---
title: "01 - Filter 2026 Hummock tracking data"
author: "Diane Pavat"
date: "2026-03-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Script to assess and remove erroneous data points and bind all datasets from same deployment.  
  
Code was written by Diane Pavat with the help of AI (chatGPT).  
  
This code is set up for pathtrack data, for data collected on Hummock in January 2026.  
  
  
# 1. Load your libraries 
```{r}
library(dplyr)
library(readr)
library(cowplot)
library(openxlsx)
library(ggplot2)
library(patchwork)
library(viridis)
library(ggpp)           # to be able to do a table in ggplot
library(leaflet)
library(htmlwidgets)
library(ggthemes)
library(leaflet.extras)
library(move2)          # for dealing with movement data
library(sf)             # for spatial manipulation
library(tidyverse)      # for data wrangling
library(lubridate)      # dealing with time
library(ggspatial)      # to annotate on maps
library(rnaturalearth)  # background maps
library(atlastools)     # movement data cleaning 
library(here)           # avoids error-prone directories
library(units)          # to handle/convert physical units
library(ggpubr)         # multipanel ggplots
library(amt)            # movement tools
library(suncalc)        # calculate sunrise/sundown
library(janitor)        # helps with data cleaning
```
  
  
# 2. Input your files  
Set the input file working directory
```{r}
hummock26_dir <- "C:/Users/r01dp24/OneDrive - University of Aberdeen/Documents/PhD/05-Data/02 - Tracking data/01 - My data/2026-01-Hummock Island"

```
  
Input the files into a database
```{r}
#hummock 2026 raw trackindata
hummock26 <- list.files(hummock26_dir, pattern = "*.csv")

# falkland island polygon
fi <- st_read("C:/Users/r01dp24/OneDrive - University of Aberdeen/Documents/PhD/05-Data/01 - Shapefile/boundaries/falklands_boundaries.shp")
fi <- st_transform(fi, 32721)

# Beauchene polygon
beauchene <- st_read("C:/Users/r01dp24/OneDrive - University of Aberdeen/Documents/PhD/05-Data/01 - Shapefile/beauchene.shp") 
beauchene <- st_transform(beauchene, 32721)
```
  
Remove beauchene for easier map display
```{r}
#difference on geometries only
fi_geom_diff <- st_difference(
  st_geometry(fi),
  st_union(st_geometry(beauchene)))

#union back into ONE geometry (so attributes stay 1-row)
fi_geom_diff <- st_union(fi_geom_diff)

#rebuild sf with original attributes
fi_no_beauchene_utm <- fi
st_geometry(fi_no_beauchene_utm) <- fi_geom_diff
fi_no_beauchene_utm <- fi_no_beauchene_utm[!st_is_empty(fi_no_beauchene_utm), ]

#have lon/lat for plotting with leaflet/ggplot
fi_no_beauchene <- st_transform(fi_no_beauchene_utm, 4326)
fi <- fi_no_beauchene
```
  
Get the files from the Hummock 2026 folder
```{r}
firstime <- TRUE
for (f in 1:length(hummock26)) {
  hummock26_f <- read_csv(paste(hummock26_dir, hummock26[f], sep = "/"))
  
  if (firstime)
  {
    fulldfhummock26 <- hummock26_f
    firstime <- FALSE
  }
  
  else{
    fulldfhummock26 <- rbind.data.frame(fulldfhummock26, hummock26_f)
  }
}
```
  
Organise the combined tracking dataset
```{r}
# Clean column names
colnames(fulldfhummock26) <- trimws(colnames(fulldfhummock26))

# Convert columns to proper types
fulldfhummock26 <- fulldfhummock26 %>%
  mutate(
    # Convert to numeric
    day = as.numeric(day),
    month = as.numeric(month),
    year = as.numeric(year),
    hour = as.numeric(hour),
    minute = as.numeric(minute),
    second = as.numeric(second),
    satellite = as.numeric(satellite),
    lat = as.numeric(lat),
    long = as.numeric(long),
    battery = as.numeric(battery),
    height = as.numeric(height),
    tagnumber = as.character(tagnumber)
  )

cat("  Tag number:", unique(fulldfhummock26$tagnumber), "\n")
cat("  Number of rows:", nrow(fulldfhummock26), "\n")
```
  
Create datetime column  
Here we apply a 3hr correction. This is because while the GPS data were recorded (Start/end time) a UTC-3 (FI timezone), the raw datafiles show the time in UTC. We therefore remove 3hr here to make sure the time is displayed in the local timezone. 
```{r}
fulldfhummock26 <- fulldfhummock26 %>%
  mutate(
    # Fix year (assuming 2000s)
    year_full = ifelse(year < 100, year + 2000, year),
    # Create datetime
    datetime = make_datetime(year_full, month, day, hour, minute, second, tz = "UTC") - hours(3)
  ) %>%
  
  mutate(
    year = year(datetime),
    month = month(datetime),
    day = day(datetime),
    hour = hour(datetime),
    minute = minute(datetime),
    second = second(datetime),
    date = as.Date(datetime),
    time = format(datetime, "%H:%M:%S"),
    year_full = as.factor(year_full)
  )
```

Check the dataset structure
```{r}
str(fulldfhummock26)

#Add a deployment ID column
fulldfhummock26$colony_name <- "Hummock Island"
fulldfhummock26$colony_code <- "HI"
fulldfhummock26$deploymentid <- "2026-01-Hummock"

str(fulldfhummock26)
```
  
# 3. Pre-filter for empty GPS points and unreasonable speeds  
Filter out invalid data points (where lat/long are 0 or NA)
```{r}
hummock26_filtered <- fulldfhummock26 %>%
  mutate(row_id = row_number())

hummock26_filtered <- hummock26_filtered %>%
  filter(lat != 0, long != 0, !is.na(lat), !is.na(long))

cat("  Valid data points (non-zero lat/long):", nrow(hummock26_filtered), "\n")

if (nrow(hummock26_filtered) == 0) {
  cat("  No valid data points found\n")
  return(NULL)
}
```
  
Filter out unreasonable speed/ step-length quality check filter
```{r}
# Start conservative, then tighten after you look at summaries.
max_speed_m_s <- 12        # e.g., 10 m/s = 36.0 km/h (very conservative)
max_step_m    <- 20000    # e.g., 20 km between successive fixes (also conservative)

pick_col <- function(df, candidates) {
  nm <- names(df)
  hit <- candidates[candidates %in% nm]
  if (length(hit) == 0) NA_character_ else hit[1]
}

auto_rows_to_remove <- integer(0)
step_metrics_all <- list()

for (b in unique(hummock26_filtered$tagnumber)) {
  
  dfb <- hummock26_filtered %>%
    filter(tagnumber == b) %>%
    filter(is.finite(long), is.finite(lat), !is.na(datetime)) %>%
    arrange(datetime)
  
  if (nrow(dfb) < 3) next
  
  # Build amt track, then project to metres (UTM 21S)
  trk_ll  <- make_track(dfb, .x = long, .y = lat, .t = datetime, crs = 4326)
  trk_utm <- transform_coords(trk_ll, 32721)
  
  st <- steps(trk_utm, keep_cols = "start")
  
  sl_col <- pick_col(st, c("sl_", "step_length", "sl"))
  dt_col <- pick_col(st, c("dt_", "dt"))
  t2_col <- pick_col(st, c("t2_", "t2"))  # end time of step (preferred)
  
  if (is.na(sl_col) || is.na(dt_col)) {
    warning("Could not find step-length/dt columns in steps() for tag ", b,
            ". Found: ", paste(names(st), collapse = ", "))
    next
  }
  
  dt_s <- as.numeric(st[[dt_col]], units = "secs")
  step_m <- as.numeric(st[[sl_col]])
  speed_m_s <- step_m / dt_s
  
  # flag bad steps (per step, i.e. between fixes)
  bad_step <- is.na(dt_s) | dt_s <= 0 |
    is.na(step_m) | step_m > max_step_m |
    is.na(speed_m_s) | speed_m_s > max_speed_m_s
  
  # collect step metrics, attached to END point time (t2_)
  if (!is.na(t2_col)) {
    step_metrics_all[[as.character(b)]] <- tibble(
      tagnumber = b,
      datetime  = st[[t2_col]],
      step_m    = step_m,
      speed_m_s = speed_m_s
    )
    
    # remove END points of bad steps
    bad_end_times <- st[[t2_col]][bad_step]
    auto_rows_to_remove <- c(
      auto_rows_to_remove,
      dfb$row_id[dfb$datetime %in% bad_end_times]
    )
    
  } else {
    # fallback alignment by index: step i corresponds to point i+1
    step_metrics_all[[as.character(b)]] <- tibble(
      tagnumber = b,
      datetime  = dfb$datetime[-1],
      step_m    = step_m,
      speed_m_s = speed_m_s
    )
    
    idx_bad <- which(bad_step)
    drop_idx <- idx_bad + 1
    drop_idx <- drop_idx[drop_idx <= nrow(dfb)]
    auto_rows_to_remove <- c(auto_rows_to_remove, dfb$row_id[drop_idx])
  }
}

auto_rows_to_remove <- sort(unique(auto_rows_to_remove))
cat(length(auto_rows_to_remove), "auto QC flagged rows", "\n")
```
  
Join step metrics onto the full filtered dataset (first point per tag will have NA step/speed)
```{r}
step_metrics_df <- bind_rows(step_metrics_all)

hummock26_with_steps <- hummock26_filtered %>%
  left_join(step_metrics_df, by = c("tagnumber", "datetime"))
```
  
"Fake cleaned" dataset (for stats only; actual removal happens later)
```{r}
hummock26_filtered2 <- hummock26_with_steps %>%
  filter(!row_id %in% auto_rows_to_remove)

cat("\n--- SPEED/STEP QC SUMMARY (AMT) ---\n")
cat("Points after coord filter:", nrow(hummock26_filtered), "\n")
cat("Points flagged by QC:", length(auto_rows_to_remove), "\n")
cat("Points after QC (if removed):", nrow(hummock26_filtered2), "\n")
cat("Max step (m) before QC:", suppressWarnings(max(hummock26_with_steps$step_m, na.rm = TRUE)), "\n")
cat("Max step (m) after QC:", suppressWarnings(max(hummock26_filtered2$step_m, na.rm = TRUE)), "\n")
cat("Max speed (m/s) before QC:", suppressWarnings(max(hummock26_with_steps$speed_m_s, na.rm = TRUE)), "\n")
cat("Max speed (m/s) after QC:", suppressWarnings(max(hummock26_filtered2$speed_m_s, na.rm = TRUE)), "\n")
```
  
# 4. Set up files for interactive map  
If you have datetime/date/time columns, keep them for popups
```{r}
has_datetime <- "datetime" %in% names(hummock26_with_steps)
has_date <- "date" %in% names(hummock26_with_steps)
has_time <- "time" %in% names(hummock26_with_steps)

#colour palette by day
days <- sort(unique(hummock26_with_steps$date))
pal_day <- colorFactor(palette = viridis(length(days)), domain = days)

#hover text (label) and click text (popup)
make_html <- function(df) {
  tm <- if (has_time)     paste0("<b>Time:</b> ", df$time, "<br>")         else ""
  
  paste0(
    "<b>ID:</b> ", df$row_id, "<br>",
    "<b>Tag:</b> ", df$tagnumber, "<br>",
    "<b>Date:</b> ", df$date, "<br>",
    tm,
    "<b>Lat:</b> ", round(df$lat, 5), "<br>",
    "<b>Lon:</b> ", round(df$long, 5), "<br>",
    "<b>Speed (m/s):</b> ", round(df$speed_m_s, 3), "<br>",
    "<b>Step (m):</b> ", round(df$step_m, 3)
  )
}


hummock26_with_steps$popup_html <- make_html(hummock26_with_steps)

#labels show on hover: keep them short so they don't cover the map
hummock26_with_steps$label_txt <- paste0(
  "ID: ", hummock26_with_steps$row_id,
  " | Tag: ", hummock26_with_steps$tagnumber,
  " | Date: ", hummock26_with_steps$date,
  " | Time: ", hummock26_with_steps$time,
  " | Lat: ", round(hummock26_with_steps$lat, 4),
  " | Lon: ", round(hummock26_with_steps$long, 4),
  " | Step (m): ", round(hummock26_with_steps$step_m,3),
  " | Speed (m/s): ", round(hummock26_with_steps$speed_m_s, 3)
)

birds <- sort(unique(hummock26_with_steps$tagnumber))

#colony coordinates
colony <- list(lat = -51.613481, lon = -60.454515)

#add buffer around the colony for visualisation and future deletion
buffer_m <- 100

colony_sf <- st_as_sf(
  data.frame(lon = colony$lon, lat = colony$lat),
  coords = c("lon", "lat"),
  crs = 4326
)

#use UTM 21S to convert distances in metres
colony_buf_sf <- colony_sf %>%
  st_transform(32721) %>%      # metres
  st_buffer(dist = buffer_m) %>%
  st_transform(4326)           # back to lon/lat for leaflet + ggplot
```
  
# 5. Interactive maps  
## 5.1. Map 1 - Check raw data to find points to remove  
```{r}
#Mark points from the auto removal list (do not remove yet)
hummock26_filter_temp <- hummock26_with_steps %>%
  mutate(to_remove = row_id %in% auto_rows_to_remove)
filtered <- leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addScaleBar(position = "bottomleft")

#add per-bird layers
for (b in birds) {
  dfb <- hummock26_filter_temp %>% filter(tagnumber == b) %>% arrange(datetime)
  
  filtered <- filtered %>%
    addPolylines(data = dfb, lng = ~ long, lat = ~ lat, group = b,
                 color = "#666666", weight = 3, opacity = 0.8) %>%
    
    addCircleMarkers(data = dfb, lng = ~ long, lat = ~ lat, group = b,
                     radius = 3, # pixels (marker size only)
                     color = ~ pal_day(date), stroke = FALSE,
                     fillOpacity = 0.8, label = ~ label_txt,
                     labelOptions = labelOptions(noHide = FALSE, 
                                                 direction = "auto",
                                                 textsize = "12px",
                                                 opacity = 0.95),
                     popup = ~ popup_html) %>%
    
    # Points flagged for removal
    addCircleMarkers(
      data = dfb %>% filter(to_remove),
      lng = ~ long, lat = ~ lat, group = b,
      radius = 6, color = "red", stroke = TRUE, weight = 2,
      fillColor = "red", fillOpacity = 0.95,
      label = ~ paste0("REMOVE → ", label_txt),
      labelOptions = labelOptions(noHide = FALSE,
                                  direction = "auto",
                                  textsize = "12px",
                                  opacity = 0.95),
      popup = ~ popup_html)
}

# Add colony buffer (true 100 m)
filtered <- filtered %>%
  addPolygons(data = colony_buf_sf, color = "orange", weight = 2,
              fillColor = "orange", fillOpacity = 0.08,
              group = "Colony buffer", label = paste0("Colony buffer: ",
                                                      buffer_m, " m"))

# Add colony point
filtered <- filtered %>%
  addCircleMarkers(lng = colony$lon, lat = colony$lat, radius = 7, 
                   color = "orange", fillColor = "orange",
                   fillOpacity = 0.9, stroke = FALSE, label = "Colony",
                   popup = paste0("<b>Colony</b><br>Lat: ", colony$lat,
                                  "<br>Lon: ", colony$lon),
                   group = "Colony") %>%
  
  addLayersControl(
    #overlayGroups = c(birds, "Colony", "Colony buffer"),
    overlayGroups = c(birds),
    options = layersControlOptions(collapsed = FALSE)
  ) %>%
  
  addLegend(
    position = "bottomleft", pal = pal_day,
    values = hummock26_filter_temp$date, title = "Date", opacity = 1
  )

# Start by showing only the first bird (so toggles are visible)
filtered <- filtered %>%
  hideGroup(birds) %>%
  showGroup(birds[1]) #%>%
  # showGroup("Colony") %>%
  # showGroup("Colony buffer")

filtered
```
  
## 5.2. Map 2 - Remove erroenous points and check new map
In this new interactive map you will be able to highlight all the points to be removed in red. This the moment you want to check that nothing was missed or added by msitake.  

1. list the points to remove or to keep from the automatic list
```{r}
manual_rows_to_remove <- c(7766, #tag 14169
                           9118, 9888, #tag 30331
                           #7142, 7141, 7139, 7138, 7153, #tag 30354
                           8968, # 8555, 8110, 8090, 8091, 8092, 8093, 8095, 8624, 8627,  #tag 31440
                           #81, 367, 368, 369, 574, #tag 31583
                           #2215, 2286, 2354, #tag 31589
                           #10212, 10200, 10239, 10211, 10205, 10455, #tag 31610
                           #6216, 6198, 6203, 6202, #tag 32635
                           3033, 3169, #3262, 3536, 3036, 3472, #tag 31717
                           4571, 4630, 4785, #3763, 4035, 3788, 3784, 4080, #tag 32722
                           5718 #5104, 5163, 4912, 4893, #tag 32735
                           #10800, 13826, 13825, 13824, 13823, 13822, 13821, 
                           #11166, 10838, 10646, 10635, 11113, 10691 #tag 32748
                           
)

auto_rows_to_keep <- c(2679, #tag 31589
                       10065, 10591, #tag 31610
                       13808 #tag 32748
)

# Remove "kept" points from the auto-flagged list
auto_rows_to_remove_final <- setdiff(auto_rows_to_remove, auto_rows_to_keep)

rows_to_remove <- sort(unique(c(
  auto_rows_to_remove_final,
  manual_rows_to_remove
)))

cat(length(auto_rows_to_remove), "auto flagged rows", "-", length(auto_rows_to_keep), "flagged rows to keep", "\n")
cat(length(auto_rows_to_remove_final), "auto rows to remove",  "\n")
cat(length(manual_rows_to_remove), "manual rows to remove", "\n")
cat(length(rows_to_remove), "rows to be removed", "\n")


# Mark points for removal (do not remove yet)
hummock26_review <- hummock26_with_steps %>%
  mutate(to_remove = row_id %in% rows_to_remove)
```
  
2. map all the points with removals in red.
```{r}
review_map <- leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addScaleBar(position = "bottomleft")

for (b in birds) {
  dfb <- hummock26_review %>% filter(tagnumber == b) %>% arrange(datetime)
  
  review_map <- review_map %>%
    addPolylines(data = dfb, lng = ~ long, lat = ~ lat, group = b,
                 color = "#666666", weight = 3, opacity = 0.8) %>%
    # Normal points
    addCircleMarkers(data = dfb %>% 
                       filter(!to_remove), lng = ~ long, lat = ~ lat,
                     group = b, radius = 3, color = ~ pal_day(date), 
                     stroke = FALSE, fillOpacity = 0.7, label = ~ label_txt,
                     labelOptions = labelOptions(noHide = FALSE,
                                                 direction = "auto",
                                                 textsize = "12px",
                                                 opacity = 0.95),
                     popup = ~ popup_html) %>%
    
    # Points flagged for removal
    addCircleMarkers(
      data = dfb %>% 
        filter(to_remove), lng = ~ long, lat = ~ lat, group = b,
      radius = 6, color = "red", stroke = TRUE, weight = 2,
      fillColor = "red", fillOpacity = 0.95,
      label = ~ paste0("REMOVE → ", label_txt),
      labelOptions = labelOptions(noHide = FALSE, direction = "auto",
                                  textsize = "12px", opacity = 0.95),
      popup = ~ popup_html)
}

# Add colony buffer + colony point
review_map <- review_map %>%
  addPolygons(
    data = colony_buf_sf, color = "orange", weight = 2, 
    fillColor = "orange", fillOpacity = 0.08, group = "Colony buffer",
    label = paste0("Colony buffer: ", buffer_m, " m")
  ) %>% 
  addCircleMarkers(lng = colony$lon, lat = colony$lat, radius = 7,
                   color = "orange", fillColor = "orange", 
                   fillOpacity = 0.9, stroke = FALSE, label = "Colony",
                   popup = paste0("<b>Colony</b><br>Lat: ", colony$lat,
                                  "<br>Lon: ", colony$lon),
                   group = "Colony"
  ) %>%
  addLayersControl(
    overlayGroups = c(birds, "Colony", "Colony buffer"),
    options = layersControlOptions(collapsed = FALSE)
  ) %>%
  addLegend(
    position = "bottomleft",
    pal = pal_day,
    values = hummock26_review$date,
    title = "Date",
    opacity = 1
  ) %>%
  hideGroup(birds) %>%
  showGroup(birds[1]) %>%
  showGroup("Colony") %>%
  showGroup("Colony buffer")

review_map
```
  
## 5.3. Map 3 - Final check cleaned data
```{r}
hummock26_clean <- hummock26_review %>%
  filter(!row_id %in% rows_to_remove)

birds_clean <- sort(unique(as.character(hummock26_clean$tagnumber)))

cleaned <- leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addScaleBar(position = "bottomright")

for (b in birds_clean) {
  dfb <- hummock26_clean %>% filter(tagnumber == b) %>% arrange(datetime)
  
  cleaned <- cleaned %>%
    addPolylines(data = dfb, lng = ~ long, lat = ~ lat, group = b,
                 color = "#666666", weight = 3, opacity = 0.8) %>%
    
    addCircleMarkers(data = dfb, lng = ~ long, lat = ~ lat, group = b,
                     radius = 3, color = ~ pal_day(date), stroke = FALSE,
                     fillOpacity = 0.8, 
                     label = ~ label_txt, 
                     labelOptions = labelOptions(noHide = FALSE,
                                                 direction = "auto",
                                                 textsize = "12px",
                                                 opacity = 0.95),
                     popup = ~ popup_html)
}

cleaned <- cleaned %>%
  addPolygons(data = colony_buf_sf, color = "orange", weight = 2,
              fillColor = "orange", fillOpacity = 0.08, 
              group = "Colony buffer", label = paste0("Colony buffer: ",
                                                      buffer_m, " m")
  ) %>%
  addCircleMarkers(lng = colony$lon, lat = colony$lat, radius = 7,
                   color = "orange", fillColor = "orange",
                   fillOpacity = 0.9, stroke = FALSE, label = "Colony",
                   popup = paste0("<b>Colony</b><br>Lat: ", colony$lat,
                                  "<br>Lon: ", colony$lon), 
                   group = "Colony"
  ) %>%
  addLayersControl(
    overlayGroups = c(birds_clean, "Colony", "Colony buffer"),
    options = layersControlOptions(collapsed = FALSE)
  ) %>%
  addLegend(
    position = "bottomleft",
    pal = pal_day,
    values = hummock26_clean$date,
    title = "Date",
    opacity = 1
  ) %>%
  hideGroup(birds_clean) %>%
  showGroup(birds_clean[1]) %>%
  showGroup("Colony") %>%
  showGroup("Colony buffer")

cleaned
```
Save the files and plots
```{r}
filename <- paste("C:/Users/r01dp24/OneDrive - University of Aberdeen/Documents/PhD/05-Data/02 - Tracking data/03 - Processed data/2026_01_Hummock.csv",
  sep = "")
write.csv(hummock26_clean, filename, row.names = F)
```  

# 6. Checking time-lag between fixes
Use move2 functions so need to transform our data into a move2 object
```{r}
# hummock26_clean_2 <- mt_as_move2(hummock26_clean,
#                                  coords = c("long","lat"),
#                                  crs = "EPSG:4326",
#                                  time_column = "datetime",
#                                  track_id_column = "tagnumber",
#                                  track_attributes = c("tagnumber"),  # optional
#                                  na.fail = F) # if set to FALSE empty points are allowed 
# 
# time_lag <- mt_time_lags(hummock26_clean_2, units = "min")  # minutes
# summary(time_lag)
# 
# # if you want to add that to your df
# # keep in mind that you have to shift it 1 row below
# hummock26_clean_2 <- bind_cols(hummock26_clean_2, as_tibble(c(NA, time_lag[-length(time_lag)])))
# glimpse(hummock26_clean_2)
# 
# # visualise but cutting above 30min
# # for graphical ease
# ggplot(tibble(track = mt_track_id(hummock26_clean_2),
#               time_lag = as.numeric(time_lag))) +
#   geom_histogram(aes(time_lag),
#                  bins = 200) +
#   labs(x = "Time lag (min)", y = "Count", title = "Distribution of time lags") +
#   #xlim(0, 30) +
#   theme_minimal()
# 
# summary(time_lag)
```
  
# 7. Plots to share
## 7.1. Preparation for all mapping
```{r}
# Ensure tagnumber is character and date has global factor levels (so colours match across facets)
all_dates <- sort(unique(hummock26_clean$date))

plot_df_hummock26 <- hummock26_clean %>%
  mutate(tagnumber = as.character(tagnumber),
         date_f = factor(date, levels = all_dates)) %>%
  arrange(tagnumber, datetime)


# Colony point df
colony_df_hummock26 <- data.frame(lon = colony$lon, lat = colony$lat)


#set the min and max extent for the dataset to include in the facet wrap
lon_min <- min(plot_df_hummock26$long, na.rm = TRUE) - 0.002
lon_max <- max(plot_df_hummock26$long, na.rm = TRUE) + 0.002
lat_min <- min(plot_df_hummock26$lat, na.rm = TRUE) - 0.002
lat_max <- max(plot_df_hummock26$lat, na.rm = TRUE) + 0.002


# remove only the extended-trip portion for selected birds, not the whole track (for mapping)
extended_trip_cutoffs <- tibble(
  tagnumber = c("32748", "31583"),
  cutoff_datetime = as.POSIXct(
    c("2026-01-16 00:06:00",   # start of long trip for 32748
      "2026-01-16 00:10:00"),  # start of long trip for 31583
    tz = "UTC"
  )
)

plot_df_minus <- plot_df_hummock26 %>%
  left_join(extended_trip_cutoffs, by = "tagnumber") %>%
  filter(is.na(cutoff_datetime) | datetime < cutoff_datetime) %>%
  select(-cutoff_datetime)

short_lon_min <- min(plot_df_minus$long, na.rm = TRUE) - 0.002
short_lon_max <- max(plot_df_minus$long, na.rm = TRUE) + 0.002
short_lat_min <- min(plot_df_minus$lat, na.rm = TRUE) - 0.002
short_lat_max <- max(plot_df_minus$lat, na.rm = TRUE) + 0.002
```
  
## 7.2. Maps for black backgrounds (powerpoint)
### 7.2.1. Facet map | 1 panel per bird (same extent) | colour per day | with grid
```{r}
facet_tracks <- ggplot() +
  # Falklands polygon (same in every facet; will be clipped by free scales)
  geom_sf(data = fi, fill = "white", colour = "grey70", linewidth = 0.3) +
  
  # Coordinates extent
  coord_sf(xlim = c(lon_min, lon_max),
           ylim = c(lat_min, lat_max),
           expand = FALSE) +
  
  # Track line
  geom_path(data = plot_df_hummock26,
            aes(x = long, y = lat, group = interaction(tagnumber, date_f),
                colour = date_f),
            linewidth = 0.8, alpha = 1) +
  
  # Colony marker
  geom_point(data = data.frame(lon = colony$lon, lat = colony$lat),
             aes(x = lon, y = lat),
             colour = "red", size = 1) +

  facet_wrap( ~ tagnumber, nrow = 3) +
  
  scale_color_hue(l = 75, c = 100) + #makes the colours brighter for black powerpoint slides
  
  geom_label(data = plot_df_hummock26 %>%
              distinct(tagnumber),
            aes(x = Inf, y = Inf, label = tagnumber),
            hjust = 1, vjust = 1.5,
            fill = "white", colour = "black", label.size=0, size = 5) +
  
  theme_void() +
  theme(
    panel.grid.major = element_line(colour = "grey85", linewidth = 0.2),
    strip.text = element_blank(),
    legend.position = "bottom",
    legend.text = element_text(colour="white", size=20),
    legend.title = element_blank(),
    axis.title = element_blank()
  )

facet_tracks
```
  
```{r}
ggsave(filename = file.path("C:/Users/r01dp24/OneDrive - University of Aberdeen/Documents/PhD/05-Data/02 - Tracking data/03 - Processed data/2026_Hummock_facet_tracks.png"),
       plot = facet_tracks,
       width = 30,#size for powerpoint
       height = 19,#size for powerpoint
       dpi = 300,
       unit = "cm"
)
```

  
### 7.2.2. One map | all birds | colour per bird | no grid
```{r}
all_tracks_PPX <- ggplot() +
  geom_sf(data = fi, fill = "white", colour = "grey40", linewidth = 0.3) +
  
  geom_path(data  = plot_df_hummock26,
            aes(x = long, y = lat, group = tagnumber, colour = tagnumber),
            alpha = 1, linewidth = 0.8 ) +
  
  scale_color_hue(l = 75, c = 100) + #makes the colours brighter for black powerpoint slides
  
  coord_sf(
    xlim = c(lon_min, lon_max),
    ylim = c(lat_min, lat_max),
    expand = FALSE
  ) +
  
  theme_void() +
  theme(
    legend.position = "none",
    axis.title = element_blank(),
    text = element_text(colour = "white", size = 15)
  )

print(all_tracks_PPX)
```
  
```{r}
ggsave(filename = file.path("C:/Users/r01dp24/OneDrive - University of Aberdeen/Documents/PhD/05-Data/02 - Tracking data/03 - Processed data/2026_Hummock_all_tracks_PPX.png"),
  plot = all_tracks_PPX,
  width = 23.64,
  height = 18.90,
  dpi = 300,
  unit = "cm"
)
```
  
  
### 7.2.3. One map | all birds | 1 colour for 2026 | with inset map
```{r}
################################################################################
# import world dataset to map the south american continuent
world <- ne_countries(scale = "medium", returnclass = "sf")
world_32721 <- st_transform(world, 32721)

south_america <- world %>%
  filter(continent == "South America") %>%
  st_transform(32721)


# Build a bounding box from the tracks for the inset map (red location square)
stopifnot(all(c("long", "lat") %in% names(plot_df_hummock26)))

plot_df_hummock26_xy <- plot_df_hummock26 %>%
  filter(!is.na(long), !is.na(lat))

if (nrow(plot_df_hummock26_xy) == 0) {
  stop("plot_df_hummock26 has no non-NA long/lat values, cannot build bounding box.")
}

xmin <- min(plot_df_hummock26_xy$long)
xmax <- max(plot_df_hummock26_xy$long)
ymin <- min(plot_df_hummock26_xy$lat)
ymax <- max(plot_df_hummock26_xy$lat)

# small padding in degrees (safe here)
pad_x <- 0.05
pad_y <- 0.05

track_bbox <- st_bbox(c(xmin = xmin,xmax = xmax,
                        ymin = ymin, ymax = ymax),
                      crs = st_crs(4326))

track_bbox_pad <- st_bbox(c(xmin = xmin - pad_x, xmax = xmax + pad_x,
                            ymin = ymin - pad_y, ymax = ymax + pad_y),
                          crs = st_crs(4326))

bb_poly     <- st_as_sfc(track_bbox) %>% st_as_sf()
bb_poly_pad <- st_as_sfc(track_bbox_pad) %>% st_as_sf()
```


```{r}
################################################################################
#Inset map: south america + FI
inset_map2 <- ggplot() +
  geom_sf(data = fi, fill = "#FFFEEF", colour = "grey70", linewidth = 0.3) +
  geom_sf(data = south_america, fill = "#FFFEEF", colour = "grey70", linewidth = 0.3) +
  geom_sf(data = bb_poly_pad, fill = NA, colour = "red", linewidth = 0.9) +
  
  coord_sf(expand = TRUE) +
  
  theme_void() +
  theme(
    #panel.background = element_rect(fill = "white", colour = "black", linewidth = 0.4),
    #plot.background  = element_rect(fill = "white", colour = NA),
    plot.margin = margin(2, 2, 2, 2)
  )

#Main map: with tracks
all_tracks_report_inset2 <- ggplot() +
  geom_sf(data = fi, fill = "#FFFEEF", colour = "grey70", linewidth = 0.3) +
  geom_sf(data = south_america, fill = "#FFFEEF", colour = "grey70", linewidth = 0.3) +
  
  geom_path(data  = plot_df_hummock26,
            aes(x = long, y = lat, group = tagnumber),
            colour = "#6DDFE5",
            alpha = 1, linewidth = 2) +
  
  coord_sf(xlim = c(-70, lon_max), #allows wider to see argentinian coast
           ylim = c(lat_min, lat_max),
           expand = TRUE) +
  
  theme_void() +
  theme(
    legend.position = "none",
    axis.text = element_text(colour = "grey", size = 28),
    axis.title = element_blank(),
    axis.text.y = element_text(margin = margin(r = 5)), #(r)ight, (l)eft
    axis.text.x = element_text(margin = margin(t = 5)), #(t)op, (b)ottom
    text = element_text(colour = "black", size = 15),
    panel.grid.major = element_line(colour="grey80"),
    panel.grid.minor = element_line(colour="grey50"),
    plot.margin = margin(10, 18, 10, 10) #Margin space so the inset can sit without hiding data (top, right, bottom, left)
    
  ) 
  
  # annotate("text", x = -70.35, y = lat_max - 0.02,
  #   label = "Argentina",
  #   hjust = 0,   # left align
  #   vjust = 1,   # top align
  #   size = 5,
  #   colour = "black"
  # ) +
  # 
  # annotate("text", x = lon_max + 0.1, y = lat_min + 0.15,
  #          label = "FI",
  #          hjust = 0,   # left align
  #          vjust = 1,   # top align
  #          size = 5,
  #          colour = "black"
  # )

#Combine: place inset in the margin area (not over the tracks)
all_tracks_report_inset2 <- ggdraw() +
  draw_plot(all_tracks_report_inset2) +
  draw_plot(inset_map2, x = 0.75, y = 0.55, # move into the top-right margin
            width = 0.26, height = 0.26)

print(all_tracks_report_inset2)
```
  
```{r}
ggsave(filename = file.path("C:/Users/r01dp24/OneDrive - University of Aberdeen/Documents/PhD/05-Data/02 - Tracking data/03 - Processed data/2026_Hummock_all_tracks_report_inset_white.png"),
       plot = all_tracks_report_inset2,
       width = 42,
       height = 33,
       dpi = 300,
       unit = "cm"
)
```
   
   
### 7.2.4. One map | all birds minus extended trips | colour per bird | no grid
```{r}
short_tracks_PPX <- ggplot() +
  geom_sf(data = fi, fill = "white", colour = "grey40", linewidth = 0.3) +
  
  geom_path(data  = plot_df_minus,
            aes(x = long, y = lat, group = tagnumber, colour = tagnumber),
            alpha = 1, linewidth = 0.8) +
  
  scale_color_hue(l = 75, c = 100) + #makes the colours brighter for black powerpoint slides
  
  coord_sf(xlim = c(short_lon_min, short_lon_max),
           ylim = c(short_lat_min, short_lat_max),
           expand = FALSE) +
  
  theme_void() +
  theme(
    legend.position = "none",
    axis.title = element_blank(),
    text = element_text(colour = "white", size = 15)
  )

print(short_tracks_PPX)
```
  
```{r}
ggsave(filename = file.path("C:/Users/r01dp24/OneDrive - University of Aberdeen/Documents/PhD/05-Data/02 - Tracking data/03 - Processed data/2026_Hummock_short_tracks_PPX.png"),
       plot = short_tracks_PPX,
       width = 23.64,
       height = 18.90,
       dpi = 300,
       unit = "cm"
)
```

  
## 7.3. Maps for white backgrounds (report)
### 7.3.1. One map | all birds | colour per individual | with inset map
```{r}
################################################################################
# import world dataset to map the south american continuent
world <- ne_countries(scale = "medium", returnclass = "sf")
world_32721 <- st_transform(world, 32721)

south_america <- world %>%
  filter(continent == "South America") %>%
  st_transform(32721)


# Build a bounding box from the tracks for the inset map (red location square)
stopifnot(all(c("long", "lat") %in% names(plot_df_hummock26)))

plot_df_hummock26_xy <- plot_df_hummock26 %>%
  filter(!is.na(long), !is.na(lat))

if (nrow(plot_df_hummock26_xy) == 0) {
  stop("plot_df_hummock26 has no non-NA long/lat values, cannot build bounding box.")
}

xmin <- min(plot_df_hummock26_xy$long)
xmax <- max(plot_df_hummock26_xy$long)
ymin <- min(plot_df_hummock26_xy$lat)
ymax <- max(plot_df_hummock26_xy$lat)

# small padding in degrees (safe here)
pad_x <- 0.05
pad_y <- 0.05

track_bbox <- st_bbox(c(xmin = xmin,xmax = xmax,
                        ymin = ymin, ymax = ymax),
                      crs = st_crs(4326))

track_bbox_pad <- st_bbox(c(xmin = xmin - pad_x, xmax = xmax + pad_x,
                            ymin = ymin - pad_y, ymax = ymax + pad_y),
                          crs = st_crs(4326))

bb_poly     <- st_as_sfc(track_bbox) %>% st_as_sf()
bb_poly_pad <- st_as_sfc(track_bbox_pad) %>% st_as_sf()


################################################################################
#Inset map: south america + FI
inset_map <- ggplot() +
  geom_sf(data = fi, fill = "grey85", colour = "grey40", linewidth = 0.3) +
  geom_sf(data = south_america, fill = "grey85", colour = "grey40", linewidth = 0.3) +
  geom_sf(data = bb_poly_pad, fill = NA, colour = "red", linewidth = 0.9) +
  
  coord_sf(expand = TRUE) +
  
  theme_void() +
  theme(
    panel.background = element_rect(fill = "white", colour = "black", linewidth = 0.4),
    plot.background  = element_rect(fill = "white", colour = NA),
    plot.margin = margin(2, 2, 2, 2)
  )

#Main map: with tracks
all_tracks_report_inset <- ggplot() +
  geom_sf(data = fi, fill = "grey85", colour = "grey40", linewidth = 0.3) +
  geom_sf(data = south_america, fill = "grey85", colour = "grey40", linewidth = 0.3) +
  
  geom_path(data  = plot_df_hummock26,
            aes(x = long, y = lat, group = tagnumber, colour = tagnumber),
            alpha = 1, linewidth = 0.8) +
  
  coord_sf(xlim = c(-70, lon_max), #allows wider to see argentinian coast
           ylim = c(lat_min, lat_max),
           expand = TRUE) +
  
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text = element_text(colour = "grey"),
    axis.title = element_blank(),
    axis.text.x = element_text(margin = margin(t = 4)),
    text = element_text(colour = "black", size = 15),
    panel.grid.major = element_line(colour="grey80"),
    panel.grid.minor = element_line(colour="grey50"),
    plot.margin = margin(10, 18, 10, 10) #Margin space so the inset can sit without hiding data (top, right, bottom, left)
  ) +
  
  annotate("text", x = -70.35, y = lat_max - 0.02,
    label = "Argentina",
    hjust = 0,   # left align
    vjust = 1,   # top align
    size = 5,
    colour = "black"
  ) +
  
  annotate("text", x = lon_max + 0.1, y = lat_min + 0.15,
           label = "FI",
           hjust = 0,   # left align
           vjust = 1,   # top align
           size = 5,
           colour = "black"
  )

#Combine: place inset in the margin area (not over the tracks)
all_tracks_report_inset <- ggdraw() +
  draw_plot(all_tracks_report_inset) +
  draw_plot(inset_map, x = 0.70, y = 0.60, # move into the top-right margin
    width = 0.26, height = 0.26)

print(all_tracks_report_inset)
```
  
```{r}
ggsave(filename = file.path("C:/Users/r01dp24/OneDrive - University of Aberdeen/Documents/PhD/05-Data/02 - Tracking data/03 - Processed data/2026_Hummock_all_tracks_report_inset.png"),
       plot = all_tracks_report_inset,
       width = 23.64,
       height = 18.90,
       dpi = 300,
       unit = "cm"
)
```

  
### 7.3.2. One map | all birds minus extended trips | colour per bird |  without inset map
```{r}
short_tracks_report <- ggplot() +
  geom_sf(data = fi, fill = "lightgrey", colour = "grey40", linewidth = 0.3) +
  
  geom_path(data  = plot_df_minus,
            aes(x = long, y = lat, group = tagnumber, colour = tagnumber),
            alpha = 1, linewidth = 0.8) +
  
  scale_color_hue(l = 75, c = 100) + #makes the colours brighter for black powerpoint slides
  
  coord_sf(
    xlim = c(short_lon_min, short_lon_max),
    ylim = c(short_lat_min, short_lat_max),
    expand = FALSE
  ) +
  
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.title = element_blank(),
    axis.text = element_text(colour = "grey70", size = 14),
    axis.ticks = element_line(colour = "grey70"),
    panel.grid.major = element_line(colour = "grey70", linewidth = 0.3),
    panel.grid.minor = element_line(colour = "grey70", linewidth = 0.2)
  ) +
  
  
  annotate("label", x = short_lon_max - 0.005, y = short_lat_max - 0.005,
           label = "2026-01-Hummock", hjust = 1, vjust = 1, fill = "white",
           linewidth = 0, size = 8, colour="black")

print(short_tracks_report)
```  
  
```{r}
ggsave(filename = file.path("C:/Users/r01dp24/OneDrive - University of Aberdeen/Documents/PhD/05-Data/02 - Tracking data/03 - Processed data/2026_Hummock_short_tracks_report.png"),
       plot = short_tracks_report,
       width = 23.64,
       height = 18.90,
       dpi = 300,
       unit = "cm"
)
```

  
### 7.3.3. One map | all birds minus extended trips | colour per bird | with inset map
```{r}
################################################################################
# Build a bounding box from the tracks for the inset map (red location square)
stopifnot(all(c("long", "lat") %in% names(plot_df_minus)))

plot_df_minus_xy <- plot_df_minus %>%
  filter(!is.na(long), !is.na(lat))

if (nrow(plot_df_minus_xy) == 0) {
  stop("plot_df_minus has no non-NA long/lat values, cannot build bounding box.")
}

xmin_minus <- min(plot_df_minus_xy$long)
xmax_minus <- max(plot_df_minus_xy$long)
ymin_minus <- min(plot_df_minus_xy$lat)
ymax_minus <- max(plot_df_minus_xy$lat)

# small padding in degrees (safe here)
pad_x <- 0.05
pad_y <- 0.05

track_bbox <- st_bbox(c(xmin = xmin_minus, xmax = xmax_minus,
                        ymin = ymin_minus, ymax = ymax_minus),
                      crs = st_crs(4326))

track_bbox_pad <- st_bbox(c(xmin = xmin_minus - pad_x, xmax = xmax_minus + pad_x,
                            ymin = ymin_minus - pad_y, ymax = ymax_minus + pad_y),
                          crs = st_crs(4326))

bb_poly     <- st_as_sfc(track_bbox) %>% st_as_sf()
bb_poly_pad <- st_as_sfc(track_bbox_pad) %>% st_as_sf()


################################################################################
#Inset map: south america + FI
inset_map <- ggplot() +
  geom_sf(data = fi, fill = "grey85", colour = "grey40", linewidth = 0.3) +
  geom_sf(data = bb_poly_pad, fill = NA, colour = "red", linewidth = 0.9) +
  
  coord_sf(expand = TRUE) +
  
  theme_void() +
  theme(
    panel.background = element_rect(fill = "white", colour = "black", linewidth = 0.4),
    plot.background  = element_rect(fill = "white", colour = NA),
    plot.margin = margin(2, 2, 2, 2)
  )


#Main map: with tracks
short_tracks_report_inset <- ggplot() +
  geom_sf(data = fi, fill = "grey85", colour = "grey40", linewidth = 0.3) +
  
  geom_path(data  = plot_df_minus,
            aes(x = long, y = lat, group = tagnumber, colour = tagnumber),
            alpha = 1, linewidth = 0.8) +
  
  scale_color_hue(l = 75, c = 100) + #makes the colours brighter for black powerpoint slides
  
  coord_sf(
    xlim = c(short_lon_min, short_lon_max),
    ylim = c(short_lat_min, short_lat_max),
    expand = FALSE
  ) +
  
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text = element_text(colour = "grey"),
    axis.title = element_blank(),
    axis.text.x = element_text(margin = margin(t = 4)),
    text = element_text(colour = "black", size = 15),
    panel.grid.major = element_line(colour="grey80"),
    panel.grid.minor = element_line(colour="grey50"),
    plot.margin = margin(10, 18, 10, 10) #Margin space so the inset can sit without hiding data (top, right, bottom, left)
  ) 

#Combine: place inset in the margin area (not over the tracks)
short_tracks_report_inset <- ggdraw() +
  draw_plot(short_tracks_report_inset) +
  draw_plot(inset_map, x = 0.75, y = 0.70, # move into the top-right margin
    width = 0.26, height = 0.26)

print(short_tracks_report_inset)
```
  
```{r}
ggsave(filename = file.path("C:/Users/r01dp24/OneDrive - University of Aberdeen/Documents/PhD/05-Data/02 - Tracking data/03 - Processed data/2026_Hummock_short_tracks_report_inset.png"),
       plot = short_tracks_report_inset,
       width = 23.64,
       height = 18.90,
       dpi = 300,
       unit = "cm"
)
```

