---
title: "03 - Setup and data load"
output:
  html_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
# Set global chunk options so code is shown and warnings/messages do not clutter the document.
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

# Purpose

This script is **not** a second raw-data cleaning pipeline.

The trip-identification script (02) has already done the following:
- reading the raw CSV files
- parsing timestamps
- assigning colony information
- identifying trips
- applying QA and overrides
- exporting clean outputs

So here, we only do a **light analysis setup step**:
- load the outputs from that workflow
- make sure key columns have the right type for downstream analyses
- create a few convenient fields such as `bird_id` and `year_season`
- save analysis-ready `.rds` files for the later scripts

# Packages
```{r packages}
library(dplyr) # data manipulation
library(readr) # reading and writing csv files
library(lubridate) # date-time parsing and date helpers
library(openxlsx) # optional Excel writing for metadata outputs
```

# Project folders
```{r folders}
# Use the current working directory as the project root.
project_dir <- getwd()

# Folder containing outputs produced by your trip-identification workflow.
input_dir <- file.path(project_dir, "visualisation")

# Root folder for the new paper-analysis workflow.
output_root <- file.path(project_dir, "outputs")

# Folder for the clean, analysis-ready objects created by this script.
output_dir <- file.path(output_root, "03_setup_and_load")

# Create the output folders if they do not already exist.
dir.create(output_root, recursive = TRUE, showWarnings = FALSE)
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
```

# Input files
```{r input-files}
# Final trip-level table exported by your trip-identification script.
trips_file <- file.path(input_dir, "RH_FORAGING_TRIPS.csv")

# Point-level tracking table with final trip IDs pushed back to points.
tracks_file <- file.path(input_dir, "RH_TRACK_POINTS_WITH_GLOBAL_TRIP_ID.csv")

# QA table describing accepted and rejected trip candidates.
qa_file <- file.path(input_dir, "RH_TRIP_QA_TABLE.csv")

# Stop immediately if any required file is missing.
if (!file.exists(trips_file)) stop("Missing file: ", trips_file)
if (!file.exists(tracks_file)) stop("Missing file: ", tracks_file)
if (!file.exists(qa_file)) stop("Missing file: ", qa_file)
```

# Helper function
```{r helper-function}
# This helper creates a simple year label from the deployment ID.
# It is only for downstream analysis convenience.
classify_year_season <- function(deploymentid_vec) {
  dplyr::case_when(
    grepl("^2025", deploymentid_vec) ~ "2025",
    grepl("^2026", deploymentid_vec) ~ "2026",
    TRUE ~ "other"
  )
}
```

# Load the already-clean outputs
```{r load-data}
# Read the trip-level table.
trips <- readr::read_csv(trips_file, show_col_types = FALSE)

# Read the point-level track table.
tracks <- readr::read_csv(tracks_file, show_col_types = FALSE)

# Read the QA table.
qa <- readr::read_csv(qa_file, show_col_types = FALSE)
```

# Standardise the track table for downstream analysis
```{r standardise-tracks}
# Make sure that the columns needed by later analyses have the right type.

tracks <- tracks %>%
  mutate(
    # Parse the tracking datetime as a proper POSIXct object.
    datetime = lubridate::ymd_hms(datetime, quiet = TRUE),

    # # Create a pure date field for daily summaries and plotting.
    # date = as.Date(datetime),

    # Keep identifiers as character strings to avoid factor/numeric mismatches.
    deploymentid = as.character(deploymentid),
    tagnumber = as.character(tagnumber),

    # # Rebuild a unique bird ID in a consistent format.
    # bird_id = paste(deploymentid, tagnumber, sep = "_"),

    # Add a simple year label for 2025 vs 2026 analyses.
    year_season = classify_year_season(deploymentid)
  ) %>%
  # Keep the rows ordered within each bird.
  arrange(deploymentid, tagnumber, datetime)
```

# Standardise the trip table for downstream analysis
```{r standardise-trips}
# Creating analysis-friendly fields from the already-clean trip output.

trips <- trips %>%
  mutate(
    # Keep identifiers as character strings.
    deploymentid = as.character(deploymentid),
    tag = as.character(tag),

    # Rebuild bird_id in the same format used in the tracks table.
    bird_id = paste(deploymentid, tag, sep = "_"),

    # Parse trip start and end datetimes from the saved date/time fields.
    trip_start = lubridate::ymd_hms(paste(start_date, start_time), quiet = TRUE),
    trip_end = lubridate::ymd_hms(paste(end_date, end_time), quiet = TRUE),

    # Add a simple year label for 2025 vs 2026 comparisons.
    year_season = classify_year_season(deploymentid)
  ) %>%
  # Keep trips ordered within each bird.
  arrange(deploymentid, tag, trip_start)
```

# Standardise the QA table for downstream analysis
```{r standardise-qa}
# The QA table is useful later when we want to trace back accepted/rejected segments.
# Here we only standardise types and parse the key datetime fields.

qa <- qa %>%
  mutate(
    # Keep identifiers as character strings.
    deploymentid = as.character(deploymentid),
    tagnumber = as.character(tagnumber),

    # Create the same bird_id format used elsewhere.
    bird_id = paste(deploymentid, tagnumber, sep = "_"),

    # Parse segment-level datetimes if present.
    start_datetime = lubridate::ymd_hms(start_datetime, quiet = TRUE),
    end_datetime = lubridate::ymd_hms(end_datetime, quiet = TRUE),

    # Parse trip-window datetimes if present.
    trip_window_start_datetime = lubridate::ymd_hms(trip_window_start_datetime, quiet = TRUE),
    trip_window_end_datetime = lubridate::ymd_hms(trip_window_end_datetime, quiet = TRUE),

    # Add the simple year label.
    year_season = classify_year_season(deploymentid)
  ) %>%
  # Keep QA rows ordered within bird.
  arrange(deploymentid, tagnumber, trip_window_start_datetime)
```

# Quick checks
```{r checks}
# Print a few high-level checks so you can confirm the objects loaded as expected.
cat("Number of tracking points:", nrow(tracks), "\n")
cat("Number of trips:", nrow(trips), "\n")
cat("Number of QA rows:", nrow(qa), "\n")
cat("Number of birds in tracking table:", dplyr::n_distinct(tracks$bird_id), "\n")
cat("Number of birds with trips:", dplyr::n_distinct(trips$bird_id), "\n")
```

# Save the analysis-ready objects
```{r save-rds}
# Save the standardised track table as an RDS object.
saveRDS(tracks, file.path(output_dir, "tracks_clean.rds"))

# Save the standardised trip table as an RDS object.
saveRDS(trips, file.path(output_dir, "trips_clean.rds"))

# Save the standardised QA table as an RDS object.
saveRDS(qa, file.path(output_dir, "qa_clean.rds"))
```

# Save a small metadata table
```{r save-metadata}
# Create a small metadata table so you can quickly check what was saved.
metadata_tbl <- tibble::tibble(
  object_name = c("tracks", "trips", "qa"),
  file_name = c("tracks_clean.rds", "trips_clean.rds", "qa_clean.rds"),
  n_rows = c(nrow(tracks), nrow(trips), nrow(qa)),
  created_on = as.character(Sys.time())
)

# Save the metadata table as CSV.
readr::write_csv(metadata_tbl, file.path(output_dir, "setup_metadata.csv"))
```

# End
```{r end}
cat("03_setup_and_load completed successfully.\n")
cat("Outputs saved in:", output_dir, "\n")
```
