---
title: "04 - Trip metrics"
output:
  html_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
# Set global chunk options so code is shown and warnings/messages do not clutter the document.
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

# Purpose of the script
This script uses the analysis-ready objects created by `00_setup_and_load.Rmd`.

It focuses on the **trip-level description of foraging behaviour**, including:
- trip duration
- total distance travelled
- maximum range from the colony
- straightness
- departure and arrival timing
- outward and return phase duration
- initial bearings

A special feature of this script is that it produces outputs in **two versions**:

1. **Full dataset**: includes all trips, including the exceptional extended 2026 trip.
2. **Trimmed dataset**: excludes trips with `max_distance_colony_km > 250 km`.

This means you can:
- keep the biologically real extended trip in the study
- inspect how much it affects summary statistics
- compare 2025 and 2026 both with and without the extended trip

This script is designed to be a **standalone module**:
- it reads the clean inputs from script 03
- it creates its own tables, figures, and model outputs
- it saves everything into its own output folder

Maximum foraging range was calculated as the maximum great-circle distance between each location and the colony. Total distance travelled was calculated as the cumulative sum of great-circle distances between consecutive GPS locations.

# 1. Set up workflow
## 1.1. Packages
```{r packages, message = FALSE, warning = FALSE}
library(dplyr) # data manipulation
library(readr) # writing csv outputs
library(lubridate) # time handling
library(ggplot2) # plotting
library(openxlsx) # writing a workbook of outputs
library(lme4)         # mixed models
library(lmerTest)     # p-values for lmer models
library(performance)  # ICC and model checks
library(broom.mixed)  # tidy model summaries
library(glmmTMB)      # fallback GLMMs if needed
library(gplots)
library(patchwork)
library(gridExtra)
library(GGally)
library(mosaic)
library(lattice)
library(see)
library(effects)
library(DHARMa)
```

## 1.2. Project folders
```{r folders}
# Use the current working directory as the project root.
project_dir <- getwd()

# Folder containing the analysis-ready objects from script 00.
input_dir <- file.path(project_dir, "outputs", "03_setup_and_load")

# Output folder for this script.
output_dir <- file.path(project_dir, "outputs", "04_trip_metrics")

# Create the output folder if needed.
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
```

## 1.3. Analysis settings
```{r analysis-settings}
# Define the threshold used to flag exceptionally long trips.
# Any trip with maximum range above this threshold will be marked as an extended trip.
extended_trip_threshold_km <- 250
```

## 1.4. Helper functions
```{r helpers}
# Simple plotting theme used throughout this script.
theme_paper <- function() {
  theme_minimal() +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(colour = "grey85", linewidth = 0.4),
      axis.text = element_text(colour = "grey20"),
      plot.title = element_text(face = "bold", hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5),
      legend.position = "right"
    )
}

# Bearing helper:
# compute the initial bearing in degrees between two longitude/latitude points.
calculate_bearing_deg <- function(lon1, lat1, lon2, lat2) {
  lon1 <- lon1 * pi / 180
  lat1 <- lat1 * pi / 180
  lon2 <- lon2 * pi / 180
  lat2 <- lat2 * pi / 180

  y <- sin(lon2 - lon1) * cos(lat2)
  x <- cos(lat1) * sin(lat2) - sin(lat1) * cos(lat2) * cos(lon2 - lon1)

  bearing <- atan2(y, x) * 180 / pi
  (bearing + 360) %% 360
}
```


## 1.5. Load the analysis-ready inputs
```{r load-inputs}
# Load the clean point-level tracking object.
tracks <- readRDS(file.path(input_dir, "tracks_clean.rds"))

# Load the clean trip-level object.
trips <- readRDS(file.path(input_dir, "trips_clean.rds"))

# Load the clean QA object.
qa <- readRDS(file.path(input_dir, "qa_clean.rds"))
```


## 1.6. Left join sex info
```{r load-inputs}
# Load the dataset with information about sexes of individuals
sexinfo <- read.xlsx("C:/Users/r01dp24/OneDrive - University of Aberdeen/Documents/PhD/05-Data/02 - Tracking data/01 - My data/2025-01-Hummock Island/sex_tagnumber.xlsx")

# left join
tracks <- tracks %>%
  left_join(sexinfo %>% select(bird_id, sex), by = "bird_id")

trips <- tracks %>%
  left_join(sexinfo %>% select(bird_id, sex), by = "bird_id")
```


# 2. Trip analysis
## 2.1. Build the main trip-metrics table
```{r build-trip-metrics}
# Start from the trip table exported by your trip-detection workflow.
# Here we add a few convenient derived metrics.

trip_metrics <- trips %>%
  mutate(
    # Straightness index: a simple measure of how direct the trip was.
    straightness_index = max_distance_colony_km / total_distance_km, #1=perfectly straight trip, small values= tortuous movement

    # Convert departure time into decimal hours.
    departure_hour = hour(trip_start) + minute(trip_start) / 60,

    # Convert arrival time into decimal hours.
    arrival_hour = hour(trip_end) + minute(trip_end) / 60,

    # Compute trip midpoint.
    trip_midpoint = trip_start + (trip_end - trip_start) / 2,

    # Convert trip midpoint into decimal hours.
    trip_midpoint_hour = hour(trip_midpoint) + minute(trip_midpoint) / 60,

    # Flag exceptionally long trips using the fixed threshold.
    is_extended_trip = max_distance_colony_km > extended_trip_threshold_km
  ) %>%
  arrange(deploymentid, tag, trip_start)

glimpse(trip_metrics)
```

## 2.2. Inspect the extended trip(s)
```{r inspect-extended-trips}
# Print the trips flagged as extended so you can confirm what is being classified.
extended_trip_table <- trip_metrics %>%
  filter(is_extended_trip) %>%
  select(
    deploymentid, tag, bird_id, global_trip_id, foraging_trip,
    trip_start, trip_end, trip_duration_hours,
    total_distance_km, max_distance_colony_km
  )

extended_trip_table
```

## 2.3. Extract outward and return phase metrics from point-level trip data
```{r outward-return-phase}
# For each final trip, use the point-level data to:
# - locate the point of maximum distance from the colony
# - split the trip into an outward phase and a return phase
# - estimate initial bearings

trip_phase_metrics <- lapply(seq_len(nrow(trip_metrics)), function(i) {

  # Pull one trip row.
  tr <- trip_metrics[i, ]

  # Subset the point-level data for that trip.
  pts <- tracks %>%
    filter(global_trip_id == tr$global_trip_id) %>%
    arrange(datetime)

  # Skip trips with too few points to calculate phase metrics reliably.
  if (nrow(pts) < 3) return(NULL)

  # Identify the point of maximum distance from the colony.
  max_i <- which.max(pts$dist_to_colony_km)

  # Define the outward phase as the start of the trip up to the maximum-range point.
  outward_pts <- pts[1:max_i, ]

  # Define the return phase as the maximum-range point back to the end of the trip.
  return_pts <- pts[max_i:nrow(pts), ]

  # Estimate departure bearing from the first two outward points if available.
  dep_bearing <- if (nrow(outward_pts) >= 2) {
    calculate_bearing_deg(
      outward_pts$long[1], outward_pts$lat[1],
      outward_pts$long[2], outward_pts$lat[2]
    )
  } else {
    NA_real_
  }

  # Estimate return bearing from the first two return-phase points if available.
  ret_bearing <- if (nrow(return_pts) >= 2) {
    calculate_bearing_deg(
      return_pts$long[1], return_pts$lat[1],
      return_pts$long[2], return_pts$lat[2]
    )
  } else {
    NA_real_
  }

  # Return one row of phase metrics for this trip.
  tibble::tibble(
    global_trip_id = tr$global_trip_id,
    deploymentid = tr$deploymentid,
    tag = tr$tag,
    bird_id = tr$bird_id,
    year_season = tr$year_season,
    foraging_trip = tr$foraging_trip,
    departure_bearing_deg = dep_bearing,
    return_bearing_deg = ret_bearing,
    outward_duration_h = as.numeric(difftime(max(outward_pts$datetime), min(outward_pts$datetime), units = "hours")),
    return_duration_h = as.numeric(difftime(max(return_pts$datetime), min(return_pts$datetime), units = "hours")),
    max_distance_time = pts$datetime[max_i],
    max_distance_hour = hour(pts$datetime[max_i]) + minute(pts$datetime[max_i]) / 60 # straight-line (great-circle) distance between GPS point and colon
  )
}) %>%
  dplyr::bind_rows()

# Join these phase metrics back onto the main trip-metrics table.
trip_metrics <- trip_metrics %>%
  left_join(
    trip_phase_metrics,
    by = c("global_trip_id", "deploymentid", "tag", "bird_id", "year_season", "foraging_trip")
  )
```

## 2.4. Create the full and trimmed datasets
```{r full-and-trimmed-datasets}
# Keep a full copy of the trip-metrics table including all trips.
trip_metrics_full <- trip_metrics

# Create a trimmed version excluding extended trips above the threshold.
trip_metrics_trimmed <- trip_metrics %>%
  filter(!is_extended_trip)

# Build one combined plotting table so that full and trimmed outputs can be compared easily.
trip_metrics_plot <- bind_rows(
  trip_metrics_full %>% mutate(dataset_version = "Full dataset"),
  trip_metrics_trimmed %>% mutate(dataset_version = "Without extended trip")
)
```

## 2.5. Summary tables
```{r summary-tables}
# Summary by year for the full dataset.
trip_summary_by_year_full <- trip_metrics_full %>%
  group_by(year_season) %>%
  summarise(
    n_trips = n(),
    n_birds = n_distinct(bird_id),
    mean_duration_h = mean(trip_duration_hours, na.rm = TRUE),
    sd_duration_h = sd(trip_duration_hours, na.rm = TRUE),
    mean_total_distance_km = mean(total_distance_km, na.rm = TRUE),
    sd_total_distance_km = sd(total_distance_km, na.rm = TRUE),
    mean_max_range_km = mean(max_distance_colony_km, na.rm = TRUE),
    sd_max_range_km = sd(max_distance_colony_km, na.rm = TRUE),
    mean_straightness = mean(straightness_index, na.rm = TRUE),
    sd_straightness = sd(straightness_index, na.rm = TRUE),
    mean_departure_hour = mean(departure_hour, na.rm = TRUE),
    mean_arrival_hour = mean(arrival_hour, na.rm = TRUE),
    mean_outward_duration_h = mean(outward_duration_h, na.rm = TRUE),
    mean_return_duration_h = mean(return_duration_h, na.rm = TRUE),
    .groups = "drop"
  )

# Summary by year for the trimmed dataset.
trip_summary_by_year_trimmed <- trip_metrics_trimmed %>%
  group_by(year_season) %>%
  summarise(
    n_trips = n(),
    n_birds = n_distinct(bird_id),
    mean_duration_h = mean(trip_duration_hours, na.rm = TRUE),
    sd_duration_h = sd(trip_duration_hours, na.rm = TRUE),
    mean_total_distance_km = mean(total_distance_km, na.rm = TRUE),
    sd_total_distance_km = sd(total_distance_km, na.rm = TRUE),
    mean_max_range_km = mean(max_distance_colony_km, na.rm = TRUE),
    sd_max_range_km = sd(max_distance_colony_km, na.rm = TRUE),
    mean_straightness = mean(straightness_index, na.rm = TRUE),
    sd_straightness = sd(straightness_index, na.rm = TRUE),
    mean_departure_hour = mean(departure_hour, na.rm = TRUE),
    mean_arrival_hour = mean(arrival_hour, na.rm = TRUE),
    mean_outward_duration_h = mean(outward_duration_h, na.rm = TRUE),
    mean_return_duration_h = mean(return_duration_h, na.rm = TRUE),
    .groups = "drop"
  )

# Summary by bird for the full dataset.
trip_summary_by_bird_full <- trip_metrics_full %>%
  group_by(year_season, deploymentid, tag, bird_id) %>%
  summarise(
    n_trips = n(),
    mean_duration_h = mean(trip_duration_hours, na.rm = TRUE),
    mean_total_distance_km = mean(total_distance_km, na.rm = TRUE),
    mean_max_range_km = mean(max_distance_colony_km, na.rm = TRUE),
    mean_straightness = mean(straightness_index, na.rm = TRUE),
    mean_departure_hour = mean(departure_hour, na.rm = TRUE),
    mean_arrival_hour = mean(arrival_hour, na.rm = TRUE),
    .groups = "drop"
  )

# Summary by bird for the trimmed dataset.
trip_summary_by_bird_trimmed <- trip_metrics_trimmed %>%
  group_by(year_season, deploymentid, tag, bird_id) %>%
  summarise(
    n_trips = n(),
    mean_duration_h = mean(trip_duration_hours, na.rm = TRUE),
    mean_total_distance_km = mean(total_distance_km, na.rm = TRUE),
    mean_max_range_km = mean(max_distance_colony_km, na.rm = TRUE),
    mean_straightness = mean(straightness_index, na.rm = TRUE),
    mean_departure_hour = mean(departure_hour, na.rm = TRUE),
    mean_arrival_hour = mean(arrival_hour, na.rm = TRUE),
    .groups = "drop"
  )
```
    
  
# 3. Statistical analysis
## 3.1. Check datasets
  
Why mixed models? Trips are repeated within birds, so trips are not independent observations. A mixed model allows us to test for a year effect while accounting for repeated measures within bird by including `bird_id` as a random intercept.

### 3.1.1. Check if need data transformation
```{r}
# Keep only the variables needed for inference.
stats_full <- trip_metrics_full %>%
  select(
    bird_id,
    year_season,
    trip_start,
    trip_duration_hours,
    total_distance_km,
    max_distance_colony_km
  ) %>%
  group_by(bird_id) %>%
  arrange(trip_start, .by_group = TRUE) %>%
  mutate(trip_order = row_number()) %>%
  ungroup() %>%
  mutate(
    bird_id = factor(bird_id),
    year_season = factor(year_season),
    log_trip_duration = log(trip_duration_hours),
    log_total_distance = log(total_distance_km),
    log_max_range = log(max_distance_colony_km),
    sqrt_trip_duration = sqrt(trip_duration_hours),
    sqrt_total_distance = sqrt(total_distance_km),
    sqrt_max_range = sqrt(max_distance_colony_km)
  )

stats_trimmed <- trip_metrics_trimmed %>%
  select(
    bird_id,
    year_season,
    trip_start,
    trip_duration_hours,
    total_distance_km,
    max_distance_colony_km
  ) %>%
  group_by(bird_id) %>%
  arrange(trip_start, .by_group = TRUE) %>%
  mutate(trip_order = row_number()) %>%
  ungroup() %>%
  mutate(
    bird_id = factor(bird_id),
    year_season = factor(year_season),
    log_trip_duration = log(trip_duration_hours),
    log_total_distance = log(total_distance_km),
    log_max_range = log(max_distance_colony_km),
    sqrt_trip_duration = sqrt(trip_duration_hours),
    sqrt_total_distance = sqrt(total_distance_km),
    sqrt_max_range = sqrt(max_distance_colony_km)
  )
```

```{r}
# # Design check:
# # if birds are unique to a single year, the year effect is estimated between sets of birds rather than within the same bird across years.
# birds_in_multiple_years <- stats_full %>%
#   distinct(bird_id, year_season) %>%
#   count(bird_id, name = "n_years") %>%
#   filter(n_years > 1)
# 
# birds_in_multiple_years
```

  
Dotchart of our key stat variables
```{r, warning=FALSE, message=FALSE, tidy=TRUE}
variables <- cbind(stats_full$trip_duration_hours, stats_full$total_distance_km,
                   stats_full$max_distance_colony_km, stats_trimmed$trip_duration_hours,
                   stats_trimmed$total_distance_km, stats_trimmed$max_distance_colony_km)

colnames(variables) <- c("duration - full", "total dist - full", "max range - full",
                         "duration - trimmed", "total dist - trimmed", "max range - trimmed")

dotplot(as.matrix(variables),
        groups=FALSE,
        strip=strip.custom(bg="white",
                           par.strip.text=list(cex=0.8)),
        scales=list(x=list(relation="free"),
                    y=list(realtion="free"),
                    draw=FALSE),
        col=1, cex=0.5, pch=16,
        xlab= "value of the variable",
        ylab= "Order of the data from text file")
```
It looks like there are two outliers for the full dataset, which are most likely coming from the Argentinian trip and one outlier for the trimmed dataset, most likely from the longer trip to the south (queen Charlotte's bay).
  
  
If we apply a log or sqrt transformation, can we account for those outliers?
Full dataset:
```{r, warning=FALSE, message=FALSE, tidy=TRUE}
variablesduration_full <- cbind(stats_full$trip_duration_hours,
                                stats_full$log_trip_duration,
                                stats_full$sqrt_trip_duration)

variablestotaldist_full <- cbind(stats_full$total_distance_km,
                                 stats_full$log_total_distance, 
                                 stats_full$sqrt_total_distance)

variablesmaxrange_full <- cbind(stats_full$max_distance_colony_km,
                                stats_full$log_max_range,
                                stats_full$sqrt_max_range)

colnames(variablesduration_full) <- c("full - duration", "full - duration log", "full - duration sqrt")
colnames(variablestotaldist_full) <- c("full - dist", "full - dist log", "full - dist sqrt")
colnames(variablesmaxrange_full) <- c("full - max range", "full - max range log", "full - max range sqrt")

par(mfrow=c(1, 3))
dotplot(as.matrix(variablesduration_full),
        groups=FALSE,
        strip=strip.custom(bg="white",
                           par.strip.text=list(cex=0.8)),
        scales=list(x=list(relation="free"),
                    y=list(realtion="free"),
                    draw=FALSE),
        col=1, cex=0.5, pch=16,
        xlab= "value of the variable",
        ylab= "Order of the data from text file")


dotplot(as.matrix(variablestotaldist_full),
        groups=FALSE,
        strip=strip.custom(bg="white",
                           par.strip.text=list(cex=0.8)),
        scales=list(x=list(relation="free"),
                    y=list(realtion="free"),
                    draw=FALSE),
        col=1, cex=0.5, pch=16,
        xlab= "value of the variable",
        ylab= "Order of the data from text file")


dotplot(as.matrix(variablesmaxrange_full),
        groups=FALSE,
        strip=strip.custom(bg="white",
                           par.strip.text=list(cex=0.8)),
        scales=list(x=list(relation="free"),
                    y=list(realtion="free"),
                    draw=FALSE),
        col=1, cex=0.5, pch=16,
        xlab= "value of the variable",
        ylab= "Order of the data from text file")
```
Trimmed dataset
```{r, warning=FALSE, message=FALSE, tidy=TRUE}
variablesduration_trimmed <- cbind(stats_trimmed$trip_duration_hours,
                                   stats_trimmed$log_trip_duration,
                                   stats_trimmed$sqrt_trip_duration)

variablestotaldist_trimmed <- cbind(stats_trimmed$total_distance_km,
                                    stats_trimmed$log_total_distance,
                                    stats_trimmed$sqrt_total_distance)

variablesmaxrange_trimmed <- cbind(stats_trimmed$max_distance_colony_km,
                                   stats_trimmed$log_max_range,
                                   stats_trimmed$sqrt_max_range)

colnames(variablesduration_trimmed) <- c("trimmed - duration", 
                                         "trimmed - duration log", 
                                         "trimmed - duration sqrt")

colnames(variablestotaldist_trimmed) <- c("trimmed - dist", 
                                          "trimmed - dist log",
                                          "trimmed - dist sqrt")

colnames(variablesmaxrange_trimmed) <- c("trimmed - max range",
                                         "trimmed - max range log",
                                         "trimmed - max range sqrt")

par(mfrow=c(1, 3))
dotplot(as.matrix(variablesduration_trimmed),
        groups=FALSE,
        strip=strip.custom(bg="white",
                           par.strip.text=list(cex=0.8)),
        scales=list(x=list(relation="free"),
                    y=list(realtion="free"),
                    draw=FALSE),
        col=1, cex=0.5, pch=16,
        xlab= "value of the variable",
        ylab= "Order of the data from text file")


dotplot(as.matrix(variablestotaldist_trimmed),
        groups=FALSE,
        strip=strip.custom(bg="white",
                           par.strip.text=list(cex=0.8)),
        scales=list(x=list(relation="free"),
                    y=list(realtion="free"),
                    draw=FALSE),
        col=1, cex=0.5, pch=16,
        xlab= "value of the variable",
        ylab= "Order of the data from text file")


dotplot(as.matrix(variablesmaxrange_trimmed),
        groups=FALSE,
        strip=strip.custom(bg="white",
                           par.strip.text=list(cex=0.8)),
        scales=list(x=list(relation="free"),
                    y=list(realtion="free"),
                    draw=FALSE),
        col=1, cex=0.5, pch=16,
        xlab= "value of the variable",
        ylab= "Order of the data from text file")
```
  
For both the full and trimmed datasets, it seems that log transforming helps a little to bridge the space between the main points and outliers. log-transformed variables will be used for modelling.

### 3.1.2. Data simple stat summary
```{r}
desc_full <- stats_full %>%
  group_by(year_season) %>%
  summarise(
    n_trips = n(),
    n_birds = n_distinct(bird_id),
    mean_duration = mean(trip_duration_hours, na.rm = TRUE),
    sd_duration = sd(trip_duration_hours, na.rm = TRUE),
    mean_total_distance = mean(total_distance_km, na.rm = TRUE),
    sd_total_distance = sd(total_distance_km, na.rm = TRUE),
    mean_max_range = mean(max_distance_colony_km, na.rm = TRUE),
    sd_max_range = sd(max_distance_colony_km, na.rm = TRUE),
    .groups = "drop"
  )

desc_trimmed <- stats_trimmed %>%
  group_by(year_season) %>%
  summarise(
    n_trips = n(),
    n_birds = n_distinct(bird_id),
    mean_duration = mean(trip_duration_hours, na.rm = TRUE),
    sd_duration = sd(trip_duration_hours, na.rm = TRUE),
    mean_total_distance = mean(total_distance_km, na.rm = TRUE),
    sd_total_distance = sd(total_distance_km, na.rm = TRUE),
    mean_max_range = mean(max_distance_colony_km, na.rm = TRUE),
    sd_max_range = sd(max_distance_colony_km, na.rm = TRUE),
    .groups = "drop"
  )

desc_full
desc_trimmed

# Combined table to compare full and trimmed descriptive summaries.
year_comparison_table <- bind_rows(
  desc_full %>% mutate(dataset = "full"),
  desc_trimmed %>% mutate(dataset = "trimmed")
) %>%
  select(dataset, everything())
```

## 3.2. Run LME
### 3.2.1. LME full dataset
```{r}
# trip duration
mod_duration_full <- lmer(
  log_trip_duration ~ year_season + (1 | bird_id), 
  data = stats_full,
  REML = TRUE)

# total distance
mod_distance_full <- lmer(
  log_total_distance ~ year_season + (1 | bird_id), 
  data = stats_full,
  REML = TRUE)

# maximum distance to colony
mod_range_full <- lmer(
  log_max_range ~ year_season + (1 | bird_id), 
  data = stats_full,
  REML = TRUE)
```


### 3.2.2. LME trimmed dataset
```{r}
# trip duration
mod_duration_trimmed <- lmer(
  log_trip_duration ~ year_season + (1| bird_id), 
  data = stats_trimmed,
  REML = TRUE)

mod_distance_trimmed <- lmer(
  log_total_distance ~ year_season + (1 | bird_id), 
  data = stats_trimmed,
  REML = TRUE)

mod_range_trimmed <- lmer(
  log_max_range ~ year_season + (1 | bird_id),
  data = stats_trimmed,
  REML = TRUE)
```

### 3.2.3. LME summaries
```{r}
summary(mod_duration_full)
summary(mod_distance_full)
summary(mod_range_full)

summary(mod_duration_trimmed)
summary(mod_distance_trimmed)
summary(mod_range_trimmed)
```

Full dataset model diagnostics: visual checks + simulation-based tests
```{r}
check_model(mod_duration_full)
check_model(mod_distance_full)
check_model(mod_range_full)


sim_duration_full <- simulateResiduals(mod_duration_full, plot = FALSE)
plot(sim_duration_full)

sim_distance_full <- simulateResiduals(mod_distance_full, plot = FALSE)
plot(sim_distance_full)

sim_range_full <- simulateResiduals(mod_range_full, plot = FALSE)
plot(sim_range_full)
```

Trimmed dataset model diagnostics: visual checks + simulation-based tests
```{r}
check_model(mod_duration_trimmed)
check_model(mod_distance_trimmed)
check_model(mod_range_trimmed)


sim_duration_trimmed <- simulateResiduals(mod_duration_trimmed, plot = FALSE)
plot(sim_duration_trimmed)

sim_distance_trimmed <- simulateResiduals(mod_distance_trimmed, plot = FALSE)
plot(sim_distance_trimmed)

sim_range_trimmed <- simulateResiduals(mod_range_trimmed, plot = FALSE)
plot(sim_range_trimmed)
```

### 3.2.4. Assumption tests in table form
```{r}
diagnose_lmm <- function(sim_obj, model_obj, data_obj) {
  tibble::tibble(
    dharma_uniformity_p = DHARMa::testUniformity(sim_obj)$p.value,
    dharma_dispersion_p = DHARMa::testDispersion(sim_obj)$p.value,
    dharma_outliers_p = DHARMa::testOutliers(sim_obj)$p.value,
    shapiro_residual_p = shapiro.test(residuals(model_obj))$p.value,
    temporal_autocorrelation_p = tryCatch(
      DHARMa::testTemporalAutocorrelation(
        sim_obj,
        time = data_obj$trip_order,
        group = data_obj$bird_id
      )$p.value,
      error = function(e) NA_real_
    )
  )
}

lmm_diagnostics <- dplyr::bind_rows(
  diagnose_lmm(sim_duration_full, mod_duration_full, stats_full) %>%
    mutate(dataset = "full", response = "log_trip_duration"),
  diagnose_lmm(sim_distance_full, mod_distance_full, stats_full) %>%
    mutate(dataset = "full", response = "log_total_distance"),
  diagnose_lmm(sim_range_full, mod_range_full, stats_full) %>%
    mutate(dataset = "full", response = "log_max_range"),
  diagnose_lmm(sim_duration_trimmed, mod_duration_trimmed, stats_trimmed) %>%
    mutate(dataset = "trimmed", response = "log_trip_duration"),
  diagnose_lmm(sim_distance_trimmed, mod_distance_trimmed, stats_trimmed) %>%
    mutate(dataset = "trimmed", response = "log_total_distance"),
  diagnose_lmm(sim_range_trimmed, mod_range_trimmed, stats_trimmed) %>%
    mutate(dataset = "trimmed", response = "log_max_range")
)

lmm_diagnostics %>%
  select(dataset, response, everything()) %>%
  arrange(dataset, response)
```

### 3.2.5. Gamma mixed-effects fallback for positive skewed metrics
```{r}
# If Gaussian LMM assumptions remain weak, fit Gamma GLMMs with log link on raw metrics.
# This keeps the response on its natural scale while modelling right-skew directly.
gamma_duration_full <- glmmTMB(
  trip_duration_hours ~ year_season + (1 | bird_id),
  family = Gamma(link = "log"),
  data = stats_full
)

gamma_distance_full <- glmmTMB(
  total_distance_km ~ year_season + (1 | bird_id),
  family = Gamma(link = "log"),
  data = stats_full
)

gamma_range_full <- glmmTMB(
  max_distance_colony_km ~ year_season + (1 | bird_id),
  family = Gamma(link = "log"),
  data = stats_full
)

gamma_duration_trimmed <- glmmTMB(
  trip_duration_hours ~ year_season + (1 | bird_id),
  family = Gamma(link = "log"),
  data = stats_trimmed
)

gamma_distance_trimmed <- glmmTMB(
  total_distance_km ~ year_season + (1 | bird_id),
  family = Gamma(link = "log"),
  data = stats_trimmed
)

gamma_range_trimmed <- glmmTMB(
  max_distance_colony_km ~ year_season + (1 | bird_id),
  family = Gamma(link = "log"),
  data = stats_trimmed
)
```

```{r}
# Summarise the year effect from the Gamma models.
tidy_gamma <- function(model, dataset_label, response_label) {
  broom.mixed::tidy(model, effects = "fixed") %>%
    mutate(
      dataset = dataset_label,
      response = response_label,
      estimate_ratio = exp(estimate),
      conf_low_ratio = exp(estimate - 1.96 * std.error),
      conf_high_ratio = exp(estimate + 1.96 * std.error)
    )
}

gamma_year_effects <- bind_rows(
  tidy_gamma(gamma_duration_full, "full", "trip_duration_hours"),
  tidy_gamma(gamma_distance_full, "full", "total_distance_km"),
  tidy_gamma(gamma_range_full, "full", "max_distance_colony_km"),
  tidy_gamma(gamma_duration_trimmed, "trimmed", "trip_duration_hours"),
  tidy_gamma(gamma_distance_trimmed, "trimmed", "total_distance_km"),
  tidy_gamma(gamma_range_trimmed, "trimmed", "max_distance_colony_km")
) %>%
  filter(term == "year_season2026") %>%
  select(dataset, response, term, estimate, std.error, p.value, estimate_ratio, conf_low_ratio, conf_high_ratio)

gamma_year_effects
```

```{r}
gamma_diagnostics <- bind_rows(
  tibble(
    dataset = "full",
    response = "trip_duration_hours",
    uniformity_p = testUniformity(simulateResiduals(gamma_duration_full, plot = FALSE))$p.value,
    dispersion_p = testDispersion(simulateResiduals(gamma_duration_full, plot = FALSE))$p.value,
    outliers_p = testOutliers(simulateResiduals(gamma_duration_full, plot = FALSE))$p.value
  ),
  tibble(
    dataset = "full",
    response = "total_distance_km",
    uniformity_p = testUniformity(simulateResiduals(gamma_distance_full, plot = FALSE))$p.value,
    dispersion_p = testDispersion(simulateResiduals(gamma_distance_full, plot = FALSE))$p.value,
    outliers_p = testOutliers(simulateResiduals(gamma_distance_full, plot = FALSE))$p.value
  ),
  tibble(
    dataset = "full",
    response = "max_distance_colony_km",
    uniformity_p = testUniformity(simulateResiduals(gamma_range_full, plot = FALSE))$p.value,
    dispersion_p = testDispersion(simulateResiduals(gamma_range_full, plot = FALSE))$p.value,
    outliers_p = testOutliers(simulateResiduals(gamma_range_full, plot = FALSE))$p.value
  ),
  tibble(
    dataset = "trimmed",
    response = "trip_duration_hours",
    uniformity_p = testUniformity(simulateResiduals(gamma_duration_trimmed, plot = FALSE))$p.value,
    dispersion_p = testDispersion(simulateResiduals(gamma_duration_trimmed, plot = FALSE))$p.value,
    outliers_p = testOutliers(simulateResiduals(gamma_duration_trimmed, plot = FALSE))$p.value
  ),
  tibble(
    dataset = "trimmed",
    response = "total_distance_km",
    uniformity_p = testUniformity(simulateResiduals(gamma_distance_trimmed, plot = FALSE))$p.value,
    dispersion_p = testDispersion(simulateResiduals(gamma_distance_trimmed, plot = FALSE))$p.value,
    outliers_p = testOutliers(simulateResiduals(gamma_distance_trimmed, plot = FALSE))$p.value
  ),
  tibble(
    dataset = "trimmed",
    response = "max_distance_colony_km",
    uniformity_p = testUniformity(simulateResiduals(gamma_range_trimmed, plot = FALSE))$p.value,
    dispersion_p = testDispersion(simulateResiduals(gamma_range_trimmed, plot = FALSE))$p.value,
    outliers_p = testOutliers(simulateResiduals(gamma_range_trimmed, plot = FALSE))$p.value
  )
)

gamma_diagnostics %>%
  arrange(dataset, response)
```

The assumption checks above let you choose the model family transparently:
- if log-LMM diagnostics are acceptable, keep the LMM for interpretability on the log scale
- if diagnostics remain problematic, prefer Gamma GLMM (log link) for strictly positive skewed trip metrics

### 3.2.6. Final model choice (documented decision)
```{r}
# Transparent model-choice rule:
# if Gaussian LMM diagnostics are flagged (p < 0.05), retain Gamma GLMM as final.
model_choice_table <- lmm_diagnostics %>%
  mutate(
    flagged_non_gaussian = shapiro_residual_p < 0.05 |
      dharma_uniformity_p < 0.05 |
      dharma_outliers_p < 0.05,
    final_model = ifelse(flagged_non_gaussian, "Gamma GLMM (log link)", "Log-LMM (Gaussian)")
  ) %>%
  select(dataset, response, flagged_non_gaussian, final_model)

model_choice_table
```

Given these diagnostics, the final inferential model is Gamma GLMM (log link) for the three focal trip metrics.

## 3.3. Figures comparing full and trimmed versions
```{r figures}
# Boxplot of trip duration by year, shown for both dataset versions.
fig_duration <- ggplot(trip_metrics_plot, aes(x = year_season, y = trip_duration_hours)) +
  geom_boxplot() +
  facet_wrap(~ dataset_version, scales = "free") +
  labs(
    title = "Trip duration by year",
    subtitle = "Comparison shown with and without extended trips > 250 km",
    x = "Year",
    y = "Trip duration (hours)"
  ) +
  theme_paper()

# Boxplot of total trip distance by year, shown for both dataset versions.
fig_total_distance <- ggplot(trip_metrics_plot, aes(x = year_season, y = total_distance_km)) +
  geom_boxplot() +
  facet_wrap(~ dataset_version, scales = "free") +
  labs(
    title = "Total trip distance by year",
    subtitle = "Comparison shown with and without extended trips > 250 km",
    x = "Year",
    y = "Total distance (km)"
  ) +
  theme_paper()

# Boxplot of maximum range by year, shown for both dataset versions.
fig_max_range <- ggplot(trip_metrics_plot, aes(x = year_season, y = max_distance_colony_km)) +
  geom_boxplot() +
  facet_wrap(~ dataset_version, scales = "free") +
  labs(
    title = "Maximum range by year",
    subtitle = "Comparison shown with and without extended trips > 250 km",
    x = "Year",
    y = "Maximum distance from colony (km)"
  ) +
  theme_paper()

# Boxplot of departure hour by year, shown for both dataset versions.
fig_departure <- ggplot(trip_metrics_plot, aes(x = year_season, y = departure_hour)) +
  geom_boxplot() +
  facet_wrap(~ dataset_version, scales = "free") +
  labs(
    title = "Departure hour by year",
    subtitle = "Comparison shown with and without extended trips > 250 km",
    x = "Year",
    y = "Departure hour"
  ) +
  theme_paper()

# Scatterplot of outward and return phase duration, shown for both dataset versions.
fig_outward_return <- ggplot(trip_metrics_plot, aes(x = outward_duration_h, y = return_duration_h, colour = year_season)) +
  geom_point(alpha = 0.8) +
  facet_wrap(~ dataset_version, scales = "free") +
  labs(
    title = "Outward versus return phase duration",
    subtitle = "Comparison shown with and without extended trips > 250 km",
    x = "Outward duration (hours)",
    y = "Return duration (hours)",
    colour = "Year"
  ) +
  theme_paper()

# Jitter + boxplot of maximum range, shown for both dataset versions.
fig_max_range_jitter <- ggplot(trip_metrics_plot, aes(x = year_season, y = max_distance_colony_km)) +
  geom_jitter(width = 0.12, alpha = 0.7) +
  geom_boxplot(alpha = 0.3, outlier.shape = NA) +
  facet_wrap(~ dataset_version, scales = "free") +
  labs(
    title = "Distribution of maximum foraging range",
    subtitle = "Facets use independent axis scales; comparison shown with and without extended trips > 250 km",
    x = "Year",
    y = "Maximum distance from colony (km)"
  ) +
  theme_paper()
```

## 3.4. Figures for model-based results
```{r model-result-figures}
# Prepare labels for plotting model summaries.
gamma_year_effects_plot <- gamma_year_effects %>%
  mutate(
    response_label = recode(
      response,
      "trip_duration_hours" = "Trip duration (h)",
      "total_distance_km" = "Total distance (km)",
      "max_distance_colony_km" = "Max range from colony (km)"
    ),
    dataset_label = ifelse(dataset == "full", "Full dataset", "Without extended trip")
  )

# Figure 1: forest plot of year effect ratios from Gamma GLMM.
fig_gamma_year_effect_forest <- ggplot(
  gamma_year_effects_plot,
  aes(x = estimate_ratio, y = response_label, xmin = conf_low_ratio, xmax = conf_high_ratio, colour = dataset_label)
) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey45") +
  geom_pointrange(position = position_dodge(width = 0.45), linewidth = 0.5) +
  labs(
    title = "Year effect on trip metrics (Gamma GLMM)",
    subtitle = "Effect shown as ratio: 2026 / 2025; values > 1 indicate larger metrics in 2026",
    x = "Estimated ratio (2026 vs 2025)",
    y = NULL,
    colour = "Dataset"
  ) +
  theme_paper()

# Helper to obtain population-level predicted means and CIs from Gamma GLMM.
predict_gamma_means <- function(model, dataset_label, response_label) {
  coefs <- fixef(model)$cond
  vc <- as.matrix(vcov(model)$cond)

  term_year <- grep("^year_season", names(coefs), value = TRUE)
  if (length(term_year) != 1) {
    stop("Expected one year contrast term in fixed effects.")
  }

  # Linear predictor and CI for baseline (2025).
  eta_2025 <- coefs["(Intercept)"]
  se_2025 <- sqrt(vc["(Intercept)", "(Intercept)"])

  # Linear predictor and CI for 2026.
  eta_2026 <- coefs["(Intercept)"] + coefs[term_year]
  var_2026 <- vc["(Intercept)", "(Intercept)"] + vc[term_year, term_year] + 2 * vc["(Intercept)", term_year]
  se_2026 <- sqrt(var_2026)

  tibble(
    dataset = dataset_label,
    response = response_label,
    year_season = c("2025", "2026"),
    pred_mean = exp(c(eta_2025, eta_2026)),
    pred_lcl = exp(c(eta_2025 - 1.96 * se_2025, eta_2026 - 1.96 * se_2026)),
    pred_ucl = exp(c(eta_2025 + 1.96 * se_2025, eta_2026 + 1.96 * se_2026))
  )
}

gamma_predicted_means <- bind_rows(
  predict_gamma_means(gamma_duration_full, "Full dataset", "Trip duration (h)"),
  predict_gamma_means(gamma_distance_full, "Full dataset", "Total distance (km)"),
  predict_gamma_means(gamma_range_full, "Full dataset", "Max range from colony (km)"),
  predict_gamma_means(gamma_duration_trimmed, "Without extended trip", "Trip duration (h)"),
  predict_gamma_means(gamma_distance_trimmed, "Without extended trip", "Total distance (km)"),
  predict_gamma_means(gamma_range_trimmed, "Without extended trip", "Max range from colony (km)")
)

# Figure 2: model-predicted means by year.
fig_gamma_predicted_means <- ggplot(
  gamma_predicted_means,
  aes(x = year_season, y = pred_mean, ymin = pred_lcl, ymax = pred_ucl, colour = dataset)
) +
  geom_pointrange(position = position_dodge(width = 0.3), linewidth = 0.5) +
  facet_wrap(~ response, scales = "free_y") +
  labs(
    title = "Predicted trip metrics by year (Gamma GLMM)",
    subtitle = "Points are population-level means; bars show approximate 95% confidence intervals",
    x = "Year",
    y = "Predicted mean",
    colour = "Dataset"
  ) +
  theme_paper()
```
  
# 4. Save outputs
## 4.1 Save figures
```{r save-figures}
ggsave(file.path(output_dir, "fig_trip_duration_by_year_full_vs_trimmed.png"), fig_duration, width = 9, height = 5, dpi = 300)
ggsave(file.path(output_dir, "fig_total_distance_by_year_full_vs_trimmed.png"), fig_total_distance, width = 9, height = 5, dpi = 300)
ggsave(file.path(output_dir, "fig_max_range_by_year_full_vs_trimmed.png"), fig_max_range, width = 9, height = 5, dpi = 300)
ggsave(file.path(output_dir, "fig_departure_hour_by_year_full_vs_trimmed.png"), fig_departure, width = 9, height = 5, dpi = 300)
ggsave(file.path(output_dir, "fig_outward_vs_return_duration_full_vs_trimmed.png"), fig_outward_return, width = 9, height = 5, dpi = 300)
ggsave(file.path(output_dir, "fig_max_range_distribution_full_dataset.png"), fig_max_range_jitter, width = 7, height = 5, dpi = 300)
ggsave(file.path(output_dir, "fig_gamma_year_effect_forest.png"), fig_gamma_year_effect_forest, width = 9, height = 5, dpi = 300)
ggsave(file.path(output_dir, "fig_gamma_predicted_means_by_year.png"), fig_gamma_predicted_means, width = 10, height = 6, dpi = 300)
```

## 4.2. Save tables
```{r save-tables}
# Save the full trip-metrics table.
readr::write_csv(trip_metrics_full, file.path(output_dir, "trip_metrics_full_dataset.csv"))

# Save the trimmed trip-metrics table.
readr::write_csv(trip_metrics_trimmed, file.path(output_dir, "trip_metrics_without_extended_trip.csv"))

# Save the table of extended trips only.
readr::write_csv(extended_trip_table, file.path(output_dir, "extended_trip_table.csv"))

# Save summary tables.
readr::write_csv(trip_summary_by_year_full, file.path(output_dir, "trip_summary_by_year_full_dataset.csv"))
readr::write_csv(trip_summary_by_year_trimmed, file.path(output_dir, "trip_summary_by_year_without_extended_trip.csv"))
readr::write_csv(trip_summary_by_bird_full, file.path(output_dir, "trip_summary_by_bird_full_dataset.csv"))
readr::write_csv(trip_summary_by_bird_trimmed, file.path(output_dir, "trip_summary_by_bird_without_extended_trip.csv"))

# Save the year-comparison table.
readr::write_csv(year_comparison_table, file.path(output_dir, "year_comparison_table_full_vs_trimmed.csv"))
```

## 4.3. Save model objects
```{r save-models}
# Save Gaussian log-LMM models.
saveRDS(mod_duration_full, file.path(output_dir, "lmm_log_duration_full.rds"))
saveRDS(mod_distance_full, file.path(output_dir, "lmm_log_total_distance_full.rds"))
saveRDS(mod_range_full, file.path(output_dir, "lmm_log_max_range_full.rds"))

saveRDS(mod_duration_trimmed, file.path(output_dir, "lmm_log_duration_trimmed.rds"))
saveRDS(mod_distance_trimmed, file.path(output_dir, "lmm_log_total_distance_trimmed.rds"))
saveRDS(mod_range_trimmed, file.path(output_dir, "lmm_log_max_range_trimmed.rds"))

# Save final Gamma GLMM models.
saveRDS(gamma_duration_full, file.path(output_dir, "gamma_duration_full.rds"))
saveRDS(gamma_distance_full, file.path(output_dir, "gamma_total_distance_full.rds"))
saveRDS(gamma_range_full, file.path(output_dir, "gamma_max_range_full.rds"))

saveRDS(gamma_duration_trimmed, file.path(output_dir, "gamma_duration_trimmed.rds"))
saveRDS(gamma_distance_trimmed, file.path(output_dir, "gamma_total_distance_trimmed.rds"))
saveRDS(gamma_range_trimmed, file.path(output_dir, "gamma_max_range_trimmed.rds"))

# Save diagnostic and model-choice tables.
readr::write_csv(lmm_diagnostics, file.path(output_dir, "lmm_diagnostics_table.csv"))
readr::write_csv(gamma_diagnostics, file.path(output_dir, "gamma_diagnostics_table.csv"))
readr::write_csv(model_choice_table, file.path(output_dir, "model_choice_table.csv"))
readr::write_csv(gamma_year_effects, file.path(output_dir, "gamma_year_effects_table.csv"))
readr::write_csv(gamma_predicted_means, file.path(output_dir, "gamma_predicted_means_table.csv"))
```

## 4.4. Save an Excel workbook of the main outputs
```{r workbook}
# Create a workbook to collect the main outputs from this script in one place.
wb <- openxlsx::createWorkbook()

# Add the full trip-metrics table.
openxlsx::addWorksheet(wb, "trip_metrics_full")
openxlsx::writeData(wb, "trip_metrics_full", trip_metrics_full)

# Add the trimmed trip-metrics table.
openxlsx::addWorksheet(wb, "trip_metrics_trimmed")
openxlsx::writeData(wb, "trip_metrics_trimmed", trip_metrics_trimmed)

# Add the extended-trip table.
openxlsx::addWorksheet(wb, "extended_trip_table")
openxlsx::writeData(wb, "extended_trip_table", extended_trip_table)

# Add summary tables.
openxlsx::addWorksheet(wb, "summary_year_full")
openxlsx::writeData(wb, "summary_year_full", trip_summary_by_year_full)

openxlsx::addWorksheet(wb, "summary_year_trimmed")
openxlsx::writeData(wb, "summary_year_trimmed", trip_summary_by_year_trimmed)

openxlsx::addWorksheet(wb, "summary_bird_full")
openxlsx::writeData(wb, "summary_bird_full", trip_summary_by_bird_full)

openxlsx::addWorksheet(wb, "summary_bird_trimmed")
openxlsx::writeData(wb, "summary_bird_trimmed", trip_summary_by_bird_trimmed)

# Add the comparison table.
openxlsx::addWorksheet(wb, "year_comparison")
openxlsx::writeData(wb, "year_comparison", year_comparison_table)

# Add model-diagnostic and inference tables.
openxlsx::addWorksheet(wb, "lmm_diagnostics")
openxlsx::writeData(wb, "lmm_diagnostics", lmm_diagnostics)

openxlsx::addWorksheet(wb, "gamma_diagnostics")
openxlsx::writeData(wb, "gamma_diagnostics", gamma_diagnostics)

openxlsx::addWorksheet(wb, "model_choice")
openxlsx::writeData(wb, "model_choice", model_choice_table)

openxlsx::addWorksheet(wb, "gamma_year_effects")
openxlsx::writeData(wb, "gamma_year_effects", gamma_year_effects)

openxlsx::addWorksheet(wb, "gamma_predictions")
openxlsx::writeData(wb, "gamma_predictions", gamma_predicted_means)

# Save the workbook.
openxlsx::saveWorkbook(
  wb,
  file.path(output_dir, "01_trip_metrics_temporal_and_shape_outputs.xlsx"),
  overwrite = TRUE
)
```

# 5. Draft manuscript text
The following section produces manuscript-ready draft text directly from model outputs, so the rationale for model choice and core results remains documented in the analysis script.

```{r draft-methods-results}
# Build compact helper objects for text formatting.
bird_year_table <- stats_full %>%
  distinct(bird_id, year_season) %>%
  count(year_season, name = "n_birds")

n_birds_2025 <- bird_year_table$n_birds[bird_year_table$year_season == "2025"]
n_birds_2026 <- bird_year_table$n_birds[bird_year_table$year_season == "2026"]
n_cross_year_birds <- nrow(birds_in_multiple_years)

effect_line <- function(dataset_id, response_id, label) {
  row <- gamma_year_effects %>%
    filter(dataset == dataset_id, response == response_id) %>%
    slice(1)

  paste0(
    label, " (", dataset_id, "): ratio 2026/2025 = ",
    sprintf("%.2f", row$estimate_ratio),
    " [95% CI ", sprintf("%.2f", row$conf_low_ratio), ", ", sprintf("%.2f", row$conf_high_ratio), "]",
    ", p = ", sprintf("%.3f", row$p.value), "."
  )
}

methods_text <- paste0(
  "Trip-level foraging metrics (trip duration, total trip distance, and maximum distance from colony) were analysed ",
  "for two deployment seasons (2025 and 2026). Trips were nested within individual birds, and individual identity (bird_id) ",
  "was included as a random intercept in mixed-effects models to account for repeated trips per individual. ",
  "Because metrics were strictly positive and right-skewed, we first fitted Gaussian log-transformed linear mixed models ",
  "(log-LMM) and evaluated assumptions using residual diagnostics (performance::check_model and DHARMa tests). ",
  "Gaussian assumptions were repeatedly flagged by residual non-normality and heavy tails, therefore final inference was based ",
  "on Gamma GLMMs with log link (glmmTMB), fitted separately for full and trimmed datasets. ",
  "Fixed effects were interpreted as multiplicative year effects (2026/2025 ratios). ",
  "The trimmed dataset excluded extended trips with max_distance_colony_km > 250. ",
  "Sample sizes were ", nrow(stats_full), " trips in total (2025: ", sum(stats_full$year_season == '2025'),
  " trips across ", n_birds_2025, " birds; 2026: ", sum(stats_full$year_season == '2026'),
  " trips across ", n_birds_2026, " birds). ",
  "No birds were tracked in both years (n cross-year birds = ", n_cross_year_birds, ")."
)

results_text <- paste0(
  "Gamma GLMMs found no statistically significant year effect for any focal metric in either full or trimmed datasets. ",
  effect_line("full", "trip_duration_hours", "Trip duration"), " ",
  effect_line("full", "total_distance_km", "Total distance"), " ",
  effect_line("full", "max_distance_colony_km", "Maximum range"), " ",
  effect_line("trimmed", "trip_duration_hours", "Trip duration"), " ",
  effect_line("trimmed", "total_distance_km", "Total distance"), " ",
  effect_line("trimmed", "max_distance_colony_km", "Maximum range"), " ",
  "Direction and magnitude of effects were broadly stable between full and trimmed analyses, indicating that exclusion ",
  "of the single extended trip did not materially change the year-comparison conclusion."
)

cat("METHODS DRAFT:\n\n")
cat(methods_text, "\n\n")
cat("RESULTS DRAFT:\n\n")
cat(results_text, "\n")

# Save text draft to file for easy copy into manuscript notes.
writeLines(
  c(
    "Methods (draft):",
    methods_text,
    "",
    "Results (draft):",
    results_text
  ),
  con = file.path(output_dir, "draft_methods_results_trip_metrics.txt")
)
```

# End
```{r end}
cat("01_trip_metrics_temporal_and_shape completed successfully.\n")
cat("Outputs saved in:", output_dir, "\n")
```
