---
title: "04 - Trip metrics"
output:
  html_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
# Set global chunk options so code is shown and warnings/messages do not clutter the document.
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

# Purpose of the script
This script uses the analysis-ready objects created by `00_setup_and_load.Rmd`.

It focuses on the **trip-level description of foraging behaviour**, including:
- trip duration
- total distance travelled
- maximum range from the colony
- straightness
- departure and arrival timing
- outward and return phase duration
- initial bearings

A special feature of this script is that it produces outputs in **two versions**:

1. **Full dataset**: includes all trips, including the exceptional extended 2026 trip.
2. **Trimmed dataset**: excludes trips with `max_distance_colony_km > 250 km`.

This means you can:
- keep the biologically real extended trip in the study
- inspect how much it affects summary statistics
- compare 2025 and 2026 both with and without the extended trip

This script is designed to be a **standalone module**:
- it reads the clean inputs from script 03
- it creates its own tables, figures, and model outputs
- it saves everything into its own output folder

Maximum foraging range was calculated as the maximum great-circle distance between each location and the colony. Total distance travelled was calculated as the cumulative sum of great-circle distances between consecutive GPS locations.

# 1. Set up workflow
## 1.1. Packages
```{r packages, message = FALSE, warning = FALSE}
library(dplyr) # data manipulation
library(readr) # writing csv outputs
library(lubridate) # time handling
library(ggplot2) # plotting
library(openxlsx) # writing a workbook of outputs
library(lme4)         # mixed models
library(lmerTest)     # p-values for lmer models
library(performance)  # ICC and model checks
library(broom.mixed)  # tidy model summaries
library(glmmTMB)      # fallback GLMMs if needed
library(gplots)
library(patchwork)
library(gridExtra)
library(GGally)
library(mosaic)
library(lattice)
library(see)
library(effects)
library(DHARMa)
```

## 1.2. Project folders
```{r folders}
# Use the current working directory as the project root.
project_dir <- getwd()

# Folder containing the analysis-ready objects from script 00.
input_dir <- file.path(project_dir, "outputs", "03_setup_and_load")

# Output folder for this script.
output_dir <- file.path(project_dir, "outputs", "04_trip_metrics")

# Create the output folder if needed.
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
```

## 1.3. Analysis settings
```{r analysis-settings}
# Define the threshold used to flag exceptionally long trips.
# Any trip with maximum range above this threshold will be marked as an extended trip.
extended_trip_threshold_km <- 250
```

## 1.4. Helper functions
```{r helpers}
# Simple plotting theme used throughout this script.
theme_paper <- function() {
  theme_minimal() +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(colour = "grey85", linewidth = 0.4),
      axis.text = element_text(colour = "grey20"),
      plot.title = element_text(face = "bold", hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5),
      legend.position = "right"
    )
}

# Bearing helper:
# compute the initial bearing in degrees between two longitude/latitude points.
calculate_bearing_deg <- function(lon1, lat1, lon2, lat2) {
  lon1 <- lon1 * pi / 180
  lat1 <- lat1 * pi / 180
  lon2 <- lon2 * pi / 180
  lat2 <- lat2 * pi / 180

  y <- sin(lon2 - lon1) * cos(lat2)
  x <- cos(lat1) * sin(lat2) - sin(lat1) * cos(lat2) * cos(lon2 - lon1)

  bearing <- atan2(y, x) * 180 / pi
  (bearing + 360) %% 360
}
```

## 1.5. Load the analysis-ready inputs
```{r load-inputs}
# Load the clean point-level tracking object.
tracks <- readRDS(file.path(input_dir, "tracks_clean.rds"))

# Load the clean trip-level object.
trips <- readRDS(file.path(input_dir, "trips_clean.rds"))

# Load the clean QA object.
qa <- readRDS(file.path(input_dir, "qa_clean.rds"))
```

# 2. Trip analysis
## 2.1. Build the main trip-metrics table
```{r build-trip-metrics}
# Start from the trip table exported by your trip-detection workflow.
# Here we add a few convenient derived metrics.

trip_metrics <- trips %>%
  mutate(
    # Straightness index: a simple measure of how direct the trip was.
    straightness_index = max_distance_colony_km / total_distance_km, #1=perfectly straight trip, small values= tortuous movement

    # Convert departure time into decimal hours.
    departure_hour = hour(trip_start) + minute(trip_start) / 60,

    # Convert arrival time into decimal hours.
    arrival_hour = hour(trip_end) + minute(trip_end) / 60,

    # Compute trip midpoint.
    trip_midpoint = trip_start + (trip_end - trip_start) / 2,

    # Convert trip midpoint into decimal hours.
    trip_midpoint_hour = hour(trip_midpoint) + minute(trip_midpoint) / 60,

    # Flag exceptionally long trips using the fixed threshold.
    is_extended_trip = max_distance_colony_km > extended_trip_threshold_km
  ) %>%
  arrange(deploymentid, tag, trip_start)

glimpse(trip_metrics)
```

## 2.2. Inspect the extended trip(s)
```{r inspect-extended-trips}
# Print the trips flagged as extended so you can confirm what is being classified.
extended_trip_table <- trip_metrics %>%
  filter(is_extended_trip) %>%
  select(
    deploymentid, tag, bird_id, global_trip_id, foraging_trip,
    trip_start, trip_end, trip_duration_hours,
    total_distance_km, max_distance_colony_km
  )

extended_trip_table
```

## 2.3. Extract outward and return phase metrics from point-level trip data
```{r outward-return-phase}
# For each final trip, use the point-level data to:
# - locate the point of maximum distance from the colony
# - split the trip into an outward phase and a return phase
# - estimate initial bearings

trip_phase_metrics <- lapply(seq_len(nrow(trip_metrics)), function(i) {

  # Pull one trip row.
  tr <- trip_metrics[i, ]

  # Subset the point-level data for that trip.
  pts <- tracks %>%
    filter(global_trip_id == tr$global_trip_id) %>%
    arrange(datetime)

  # Skip trips with too few points to calculate phase metrics reliably.
  if (nrow(pts) < 3) return(NULL)

  # Identify the point of maximum distance from the colony.
  max_i <- which.max(pts$dist_to_colony_km)

  # Define the outward phase as the start of the trip up to the maximum-range point.
  outward_pts <- pts[1:max_i, ]

  # Define the return phase as the maximum-range point back to the end of the trip.
  return_pts <- pts[max_i:nrow(pts), ]

  # Estimate departure bearing from the first two outward points if available.
  dep_bearing <- if (nrow(outward_pts) >= 2) {
    calculate_bearing_deg(
      outward_pts$long[1], outward_pts$lat[1],
      outward_pts$long[2], outward_pts$lat[2]
    )
  } else {
    NA_real_
  }

  # Estimate return bearing from the first two return-phase points if available.
  ret_bearing <- if (nrow(return_pts) >= 2) {
    calculate_bearing_deg(
      return_pts$long[1], return_pts$lat[1],
      return_pts$long[2], return_pts$lat[2]
    )
  } else {
    NA_real_
  }

  # Return one row of phase metrics for this trip.
  tibble::tibble(
    global_trip_id = tr$global_trip_id,
    deploymentid = tr$deploymentid,
    tag = tr$tag,
    bird_id = tr$bird_id,
    year_season = tr$year_season,
    foraging_trip = tr$foraging_trip,
    departure_bearing_deg = dep_bearing,
    return_bearing_deg = ret_bearing,
    outward_duration_h = as.numeric(difftime(max(outward_pts$datetime), min(outward_pts$datetime), units = "hours")),
    return_duration_h = as.numeric(difftime(max(return_pts$datetime), min(return_pts$datetime), units = "hours")),
    max_distance_time = pts$datetime[max_i],
    max_distance_hour = hour(pts$datetime[max_i]) + minute(pts$datetime[max_i]) / 60 # straight-line (great-circle) distance between GPS point and colon
  )
}) %>%
  dplyr::bind_rows()

# Join these phase metrics back onto the main trip-metrics table.
trip_metrics <- trip_metrics %>%
  left_join(
    trip_phase_metrics,
    by = c("global_trip_id", "deploymentid", "tag", "bird_id", "year_season", "foraging_trip")
  )
```

## 2.4. Create the full and trimmed datasets
```{r full-and-trimmed-datasets}
# Keep a full copy of the trip-metrics table including all trips.
trip_metrics_full <- trip_metrics

# Create a trimmed version excluding extended trips above the threshold.
trip_metrics_trimmed <- trip_metrics %>%
  filter(!is_extended_trip)

# Build one combined plotting table so that full and trimmed outputs can be compared easily.
trip_metrics_plot <- bind_rows(
  trip_metrics_full %>% mutate(dataset_version = "Full dataset"),
  trip_metrics_trimmed %>% mutate(dataset_version = "Without extended trip")
)
```

## 2.5. Summary tables
```{r summary-tables}
# Summary by year for the full dataset.
trip_summary_by_year_full <- trip_metrics_full %>%
  group_by(year_season) %>%
  summarise(
    n_trips = n(),
    n_birds = n_distinct(bird_id),
    mean_duration_h = mean(trip_duration_hours, na.rm = TRUE),
    sd_duration_h = sd(trip_duration_hours, na.rm = TRUE),
    mean_total_distance_km = mean(total_distance_km, na.rm = TRUE),
    sd_total_distance_km = sd(total_distance_km, na.rm = TRUE),
    mean_max_range_km = mean(max_distance_colony_km, na.rm = TRUE),
    sd_max_range_km = sd(max_distance_colony_km, na.rm = TRUE),
    mean_straightness = mean(straightness_index, na.rm = TRUE),
    sd_straightness = sd(straightness_index, na.rm = TRUE),
    mean_departure_hour = mean(departure_hour, na.rm = TRUE),
    mean_arrival_hour = mean(arrival_hour, na.rm = TRUE),
    mean_outward_duration_h = mean(outward_duration_h, na.rm = TRUE),
    mean_return_duration_h = mean(return_duration_h, na.rm = TRUE),
    .groups = "drop"
  )

# Summary by year for the trimmed dataset.
trip_summary_by_year_trimmed <- trip_metrics_trimmed %>%
  group_by(year_season) %>%
  summarise(
    n_trips = n(),
    n_birds = n_distinct(bird_id),
    mean_duration_h = mean(trip_duration_hours, na.rm = TRUE),
    sd_duration_h = sd(trip_duration_hours, na.rm = TRUE),
    mean_total_distance_km = mean(total_distance_km, na.rm = TRUE),
    sd_total_distance_km = sd(total_distance_km, na.rm = TRUE),
    mean_max_range_km = mean(max_distance_colony_km, na.rm = TRUE),
    sd_max_range_km = sd(max_distance_colony_km, na.rm = TRUE),
    mean_straightness = mean(straightness_index, na.rm = TRUE),
    sd_straightness = sd(straightness_index, na.rm = TRUE),
    mean_departure_hour = mean(departure_hour, na.rm = TRUE),
    mean_arrival_hour = mean(arrival_hour, na.rm = TRUE),
    mean_outward_duration_h = mean(outward_duration_h, na.rm = TRUE),
    mean_return_duration_h = mean(return_duration_h, na.rm = TRUE),
    .groups = "drop"
  )

# Summary by bird for the full dataset.
trip_summary_by_bird_full <- trip_metrics_full %>%
  group_by(year_season, deploymentid, tag, bird_id) %>%
  summarise(
    n_trips = n(),
    mean_duration_h = mean(trip_duration_hours, na.rm = TRUE),
    mean_total_distance_km = mean(total_distance_km, na.rm = TRUE),
    mean_max_range_km = mean(max_distance_colony_km, na.rm = TRUE),
    mean_straightness = mean(straightness_index, na.rm = TRUE),
    mean_departure_hour = mean(departure_hour, na.rm = TRUE),
    mean_arrival_hour = mean(arrival_hour, na.rm = TRUE),
    .groups = "drop"
  )

# Summary by bird for the trimmed dataset.
trip_summary_by_bird_trimmed <- trip_metrics_trimmed %>%
  group_by(year_season, deploymentid, tag, bird_id) %>%
  summarise(
    n_trips = n(),
    mean_duration_h = mean(trip_duration_hours, na.rm = TRUE),
    mean_total_distance_km = mean(total_distance_km, na.rm = TRUE),
    mean_max_range_km = mean(max_distance_colony_km, na.rm = TRUE),
    mean_straightness = mean(straightness_index, na.rm = TRUE),
    mean_departure_hour = mean(departure_hour, na.rm = TRUE),
    mean_arrival_hour = mean(arrival_hour, na.rm = TRUE),
    .groups = "drop"
  )
```
    
  
# 3. Statistical analysis
## 3.1. Check datasets
  
Why mixed models? Trips are repeated within birds, so trips are not independent observations. A mixed model allows us to test for a year effect while accounting for repeated measures within bird by including `bird_id` as a random intercept.

### 3.1.1. Check if need data transformation
```{r}
# Keep only the variables needed for inference.
stats_full <- trip_metrics_full %>%
  select(
    bird_id,
    year_season,
    trip_duration_hours,
    total_distance_km,
    max_distance_colony_km
  ) %>%
  mutate(
    year_season = factor(year_season),
    log_trip_duration = log(trip_duration_hours),
    log_total_distance = log(total_distance_km),
    log_max_range = log(max_distance_colony_km),
    sqrt_trip_duration = sqrt(trip_duration_hours),
    sqrt_total_distance = sqrt(total_distance_km),
    sqrt_max_range = sqrt(max_distance_colony_km)
  )

stats_trimmed <- trip_metrics_trimmed %>%
  select(
    bird_id,
    year_season,
    trip_duration_hours,
    total_distance_km,
    max_distance_colony_km
  ) %>%
  mutate(
    year_season = factor(year_season),
    log_trip_duration = log(trip_duration_hours),
    log_total_distance = log(total_distance_km),
    log_max_range = log(max_distance_colony_km),
    sqrt_trip_duration = sqrt(trip_duration_hours),
    sqrt_total_distance = sqrt(total_distance_km),
    sqrt_max_range = sqrt(max_distance_colony_km)
  )
```
  
  
Dotchart of our key stat variables
```{r, warning=FALSE, message=FALSE, tidy=TRUE}
variables <- cbind(stats_full$trip_duration_hours, stats_full$total_distance_km,
                   stats_full$max_distance_colony_km, stats_trimmed$trip_duration_hours,
                   stats_trimmed$total_distance_km, stats_trimmed$max_distance_colony_km)

colnames(variables) <- c("duration - full", "total dist - full", "max range - full",
                         "duration - trimmed", "total dist - trimmed", "max range - trimmed")

dotplot(as.matrix(variables),
        groups=FALSE,
        strip=strip.custom(bg="white",
                           par.strip.text=list(cex=0.8)),
        scales=list(x=list(relation="free"),
                    y=list(realtion="free"),
                    draw=FALSE),
        col=1, cex=0.5, pch=16,
        xlab= "value of the variable",
        ylab= "Order of the data from text file")
```
It looks like there are two outliers for the full dataset, which are most likely coming from the Argentinian trip and one outlier for the trimmed dataset, most likely from the longer trip to the south (queen Charlotte's bay).
  
  
If we apply a log or sqrt transformation, can we account for those outliers?
Full dataset:
```{r, warning=FALSE, message=FALSE, tidy=TRUE}
variablesduration_full <- cbind(stats_full$trip_duration_hours,
                                stats_full$log_trip_duration,
                                stats_full$sqrt_trip_duration)

variablestotaldist_full <- cbind(stats_full$total_distance_km,
                                 stats_full$log_total_distance, 
                                 stats_full$sqrt_total_distance)

variablesmaxrange_full <- cbind(stats_trimmed$max_distance_colony_km,
                                stats_trimmed$log_max_range,
                                stats_trimmed$sqrt_max_range)

colnames(variablesduration_full) <- c("full - duration", "full - duration log", "full - duration sqrt")
colnames(variablestotaldist_full) <- c("full - dist", "full - dist log", "full - dist sqrt")
colnames(variablesmaxrange_full) <- c("full - max range", "full - max range log", "full - max range sqrt")

par(mfrow=c(3, 3))
dotplot(as.matrix(variablesduration_full),
        groups=FALSE,
        strip=strip.custom(bg="white",
                           par.strip.text=list(cex=0.8)),
        scales=list(x=list(relation="free"),
                    y=list(realtion="free"),
                    draw=FALSE),
        col=1, cex=0.5, pch=16,
        xlab= "value of the variable",
        ylab= "Order of the data from text file")


dotplot(as.matrix(variablestotaldist_full),
        groups=FALSE,
        strip=strip.custom(bg="white",
                           par.strip.text=list(cex=0.8)),
        scales=list(x=list(relation="free"),
                    y=list(realtion="free"),
                    draw=FALSE),
        col=1, cex=0.5, pch=16,
        xlab= "value of the variable",
        ylab= "Order of the data from text file")


dotplot(as.matrix(variablesmaxrange_full),
        groups=FALSE,
        strip=strip.custom(bg="white",
                           par.strip.text=list(cex=0.8)),
        scales=list(x=list(relation="free"),
                    y=list(realtion="free"),
                    draw=FALSE),
        col=1, cex=0.5, pch=16,
        xlab= "value of the variable",
        ylab= "Order of the data from text file")
```
Trimmed dataset
```{r, warning=FALSE, message=FALSE, tidy=TRUE}
variablesduration_trimmed <- cbind(stats_trimmed$trip_duration_hours,
                                   stats_trimmed$log_trip_duration,
                                   stats_trimmed$sqrt_trip_duration)

variablestotaldist_trimmed <- cbind(stats_trimmed$total_distance_km,
                                    stats_trimmed$log_total_distance,
                                    stats_trimmed$sqrt_total_distance)

variablesmaxrange_trimmed <- cbind(stats_trimmed$max_distance_colony_km,
                                   stats_trimmed$log_max_range,
                                   stats_trimmed$sqrt_max_range)

colnames(variablesduration_trimmed) <- c("trimmed - duration", 
                                         "trimmed - duration log", 
                                         "trimmed - duration sqrt")

colnames(variablestotaldist_trimmed) <- c("trimmed - dist", 
                                          "trimmed - dist log",
                                          "trimmed - dist sqrt")

colnames(variablesmaxrange_trimmed) <- c("trimmed - max range",
                                         "trimmed - max range log",
                                         "trimmed - max range sqrt")

par(mfrow=c(1, 2))
dotplot(as.matrix(variablesduration_trimmed),
        groups=FALSE,
        strip=strip.custom(bg="white",
                           par.strip.text=list(cex=0.8)),
        scales=list(x=list(relation="free"),
                    y=list(realtion="free"),
                    draw=FALSE),
        col=1, cex=0.5, pch=16,
        xlab= "value of the variable",
        ylab= "Order of the data from text file")


dotplot(as.matrix(variablestotaldist_trimmed),
        groups=FALSE,
        strip=strip.custom(bg="white",
                           par.strip.text=list(cex=0.8)),
        scales=list(x=list(relation="free"),
                    y=list(realtion="free"),
                    draw=FALSE),
        col=1, cex=0.5, pch=16,
        xlab= "value of the variable",
        ylab= "Order of the data from text file")


dotplot(as.matrix(variablesmaxrange_trimmed),
        groups=FALSE,
        strip=strip.custom(bg="white",
                           par.strip.text=list(cex=0.8)),
        scales=list(x=list(relation="free"),
                    y=list(realtion="free"),
                    draw=FALSE),
        col=1, cex=0.5, pch=16,
        xlab= "value of the variable",
        ylab= "Order of the data from text file")
```
  
For both the full and trimmed datasets, it seems that log transforming helps a little to bridge the space between the main points and outliers. log-transformed variables will be used for modelling.

### 3.1.2. Data simple stat summary
```{r}
desc_full <- stats_full %>%
  group_by(year_season) %>%
  summarise(
    n_trips = n(),
    n_birds = n_distinct(bird_id),
    mean_duration = mean(trip_duration_hours, na.rm = TRUE),
    sd_duration = sd(trip_duration_hours, na.rm = TRUE),
    mean_total_distance = mean(total_distance_km, na.rm = TRUE),
    sd_total_distance = sd(total_distance_km, na.rm = TRUE),
    mean_max_range = mean(max_distance_colony_km, na.rm = TRUE),
    sd_max_range = sd(max_distance_colony_km, na.rm = TRUE),
    .groups = "drop"
  )

desc_trimmed <- stats_trimmed %>%
  group_by(year_season) %>%
  summarise(
    n_trips = n(),
    n_birds = n_distinct(bird_id),
    mean_duration = mean(trip_duration_hours, na.rm = TRUE),
    sd_duration = sd(trip_duration_hours, na.rm = TRUE),
    mean_total_distance = mean(total_distance_km, na.rm = TRUE),
    sd_total_distance = sd(total_distance_km, na.rm = TRUE),
    mean_max_range = mean(max_distance_colony_km, na.rm = TRUE),
    sd_max_range = sd(max_distance_colony_km, na.rm = TRUE),
    .groups = "drop"
  )

desc_full
desc_trimmed
```

## 3.2. Run LME
### 3.2.1. LME full dataset
```{r}
# trip duration
mod_duration_full <- lmer(
  log_trip_duration ~ year_season + (1 | bird_id), 
  data = stats_full,
  REML = TRUE)

# total distance
mod_distance_full <- lmer(
  log_total_distance ~ year_season + (1 | bird_id), 
  data = stats_full,
  REML = TRUE)

# maximum distance to colony
mod_range_full <- lmer(
  log_max_range ~ year_season + (1 | bird_id), 
  data = stats_full,
  REML = TRUE)
```


### 3.2.2. LME trimmed dataset
```{r}
# trip duration
mod_duration_trimmed <- lmer(
  log_trip_duration ~ year_season + (1| bird_id), 
  data = stats_trimmed,
  REML = TRUE)

mod_distance_trimmed <- lmer(
  log_total_distance ~ year_season + (1 | bird_id), 
  data = stats_trimmed,
  REML = TRUE)

mod_range_trimmed <- lmer(
  log_max_range ~ year_season + (1 | bird_id),
  data = stats_trimmed,
  REML = TRUE)
```

Full dataset model diagnostics, check assumptions
```{r}
check_model(mod_duration_full)
check_model(mod_distance_full)
check_model(mod_range_full)


sim_duration_full <- simulateResiduals(mod_duration_full)
plot(sim_duration_full)

sim_distance_full <- simulateResiduals(mod_distance_full)
plot(sim_distance_full)

sim_range_full <- simulateResiduals(mod_range_full)
plot(sim_range_full)
```

Trimmed dataset model diagnostics, check assumptions
```{r}
check_model(mod_duration_trimmed)
check_model(mod_distance_trimmed)
check_model(mod_range_trimmed)


sim_duration_trimmed <- simulateResiduals(mod_duration_trimmed)
plot(sim_duration_trimmed)

sim_distance_trimmed <- simulateResiduals(mod_distance_trimmed)
plot(sim_distance_trimmed)

sim_range_trimmed <- simulateResiduals(mod_range_trimmed)
plot(sim_range_trimmed)
```
From the diagnostic plots, it is clear that issues are present when looking at the trip duration and total distance variables. Specifically, we can identify some non-normally distributed residuals under the Gaussian model and the presence of heteroscedasticity. Therefore, I will fit a Gamma mixed-effects model (GAMM) with a `log link` function, which is appropriate for positive, right-skewed data.
  


  




















## 3.2. Figures comparing full and trimmed versions
```{r figures}
# Boxplot of trip duration by year, shown for both dataset versions.
fig_duration <- ggplot(trip_metrics_plot, aes(x = year_season, y = trip_duration_hours)) +
  geom_boxplot() +
  facet_wrap(~ dataset_version, scales = "free") +
  labs(
    title = "Trip duration by year",
    subtitle = "Comparison shown with and without extended trips > 250 km",
    x = "Year",
    y = "Trip duration (hours)"
  ) +
  theme_paper()

# Boxplot of total trip distance by year, shown for both dataset versions.
fig_total_distance <- ggplot(trip_metrics_plot, aes(x = year_season, y = total_distance_km)) +
  geom_boxplot() +
  facet_wrap(~ dataset_version, scales = "free") +
  labs(
    title = "Total trip distance by year",
    subtitle = "Comparison shown with and without extended trips > 250 km",
    x = "Year",
    y = "Total distance (km)"
  ) +
  theme_paper()

# Boxplot of maximum range by year, shown for both dataset versions.
fig_max_range <- ggplot(trip_metrics_plot, aes(x = year_season, y = max_distance_colony_km)) +
  geom_boxplot() +
  facet_wrap(~ dataset_version, scales = "free") +
  labs(
    title = "Maximum range by year",
    subtitle = "Comparison shown with and without extended trips > 250 km",
    x = "Year",
    y = "Maximum distance from colony (km)"
  ) +
  theme_paper()

# Boxplot of departure hour by year, shown for both dataset versions.
fig_departure <- ggplot(trip_metrics_plot, aes(x = year_season, y = departure_hour)) +
  geom_boxplot() +
  facet_wrap(~ dataset_version, scales = "free") +
  labs(
    title = "Departure hour by year",
    subtitle = "Comparison shown with and without extended trips > 250 km",
    x = "Year",
    y = "Departure hour"
  ) +
  theme_paper()

# Scatterplot of outward and return phase duration, shown for both dataset versions.
fig_outward_return <- ggplot(trip_metrics_plot, aes(x = outward_duration_h, y = return_duration_h, colour = year_season)) +
  geom_point(alpha = 0.8) +
  facet_wrap(~ dataset_version, scales = "free") +
  labs(
    title = "Outward versus return phase duration",
    subtitle = "FComparison shown with and without extended trips > 250 km",
    x = "Outward duration (hours)",
    y = "Return duration (hours)",
    colour = "Year"
  ) +
  theme_paper()

# Jitter + boxplot of maximum range, shown for both dataset versions.
fig_max_range_jitter <- ggplot(trip_metrics_plot, aes(x = year_season, y = max_distance_colony_km)) +
  geom_jitter(width = 0.12, alpha = 0.7) +
  geom_boxplot(alpha = 0.3, outlier.shape = NA) +
  facet_wrap(~ dataset_version, scales = "free") +
  labs(
    title = "Distribution of maximum foraging range",
    subtitle = "Facets use independent axis scales; comparison shown with and without extended trips > 250 km",
    x = "Year",
    y = "Maximum distance from colony (km)"
  ) +
  theme_paper()
```
  
# 4. Save outputs
## 4.1 Save figures
```{r save-figures}
ggsave(file.path(output_dir, "fig_trip_duration_by_year_full_vs_trimmed.png"), fig_duration, width = 9, height = 5, dpi = 300)
ggsave(file.path(output_dir, "fig_total_distance_by_year_full_vs_trimmed.png"), fig_total_distance, width = 9, height = 5, dpi = 300)
ggsave(file.path(output_dir, "fig_max_range_by_year_full_vs_trimmed.png"), fig_max_range, width = 9, height = 5, dpi = 300)
ggsave(file.path(output_dir, "fig_departure_hour_by_year_full_vs_trimmed.png"), fig_departure, width = 9, height = 5, dpi = 300)
ggsave(file.path(output_dir, "fig_outward_vs_return_duration_full_vs_trimmed.png"), fig_outward_return, width = 9, height = 5, dpi = 300)
ggsave(file.path(output_dir, "fig_max_range_distribution_full_dataset.png"), fig_max_range_jitter, width = 7, height = 5, dpi = 300)
```

## 4.2. Save tables
```{r save-tables}
# Save the full trip-metrics table.
readr::write_csv(trip_metrics_full, file.path(output_dir, "trip_metrics_full_dataset.csv"))

# Save the trimmed trip-metrics table.
readr::write_csv(trip_metrics_trimmed, file.path(output_dir, "trip_metrics_without_extended_trip.csv"))

# Save the table of extended trips only.
readr::write_csv(extended_trip_table, file.path(output_dir, "extended_trip_table.csv"))

# Save summary tables.
readr::write_csv(trip_summary_by_year_full, file.path(output_dir, "trip_summary_by_year_full_dataset.csv"))
readr::write_csv(trip_summary_by_year_trimmed, file.path(output_dir, "trip_summary_by_year_without_extended_trip.csv"))
readr::write_csv(trip_summary_by_bird_full, file.path(output_dir, "trip_summary_by_bird_full_dataset.csv"))
readr::write_csv(trip_summary_by_bird_trimmed, file.path(output_dir, "trip_summary_by_bird_without_extended_trip.csv"))

# Save the year-comparison table.
readr::write_csv(year_comparison_table, file.path(output_dir, "year_comparison_table_full_vs_trimmed.csv"))
```

## 4.3. Save model objects
```{r save-models}
# Save the simple linear models for the full dataset.
saveRDS(lm_duration_full, file.path(output_dir, "lm_duration_by_year_full_dataset.rds"))
saveRDS(lm_max_range_full, file.path(output_dir, "lm_max_range_by_year_full_dataset.rds"))

# Save the simple linear models for the trimmed dataset.
saveRDS(lm_duration_trimmed, file.path(output_dir, "lm_duration_by_year_without_extended_trip.rds"))
saveRDS(lm_max_range_trimmed, file.path(output_dir, "lm_max_range_by_year_without_extended_trip.rds"))
```

## 4.4. Save an Excel workbook of the main outputs
```{r workbook}
# Create a workbook to collect the main outputs from this script in one place.
wb <- openxlsx::createWorkbook()

# Add the full trip-metrics table.
openxlsx::addWorksheet(wb, "trip_metrics_full")
openxlsx::writeData(wb, "trip_metrics_full", trip_metrics_full)

# Add the trimmed trip-metrics table.
openxlsx::addWorksheet(wb, "trip_metrics_trimmed")
openxlsx::writeData(wb, "trip_metrics_trimmed", trip_metrics_trimmed)

# Add the extended-trip table.
openxlsx::addWorksheet(wb, "extended_trip_table")
openxlsx::writeData(wb, "extended_trip_table", extended_trip_table)

# Add summary tables.
openxlsx::addWorksheet(wb, "summary_year_full")
openxlsx::writeData(wb, "summary_year_full", trip_summary_by_year_full)

openxlsx::addWorksheet(wb, "summary_year_trimmed")
openxlsx::writeData(wb, "summary_year_trimmed", trip_summary_by_year_trimmed)

openxlsx::addWorksheet(wb, "summary_bird_full")
openxlsx::writeData(wb, "summary_bird_full", trip_summary_by_bird_full)

openxlsx::addWorksheet(wb, "summary_bird_trimmed")
openxlsx::writeData(wb, "summary_bird_trimmed", trip_summary_by_bird_trimmed)

# Add the comparison table.
openxlsx::addWorksheet(wb, "year_comparison")
openxlsx::writeData(wb, "year_comparison", year_comparison_table)

# Save the workbook.
openxlsx::saveWorkbook(
  wb,
  file.path(output_dir, "01_trip_metrics_temporal_and_shape_outputs.xlsx"),
  overwrite = TRUE
)
```

# End
```{r end}
cat("01_trip_metrics_temporal_and_shape completed successfully.\n")
cat("Outputs saved in:", output_dir, "\n")
```
