---
title: "05 - Habitat Use (Autocorrelation-Aware KDE + Environmental Covariates)"
output:
  html_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

# Purpose
This script is a **separate workflow** from `04_trip_metrics.Rmd`, focused on:

1. estimating space use with kernel-type utilization distributions while accounting for serial autocorrelation in tracking data;
2. fitting environmental-covariate habitat models with an autocorrelation-aware likelihood.

The workflow is designed for seabird central-place foraging tracks and uses the cleaned inputs from `outputs/03_setup_and_load`.

# Method choice (literature-informed)
Why this approach:

- **Classical KDE is widely used in seabird conservation workflows** (e.g., `track2KBA` / mIBA-style analyses), but requires careful smoothing and representativeness checks.
- **Tracking data are typically autocorrelated**, and IID KDE can be biased on modern high-frequency data.
- **AKDE (autocorrelated KDE)** from `ctmm` is specifically designed for autocorrelated telemetry and is generally preferred for robust home-range/UD estimation.
- For covariates, **integrated RSF/iRSF with autocorrelation-adjusted likelihood** is preferable to naive used-vs-available models that ignore serial dependence.

References used to guide this script:

- Fleming et al. 2015 (AKDE): https://doi.org/10.1890/14-2010.1
- Noonan et al. 2019 (autocorrelation bias in home-range estimators): https://doi.org/10.1002/ecm.1344
- ctmm package / AKDE / RSF docs: https://ctmm-initiative.github.io/ctmm/
- Alston et al. 2023 (autocorrelation-informed RSF weighting): https://doi.org/10.1111/2041-210X.14025
- track2KBA paper (seabird-oriented KDE workflow context): https://ueaeprints.uea.ac.uk/81330/1/Accepted_Manuscript.pdf
- Example seabird habitat-model workflow combining distributions + environmental covariates:
  https://link.springer.com/article/10.1007/s00300-018-2404-4

# 1. Setup
## 1.1 Packages
```{r packages}
required_packages <- c(
  "dplyr", "tidyr", "purrr", "stringr", "lubridate",
  "ggplot2", "readr", "sf", "terra", "raster", "ctmm"
)


missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]

if (length(missing_packages) > 0) {
  stop(
    paste0(
      "Missing required packages: ",
      paste(missing_packages, collapse = ", "),
      ".\nInstall them before running this script."
    )
  )
}

library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(lubridate)
library(ggplot2)
library(readr)
library(sf)
library(terra)
library(raster)
library(ctmm)
```

## 1.2 Folders
```{r folders}
project_dir <- getwd()
input_dir <- file.path(project_dir, "outputs", "03_setup_and_load")
output_dir <- file.path(project_dir, "outputs", "05_habitat_use_kde_environment")
fig_dir <- file.path(output_dir, "figures")
table_dir <- file.path(output_dir, "tables")
model_dir <- file.path(output_dir, "models")
akde_raster_dir <- file.path(output_dir, "akde_rasters")
akde_vector_dir <- file.path(output_dir, "akde_vectors")
env_input_dir <- file.path(output_dir, "inputs_env")
map_input_dir <- file.path(output_dir, "inputs_map")
env_projected_dir <- file.path(output_dir, "env_projected")
rsf_map_dir <- file.path(output_dir, "rsf_maps")

dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(table_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(model_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(akde_raster_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(akde_vector_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(env_input_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(map_input_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(env_projected_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(rsf_map_dir, recursive = TRUE, showWarnings = FALSE)
```

## 1.3 User settings
```{r settings}
# Use at-sea locations only for habitat-use inference.
at_sea_state_label <- "OUT"

# Optional land polygon (recommended for publication-style maps).
# Put a coastline/land polygon in inputs_map and set this path.
# Supported formats include .gpkg / .shp.
land_vector_path <- if (file.exists(file.path(project_dir, "FI_minusBeauchene.shp"))) {
  file.path(project_dir, "FI_minusBeauchene.shp")
} else {
  file.path(map_input_dir, "falklands_land.gpkg")
}

# Dataset scope:
# - "trimmed": excludes trips whose max_distance_colony_km > extended_trip_threshold_km
# - "full": keeps all trips
analysis_dataset <- "trimmed"
extended_trip_threshold_km <- 250

# Confidence and utilization levels for area summaries.
ci_level <- 0.95
ud_levels <- c(0.50, 0.95)

# Number of individual variograms to plot as diagnostics.
n_variogram_examples <- 6

# Speed controls for AKDE (section 3.3)
# These settings strongly affect runtime.
akde_use_model_selection <- FALSE  # FALSE = ctmm.fit on initial guess (faster), TRUE = ctmm.select (slower, more exhaustive)
akde_use_weights <- FALSE          # TRUE is more robust to sampling bias but slower
akde_res <- 6                      # Smaller = coarser raster, faster (default ctmm is often finer)
akde_error <- 0.01                 # Higher tolerated numerical error = faster (default often 0.001)

# Force all AKDEs onto a fixed grid resolution (meters) for compatibility when averaging UDs.
# Set to NULL to let ctmm choose individual-specific grids (can cause mean.UD grid mismatch).
akde_grid_dr_m <- c(5000, 5000)

# Optional thinning for very high-frequency tracks.
# Set to NULL to disable, or e.g. 10 for ~10-minute thinning.
thin_to_minutes <- NULL

# Optional minimum fixes required per individual to fit AKDE.
min_fixes_per_individual <- 30

# Statistical resampling settings for overlap inference.
# Increase to 1000+ for final inference if compute time allows.
n_overlap_boot <- 300
n_overlap_perm <- 300

# Publication map settings.
year_map_contours <- c(0.50, 0.75, 0.95)
map_palette <- "C"
```

# 2. Load and prepare tracking data
```{r load-prepare}
tracks <- readRDS(file.path(input_dir, "tracks_clean.rds"))
trips <- readRDS(file.path(input_dir, "trips_clean.rds"))

# Select full or trimmed trip IDs.
if (analysis_dataset == "trimmed") {
  selected_trip_ids <- trips %>%
    filter(max_distance_colony_km <= extended_trip_threshold_km) %>%
    pull(global_trip_id) %>%
    unique()
} else if (analysis_dataset == "full") {
  selected_trip_ids <- unique(trips$global_trip_id)
} else {
  stop("analysis_dataset must be either 'trimmed' or 'full'.")
}

tracks_at_sea <- tracks %>%
  filter(global_trip_id %in% selected_trip_ids) %>%
  filter(state == at_sea_state_label) %>%
  mutate(
    datetime = as.POSIXct(datetime, tz = "UTC"),
    year_season = as.character(year_season),
    bird_id = as.character(bird_id)
  ) %>%
  filter(!is.na(datetime), !is.na(long), !is.na(lat), !is.na(bird_id), !is.na(year_season)) %>%
  arrange(year_season, bird_id, datetime) %>%
  distinct(year_season, bird_id, datetime, .keep_all = TRUE)

# Optional regular-time thinning for speed and comparability across years.
if (!is.null(thin_to_minutes)) {
  thin_seconds <- thin_to_minutes * 60
  tracks_at_sea <- tracks_at_sea %>%
    mutate(time_bin = floor(as.numeric(datetime) / thin_seconds)) %>%
    group_by(year_season, bird_id, time_bin) %>%
    slice(1) %>%
    ungroup() %>%
    select(-time_bin)
}

glimpse(tracks_at_sea)

tracking_counts <- tracks_at_sea %>%
  group_by(year_season) %>%
  summarise(
    n_points = n(),
    n_individuals = n_distinct(bird_id),
    .groups = "drop"
  )

tracking_counts

readr::write_csv(tracking_counts, file.path(table_dir, "tracking_counts_after_filters.csv"))
```

## 2.2 Optional map layers (land + colony point)
```{r map-layers}
# Colony point from available metadata (for map reference).
colony_ll <- tracks_at_sea %>%
  summarise(
    colony_lon = median(colony_lon, na.rm = TRUE),
    colony_lat = median(colony_lat, na.rm = TRUE)
  )

colony_sf_ll <- sf::st_as_sf(colony_ll, coords = c("colony_lon", "colony_lat"), crs = 4326)

# Optional land polygon.
land_available <- file.exists(land_vector_path)
land_sf_raw <- NULL
if (land_available) {
  land_sf_raw <- sf::st_read(land_vector_path, quiet = TRUE)
}
```

## 2.1 Sampling-interval diagnostics (important for interpretation)
```{r sampling-interval}
sampling_diag <- tracks_at_sea %>%
  group_by(bird_id, year_season) %>%
  arrange(datetime, .by_group = TRUE) %>%
  mutate(dt_min = as.numeric(difftime(datetime, lag(datetime), units = "mins"))) %>%
  ungroup() %>%
  filter(!is.na(dt_min), dt_min > 0, dt_min < 24 * 60)

sampling_summary <- sampling_diag %>%
  group_by(year_season) %>%
  summarise(
    n_intervals = n(),
    mean_dt_min = mean(dt_min, na.rm = TRUE),
    median_dt_min = median(dt_min, na.rm = TRUE),
    q1_dt_min = quantile(dt_min, 0.25, na.rm = TRUE),
    q3_dt_min = quantile(dt_min, 0.75, na.rm = TRUE),
    .groups = "drop"
  )

sampling_summary

readr::write_csv(sampling_summary, file.path(table_dir, "sampling_interval_summary_by_year.csv"))
```

```{r sampling-interval-fig}
fig_sampling <- ggplot(sampling_diag, aes(x = year_season, y = dt_min)) +
  geom_boxplot(outlier.alpha = 0.3) +
  labs(
    title = "Sampling interval by year",
    subtitle = "Differences in fix interval can affect path-based metrics and should be reported",
    x = "Year",
    y = "Time between fixes (minutes)"
  ) +
  theme_minimal()

fig_sampling

ggsave(
  file.path(fig_dir, "fig_sampling_interval_by_year.png"),
  fig_sampling,
  width = 8,
  height = 5,
  dpi = 300
)
```

# 3. Autocorrelation-aware KDE using ctmm (AKDE)
## 3.1 Build telemetry objects
```{r telemetry}
# ctmm::as.telemetry expects Movebank-style column names.
# Keep an explicit lookup table instead of parsing IDs with string splits later.
id_lookup <- tracks_at_sea %>%
  distinct(bird_id, year_season) %>%
  arrange(year_season, bird_id)
id_year_lookup <- setNames(id_lookup$year_season, id_lookup$bird_id)

ctmm_df <- tracks_at_sea %>%
  transmute(
    individual.local.identifier = bird_id,
    timestamp = format(datetime, "%Y-%m-%d %H:%M:%S"),
    location.long = long,
    location.lat = lat,
    year_season = year_season,
    bird_id = bird_id
  )

telemetry_list <- ctmm::as.telemetry(
  ctmm_df,
  timeformat = "%Y-%m-%d %H:%M:%S",
  timezone = "UTC",
  keep = TRUE,
  drop = FALSE
)

if (!is.list(telemetry_list)) {
  telemetry_list <- list(telemetry_list)
}

length(telemetry_list)
```

## 3.2 Variogram diagnostics for autocorrelation
```{r variograms}
example_ids <- names(telemetry_list)[seq_len(min(n_variogram_examples, length(telemetry_list)))]

for (id in example_ids) {
  vg <- ctmm::variogram(telemetry_list[[id]])
  png(file.path(fig_dir, paste0("fig_variogram_", make.names(id), ".png")), width = 900, height = 600, res = 120)
  plot(vg, main = paste("Variogram:", id))
  dev.off()
}
```

## 3.3 Fit movement models and AKDE per individual
```{r fit-akde}
fit_akde_one <- function(tel, id) {
  n_fix <- tryCatch(nrow(as.data.frame(tel)), error = function(e) NA_integer_)

  if (!is.na(n_fix) && n_fix < min_fixes_per_individual) {
    return(list(
      id = id,
      fit = NULL,
      ud = NULL,
      n_fix = n_fix,
      runtime_sec = NA_real_,
      status = "skipped_too_few_fixes"
    ))
  }

  t0 <- Sys.time()
  guess <- ctmm::ctmm.guess(tel, interactive = FALSE)

  fit <- if (akde_use_model_selection) {
    ctmm::ctmm.select(tel, guess, trace = FALSE)
  } else {
    ctmm::ctmm.fit(tel, guess)
  }

  grid_arg <- if (is.null(akde_grid_dr_m)) {
    NULL
  } else {
    list(dr = akde_grid_dr_m, align.to.origin = TRUE)
  }

  ud <- ctmm::akde(
    tel,
    fit,
    weights = akde_use_weights,
    res = akde_res,
    error = akde_error,
    grid = grid_arg
  )

  runtime_sec <- as.numeric(difftime(Sys.time(), t0, units = "secs"))

  list(
    id = id,
    fit = fit,
    ud = ud,
    n_fix = n_fix,
    runtime_sec = runtime_sec,
    status = "ok"
  )
}

akde_results <- purrr::imap(telemetry_list, ~purrr::possibly(fit_akde_one, otherwise = NULL)(.x, .y))
akde_results <- akde_results[!vapply(akde_results, is.null, logical(1))]

akde_runtime_table <- purrr::map_dfr(akde_results, ~tibble(
  individual_id = .x$id,
  n_fix = .x$n_fix,
  runtime_sec = .x$runtime_sec,
  status = .x$status
)) %>%
  left_join(id_lookup %>% rename(individual_id = bird_id), by = "individual_id") %>%
  mutate(bird_id = individual_id, .after = individual_id) %>%
  relocate(year_season, bird_id, .after = individual_id)

readr::write_csv(akde_runtime_table, file.path(table_dir, "akde_runtime_by_individual.csv"))

akde_runtime_table %>%
  arrange(desc(runtime_sec))

akde_results <- akde_results[vapply(akde_results, function(x) identical(x$status, "ok"), logical(1))]

length(akde_results)
```

## 3.4 Summarise AKDE areas and export outputs
```{r akde-summary-export}
if (length(akde_results) == 0) {
  stop("No successful AKDE fits. Check akde_runtime_by_individual.csv and speed/filter settings.")
}

# Robust extractor for ctmm summary output across versions.
# Some versions return a list with $CI matrix, others return matrix/vector directly.
extract_area_ci <- function(ud_object, ci_level, ud_level) {
  sm <- summary(ud_object, level = ci_level, level.UD = ud_level, units = TRUE)
  ci <- if (is.list(sm) && !is.null(sm$CI)) sm$CI else sm

  if (is.data.frame(ci)) {
    ci <- as.matrix(ci)
  }

  if (is.matrix(ci)) {
    idx <- 1
    if (!is.null(rownames(ci))) {
      area_idx <- which(grepl("^area", rownames(ci), ignore.case = TRUE))
      if (length(area_idx) > 0) idx <- area_idx[1]
    }
    row <- ci[idx, , drop = TRUE]
    nm <- names(row)
  } else {
    row <- ci
    nm <- names(row)
  }

  pick <- function(x, candidates, fallback_index = 1) {
    hit <- intersect(candidates, names(x))
    if (length(hit) > 0) return(as.numeric(x[[hit[1]]]))
    as.numeric(x[[fallback_index]])
  }

  low <- pick(row, c("low", "LOW", "Low"), fallback_index = 1)
  est <- pick(row, c("est", "EST", "Est", "estimate"), fallback_index = min(2, length(row)))
  high <- pick(row, c("high", "HIGH", "High"), fallback_index = length(row))

  tibble(
    area_low = low,
    area_estimate = est,
    area_high = high
  )
}

akde_area_table <- purrr::map_dfr(akde_results, function(res) {
  purrr::map_dfr(ud_levels, function(level_ud) {
    area_ci <- extract_area_ci(res$ud, ci_level = ci_level, ud_level = level_ud)
    tibble(
      individual_id = res$id,
      ud_level = level_ud,
      area_low = area_ci$area_low,
      area_estimate = area_ci$area_estimate,
      area_high = area_ci$area_high
    ) %>%
      mutate(across(c(area_low, area_estimate, area_high), as.numeric))
  })
}) %>%
  left_join(id_lookup %>% rename(individual_id = bird_id), by = "individual_id") %>%
  mutate(bird_id = individual_id, .after = individual_id) %>%
  relocate(year_season, bird_id, .after = individual_id)

readr::write_csv(akde_area_table, file.path(table_dir, "akde_area_summary.csv"))

for (res in akde_results) {
  id_safe <- make.names(res$id)

  # Save fitted movement model and UD object.
  saveRDS(res$fit, file.path(model_dir, paste0("ctmm_fit_", id_safe, ".rds")))
  saveRDS(res$ud, file.path(model_dir, paste0("akde_ud_", id_safe, ".rds")))

  # Save rasterized UD (probability mass function).
  ud_pmf <- ctmm::raster(res$ud, DF = "PMF")
  ctmm::writeRaster(ud_pmf, filename = file.path(akde_raster_dir, paste0("akde_pmf_", id_safe, ".tif")), overwrite = TRUE)

  # Save UD contours (50% and 95%) as vector files.
  contour_file <- file.path(akde_vector_dir, paste0("akde_contours_", id_safe, ".gpkg"))
  if (file.exists(contour_file)) unlink(contour_file, recursive = TRUE, force = TRUE)
  ctmm::writeVector(
    res$ud,
    filename = contour_file,
    level.UD = ud_levels,
    level = ci_level
  )
}
```

## 3.5 Mean AKDE by year and overlap
```{r mean-akde-year}
extract_overlap_ci <- function(overlap_obj, id1, id2) {
  # ctmm::overlap usually returns a list with $CI[,,c("low","est","high")].
  if (is.list(overlap_obj) && !is.null(overlap_obj$CI)) {
    ci_array <- overlap_obj$CI
    get_ci <- function(level_name) {
      dims <- dim(ci_array)
      if (is.null(dims) || length(dims) < 3) return(NA_real_)
      dim3 <- dimnames(ci_array)[[3]]
      if (is.null(dim3) || !level_name %in% dim3) return(NA_real_)
      as.numeric(ci_array[id1, id2, level_name])
    }
    return(tibble(
      overlap_low = get_ci("low"),
      overlap_estimate = get_ci("est"),
      overlap_high = get_ci("high")
    ))
  }

  # Fallback for matrix-only outputs (estimate only, no CI).
  if (is.matrix(overlap_obj) && all(c(id1, id2) %in% rownames(overlap_obj))) {
    return(tibble(
      overlap_low = NA_real_,
      overlap_estimate = as.numeric(overlap_obj[id1, id2]),
      overlap_high = NA_real_
    ))
  }

  tibble(overlap_low = NA_real_, overlap_estimate = NA_real_, overlap_high = NA_real_)
}

akde_results_year <- purrr::map(akde_results, function(x) {
  x$year_season <- unname(id_year_lookup[[x$id]])
  x
})

akde_results_year <- akde_results_year[!is.na(vapply(akde_results_year, function(x) x$year_season, character(1)))]
akde_by_year <- split(akde_results_year, vapply(akde_results_year, function(x) x$year_season, character(1)))

year_ud <- purrr::imap(akde_by_year, function(res_year, yy) {
  uds <- purrr::map(res_year, "ud")

  # Try direct averaging first.
  direct_mean <- tryCatch(mean(uds, sample = FALSE), error = function(e) e)
  if (!inherits(direct_mean, "error")) {
    return(direct_mean)
  }

  message("Grid mismatch while averaging UDs for ", yy, "; recomputing this year's UDs on a shared grid.")

  ids <- purrr::map_chr(res_year, "id")
  data_year <- telemetry_list[ids]
  fit_year <- purrr::map(res_year, "fit")

  shared_grid <- if (is.null(akde_grid_dr_m)) {
    NULL
  } else {
    list(dr = akde_grid_dr_m, align.to.origin = TRUE)
  }

  uds_shared <- ctmm::akde(
    data_year,
    fit_year,
    weights = akde_use_weights,
    res = akde_res,
    error = akde_error,
    grid = shared_grid
  )

  mean(uds_shared, sample = FALSE)
})

for (yy in names(year_ud)) {
  ud_r <- ctmm::raster(year_ud[[yy]], DF = "PMF")
  ctmm::writeRaster(ud_r, filename = file.path(akde_raster_dir, paste0("akde_mean_pmf_", yy, ".tif")), overwrite = TRUE)

  # Also save publication-ready contour polygons for year-level maps.
  year_contour_file <- file.path(akde_vector_dir, paste0("akde_mean_contours_", yy, ".gpkg"))
  if (file.exists(year_contour_file)) unlink(year_contour_file, recursive = TRUE, force = TRUE)
  ctmm::writeVector(
    year_ud[[yy]],
    filename = year_contour_file,
    level.UD = year_map_contours,
    level = ci_level
  )
}

if (length(year_ud) >= 2) {
  year_pairs <- combn(names(year_ud), 2, simplify = FALSE)
  overlap_table <- purrr::map_dfr(year_pairs, function(pair) {
    y1 <- pair[1]
    y2 <- pair[2]
    ud_list <- list()
    ud_list[[y1]] <- year_ud[[y1]]
    ud_list[[y2]] <- year_ud[[y2]]

    ov_bh <- ctmm::overlap(ud_list, method = "Bhattacharyya", level = ci_level, debias = TRUE)
    ov_en <- ctmm::overlap(ud_list, method = "Encounter", level = ci_level, debias = TRUE)

    bind_rows(
      extract_overlap_ci(ov_bh, y1, y2) %>% mutate(method = "Bhattacharyya"),
      extract_overlap_ci(ov_en, y1, y2) %>% mutate(method = "Encounter")
    ) %>%
      mutate(year_1 = y1, year_2 = y2, .before = 1)
  })

  readr::write_csv(overlap_table, file.path(table_dir, "akde_overlap_by_year.csv"))
}
```

## 3.6 Statistical support for habitat-use comparison
```{r akde-tests}
# Individual-level 95% area comparison (non-parametric, robust for skewed areas).
akde_area_95 <- akde_area_table %>%
  dplyr::filter(abs(ud_level - 0.95) < 1e-8) %>%
  dplyr::select(individual_id, bird_id, year_season, area_estimate)

cliffs_delta <- function(x, y) {
  # Robust non-parametric effect size: P(x>y) - P(x<y).
  comp <- outer(x, y, FUN = "-")
  (sum(comp > 0) - sum(comp < 0)) / (length(x) * length(y))
}

year_pairs_area <- combn(sort(unique(akde_area_95$year_season)), 2, simplify = FALSE)
if (length(year_pairs_area) > 0) {
  area_test_table <- purrr::map_dfr(year_pairs_area, function(pair) {
    y1 <- pair[1]
    y2 <- pair[2]
    x <- akde_area_95 %>% dplyr::filter(year_season == y1) %>% dplyr::pull(area_estimate)
    y <- akde_area_95 %>% dplyr::filter(year_season == y2) %>% dplyr::pull(area_estimate)
    wt <- wilcox.test(x, y, exact = FALSE)

    tibble(
      year_1 = y1,
      year_2 = y2,
      n_1 = length(x),
      n_2 = length(y),
      median_1 = median(x, na.rm = TRUE),
      median_2 = median(y, na.rm = TRUE),
      statistic = as.numeric(wt$statistic),
      p_value = wt$p.value,
      cliffs_delta = cliffs_delta(x, y)
    )
  }) %>%
    mutate(p_value_fdr = p.adjust(p_value, method = "fdr"))

  readr::write_csv(area_test_table, file.path(table_dir, "akde_area95_year_wilcox.csv"))
}

# Bootstrap overlap with equal birds per year (handles unbalanced sample sizes).
set.seed(123)
ids_by_year <- id_lookup %>%
  dplyr::filter(bird_id %in% purrr::map_chr(akde_results, "id")) %>%
  group_by(year_season) %>%
  summarise(ids = list(bird_id), .groups = "drop") %>%
  tibble::deframe()

target_years <- if (all(c("2025", "2026") %in% names(ids_by_year))) c("2025", "2026") else sort(names(ids_by_year))[1:2]

if (length(target_years) == 2 && all(!is.na(target_years))) {
  y1 <- target_years[1]
  y2 <- target_years[2]
  n_equal <- min(length(ids_by_year[[y1]]), length(ids_by_year[[y2]]))
  ud_lookup_ind <- setNames(purrr::map(akde_results, "ud"), purrr::map_chr(akde_results, "id"))

  overlap_estimate_one <- function(ud_a, ud_b, y_a, y_b) {
    ud_list <- list()
    ud_list[[y_a]] <- ud_a
    ud_list[[y_b]] <- ud_b
    ov <- ctmm::overlap(ud_list, method = "Bhattacharyya", level = ci_level, debias = TRUE)
    extract_overlap_ci(ov, y_a, y_b)$overlap_estimate
  }

  # Observed overlap from full year means.
  observed_overlap <- NA_real_
  if (exists("overlap_table")) {
    observed_overlap <- overlap_table %>%
      dplyr::filter(year_1 == y1, year_2 == y2, method == "Bhattacharyya") %>%
      dplyr::pull(overlap_estimate) %>%
      .[1]
  }
  if (is.na(observed_overlap)) {
    ud_year <- list()
    ud_year[[y1]] <- mean(ud_lookup_ind[ids_by_year[[y1]]], sample = FALSE)
    ud_year[[y2]] <- mean(ud_lookup_ind[ids_by_year[[y2]]], sample = FALSE)
    ov_obs <- ctmm::overlap(ud_year, method = "Bhattacharyya", level = ci_level, debias = TRUE)
    observed_overlap <- extract_overlap_ci(ov_obs, y1, y2)$overlap_estimate
  }

  overlap_boot <- replicate(n_overlap_boot, {
    s1 <- sample(ids_by_year[[y1]], size = n_equal, replace = TRUE)
    s2 <- sample(ids_by_year[[y2]], size = n_equal, replace = TRUE)
    ud1 <- mean(ud_lookup_ind[s1], sample = FALSE)
    ud2 <- mean(ud_lookup_ind[s2], sample = FALSE)
    overlap_estimate_one(ud1, ud2, y1, y2)
  })

  # Permutation test (Niella-style random reassignment logic adapted to AKDE overlap):
  # if observed overlap is lower than null, groups are more spatially segregated than random expectation.
  n1 <- length(ids_by_year[[y1]])
  n2 <- length(ids_by_year[[y2]])
  pooled_ids <- c(ids_by_year[[y1]], ids_by_year[[y2]])
  overlap_perm <- replicate(n_overlap_perm, {
    shuffled <- sample(pooled_ids, size = length(pooled_ids), replace = FALSE)
    g1 <- shuffled[seq_len(n1)]
    g2 <- shuffled[(n1 + 1):(n1 + n2)]
    ud1 <- mean(ud_lookup_ind[g1], sample = FALSE)
    ud2 <- mean(ud_lookup_ind[g2], sample = FALSE)
    overlap_estimate_one(ud1, ud2, y1, y2)
  })

  overlap_boot_table <- tibble(overlap = overlap_boot)
  overlap_boot_summary <- tibble(
    year_1 = y1,
    year_2 = y2,
    n_boot = n_overlap_boot,
    n_equal_birds = n_equal,
    observed_overlap = observed_overlap,
    overlap_median = median(overlap_boot, na.rm = TRUE),
    overlap_lcl = quantile(overlap_boot, 0.025, na.rm = TRUE),
    overlap_ucl = quantile(overlap_boot, 0.975, na.rm = TRUE)
  )

  overlap_perm_table <- tibble(overlap = overlap_perm)
  overlap_perm_summary <- tibble(
    year_1 = y1,
    year_2 = y2,
    n_perm = n_overlap_perm,
    n_1 = n1,
    n_2 = n2,
    observed_overlap = observed_overlap,
    null_median = median(overlap_perm, na.rm = TRUE),
    p_value_lower_tail = mean(overlap_perm <= observed_overlap, na.rm = TRUE),
    p_value_upper_tail = mean(overlap_perm >= observed_overlap, na.rm = TRUE),
    p_value_two_sided = 2 * pmin(
      mean(overlap_perm <= observed_overlap, na.rm = TRUE),
      mean(overlap_perm >= observed_overlap, na.rm = TRUE)
    )
  )

  readr::write_csv(overlap_boot_table, file.path(table_dir, "akde_overlap_bootstrap_samples.csv"))
  readr::write_csv(overlap_boot_summary, file.path(table_dir, "akde_overlap_bootstrap_summary.csv"))
  readr::write_csv(overlap_perm_table, file.path(table_dir, "akde_overlap_permutation_samples.csv"))
  readr::write_csv(overlap_perm_summary, file.path(table_dir, "akde_overlap_permutation_summary.csv"))
}
```

```{r mean-akde-plots}
format_ud_level <- function(x) {
  x_num <- suppressWarnings(as.numeric(as.character(x)))
  if (all(is.na(x_num))) return(as.character(x))
  if (max(x_num, na.rm = TRUE) <= 1) x_num <- x_num * 100
  paste0(round(x_num), "%")
}

plot_mean_ud <- function(ud_obj, year_label) {
  r <- ctmm::raster(ud_obj, DF = "PMF")
  df <- as.data.frame(raster::rasterToPoints(r))
  names(df) <- c("x", "y", "pmf")

  ud_crs <- sf::st_crs(raster::projection(r))

  land_sf_plot <- NULL
  if (!is.null(land_sf_raw)) {
    land_sf_plot <- sf::st_transform(land_sf_raw, ud_crs)
  }

  colony_sf_plot <- sf::st_transform(colony_sf_ll, ud_crs)

  contour_path <- file.path(akde_vector_dir, paste0("akde_mean_contours_", year_label, ".gpkg"))
  contour_sf <- NULL
  if (file.exists(contour_path)) {
    contour_sf <- sf::st_read(contour_path, quiet = TRUE)
    level_col <- names(contour_sf)[grepl("level", names(contour_sf), ignore.case = TRUE)][1]
    if (!is.na(level_col)) {
      contour_sf <- contour_sf %>%
        mutate(ud_level_label = format_ud_level(.data[[level_col]]))
    }
  }

  p <- ggplot(df, aes(x = x, y = y, fill = pmf)) +
    geom_raster(interpolate = FALSE) +
    geom_contour(
      data = df,
      aes(x = x, y = y, z = pmf),
      inherit.aes = FALSE,
      bins = 10,
      colour = "white",
      alpha = 0.35,
      linewidth = 0.2
    ) +
    scale_fill_viridis_c(option = map_palette, trans = "sqrt") +
    coord_sf(datum = NA) +
    labs(
      title = paste("Mean autocorrelation-aware UD (AKDE) -", year_label),
      x = "Projected X",
      y = "Projected Y",
      fill = "Use intensity"
    ) +
    theme_minimal(base_size = 11) +
    theme(
      panel.grid.major = element_line(colour = "grey85", linewidth = 0.2),
      panel.grid.minor = element_blank(),
      panel.background = element_rect(fill = "#dff0fa", colour = NA)
    )

  if (!is.null(land_sf_plot)) {
    p <- p +
      geom_sf(data = land_sf_plot, inherit.aes = FALSE, fill = "grey88", colour = "grey35", linewidth = 0.25)
  }

  if (!is.null(contour_sf)) {
    if ("ud_level_label" %in% names(contour_sf)) {
      p <- p +
        geom_sf(
          data = contour_sf,
          aes(colour = ud_level_label),
          inherit.aes = FALSE,
          fill = NA,
          linewidth = 0.75,
          alpha = 0.85
        ) +
        scale_colour_manual(
          values = c("50%" = "#cc4c02", "75%" = "#fe9929", "95%" = "#fec44f"),
          name = "UD contour"
        )
    } else {
      p <- p +
        geom_sf(
          data = contour_sf,
          inherit.aes = FALSE,
          fill = NA,
          colour = "black",
          linewidth = 0.5,
          alpha = 0.85
        )
    }
  }

  p <- p +
    geom_sf(
      data = colony_sf_plot,
      inherit.aes = FALSE,
      shape = 21,
      size = 2.5,
      fill = "gold",
      colour = "black",
      stroke = 0.4
    )

  p
}

for (yy in names(year_ud)) {
  p <- plot_mean_ud(year_ud[[yy]], yy)
  ggsave(file.path(fig_dir, paste0("fig_akde_mean_ud_", yy, ".png")), p, width = 8, height = 6, dpi = 300)
}

# Faceted year map with identical colour scale for direct visual comparison.
mean_ud_df <- purrr::imap_dfr(year_ud, function(ud_obj, yy) {
  r <- ctmm::raster(ud_obj, DF = "PMF")
  out <- as.data.frame(raster::rasterToPoints(r))
  names(out) <- c("x", "y", "pmf")
  out$year_season <- yy
  out
})

if (nrow(mean_ud_df) > 0) {
  facet_coord <- if (!is.null(land_sf_raw)) coord_sf(datum = NA) else coord_equal()
  fig_mean_ud_facet <- ggplot(mean_ud_df, aes(x = x, y = y, fill = pmf)) +
    geom_raster(interpolate = FALSE) +
    scale_fill_viridis_c(option = map_palette, trans = "sqrt") +
    facet_wrap(~year_season, ncol = 2) +
    facet_coord +
    labs(
      title = "Year-level AKDE habitat use comparison",
      subtitle = "Shared scale across panels; PMF from autocorrelation-aware AKDE",
      x = "Projected X",
      y = "Projected Y",
      fill = "Use intensity"
    ) +
    theme_minimal(base_size = 11) +
    theme(
      panel.grid.major = element_line(colour = "grey85", linewidth = 0.2),
      panel.grid.minor = element_blank(),
      panel.background = element_rect(fill = "#dff0fa", colour = NA)
    )

  if (!is.null(land_sf_raw)) {
    ud_crs_facet <- sf::st_crs(raster::projection(ctmm::raster(year_ud[[1]], DF = "PMF")))
    land_sf_facet <- sf::st_transform(land_sf_raw, ud_crs_facet)
    fig_mean_ud_facet <- fig_mean_ud_facet +
      geom_sf(data = land_sf_facet, inherit.aes = FALSE, fill = "grey88", colour = "grey35", linewidth = 0.25)
  }

  ggsave(
    file.path(fig_dir, "fig_akde_mean_ud_by_year_facet.png"),
    fig_mean_ud_facet,
    width = 10,
    height = 6,
    dpi = 300
  )
}

# Contour-only comparison map (common publication style in seabird foraging papers).
year_contour_list <- purrr::imap(year_ud, function(.ud, yy) {
  contour_path <- file.path(akde_vector_dir, paste0("akde_mean_contours_", yy, ".gpkg"))
  if (!file.exists(contour_path)) return(NULL)
  sf_obj <- sf::st_read(contour_path, quiet = TRUE)
  level_col <- names(sf_obj)[grepl("level", names(sf_obj), ignore.case = TRUE)][1]
  if (!is.na(level_col)) {
    sf_obj <- sf_obj %>% mutate(ud_level_label = format_ud_level(.data[[level_col]]))
  }
  sf_obj$year_season <- yy
  sf_obj
})

year_contour_list <- year_contour_list[!vapply(year_contour_list, is.null, logical(1))]
year_contours <- if (length(year_contour_list) > 0) do.call(rbind, year_contour_list) else NULL

if (!is.null(year_contours) && nrow(year_contours) > 0) {
  ud_crs_cont <- sf::st_crs(year_contours)
  colony_sf_cont <- sf::st_transform(colony_sf_ll, ud_crs_cont)
  fig_contours <- ggplot() +
    theme_minimal(base_size = 11) +
    theme(
      panel.grid.major = element_line(colour = "grey85", linewidth = 0.2),
      panel.grid.minor = element_blank(),
      panel.background = element_rect(fill = "#dff0fa", colour = NA)
    ) +
    labs(
      title = "AKDE contour comparison by year",
      subtitle = "Contour rings represent 50%, 75% and 95% utilization distributions",
      x = "Projected X",
      y = "Projected Y",
      colour = "Year"
    ) +
    coord_sf(datum = NA)

  if (!is.null(land_sf_raw)) {
    land_sf_cont <- sf::st_transform(land_sf_raw, ud_crs_cont)
    fig_contours <- fig_contours +
      geom_sf(data = land_sf_cont, fill = "grey88", colour = "grey35", linewidth = 0.25)
  }

  if ("ud_level_label" %in% names(year_contours)) {
    fig_contours <- fig_contours +
      geom_sf(
        data = year_contours,
        aes(colour = year_season, linetype = ud_level_label),
        fill = NA,
        linewidth = 0.8,
        alpha = 0.9
      ) +
      scale_linetype_discrete(name = "UD contour")
  } else {
    fig_contours <- fig_contours +
      geom_sf(
        data = year_contours,
        aes(colour = year_season),
        fill = NA,
        linewidth = 0.8,
        alpha = 0.9
      )
  }

  fig_contours <- fig_contours +
    geom_sf(
      data = colony_sf_cont,
      shape = 21,
      size = 2.8,
      fill = "gold",
      colour = "black",
      stroke = 0.4
    )

  ggsave(
    file.path(fig_dir, "fig_akde_contour_comparison_by_year.png"),
    fig_contours,
    width = 9,
    height = 6,
    dpi = 300
  )
}
```

```{r overlap-bootstrap-plot}
if (exists("overlap_boot_table") && nrow(overlap_boot_table) > 0) {
  fig_overlap_boot <- ggplot(overlap_boot_table, aes(x = overlap)) +
    geom_histogram(bins = 30, fill = "#2c7fb8", alpha = 0.8, colour = "white") +
    geom_vline(
      xintercept = overlap_boot_summary$overlap_median,
      colour = "firebrick",
      linewidth = 0.8
    ) +
    geom_vline(
      xintercept = overlap_boot_summary$observed_overlap,
      colour = "black",
      linewidth = 0.8,
      linetype = "dashed"
    ) +
    labs(
      title = "Bootstrap distribution of year-overlap in habitat use",
      subtitle = "Equal-bird bootstrap across years using AKDE mean-UD overlap",
      x = paste0("Overlap (", overlap_boot_summary$year_1, " vs ", overlap_boot_summary$year_2, ")"),
      y = "Bootstrap frequency"
    ) +
    theme_minimal()

  fig_overlap_boot

  ggsave(
    file.path(fig_dir, "fig_akde_overlap_bootstrap_histogram.png"),
    fig_overlap_boot,
    width = 8,
    height = 5,
    dpi = 300
  )
}

if (exists("overlap_perm_table") && nrow(overlap_perm_table) > 0) {
  fig_overlap_perm <- ggplot(overlap_perm_table, aes(x = overlap)) +
    geom_histogram(bins = 30, fill = "#969696", alpha = 0.9, colour = "white") +
    geom_vline(
      xintercept = overlap_perm_summary$observed_overlap,
      colour = "#cb181d",
      linewidth = 0.9
    ) +
    labs(
      title = "Permutation null distribution of overlap",
      subtitle = "Observed overlap (red) vs random year-label reassignment null",
      x = "Bhattacharyya overlap",
      y = "Permutation frequency"
    ) +
    theme_minimal()

  ggsave(
    file.path(fig_dir, "fig_akde_overlap_permutation_histogram.png"),
    fig_overlap_perm,
    width = 8,
    height = 5,
    dpi = 300
  )
}
```

# 4. Environmental covariates with autocorrelation-aware iRSF
Put environmental raster covariates (`.tif`) into:

`r env_input_dir`

Typical candidates:
- bathymetry/depth
- slope
- SST
- chlorophyll-a
- distance to colony / distance to shelf break

```{r load-env}
env_files <- list.files(env_input_dir, pattern = "\\.tif$", full.names = TRUE)

if (length(env_files) == 0) {
  warning(
    paste0(
      "No environmental .tif files found in ", env_input_dir,
      ". Add raster covariates, then rerun section 4."
    )
  )
} else {
  # Use projection from the first telemetry object as target.
  target_proj <- ctmm::projection(akde_results[[1]]$ud)

  env_terra <- setNames(lapply(env_files, terra::rast), tools::file_path_sans_ext(basename(env_files)))

  env_projected <- purrr::imap(env_terra, function(r, nm) {
    r_proj <- terra::project(r, target_proj, method = "bilinear")
    terra::writeRaster(r_proj, file.path(env_projected_dir, paste0(nm, "_projected.tif")), overwrite = TRUE)
    raster::raster(r_proj)
  })

  env_raster_list <- env_projected
  env_names <- names(env_raster_list)

  rsf_formula <- as.formula(paste("~", paste(env_names, collapse = " + ")))
  rsf_formula
}
```

```{r fit-rsf, eval=TRUE}
if (exists("env_raster_list")) {
  fit_rsf_one <- function(res) {
    rsf_fit <- ctmm::rsf.fit(
      data = telemetry_list[[res$id]],
      UD = res$ud,
      R = env_raster_list,
      formula = rsf_formula,
      integrated = TRUE,
      trace = TRUE
    )

    id_safe <- make.names(res$id)
    saveRDS(rsf_fit, file.path(model_dir, paste0("irsf_", id_safe, ".rds")))

    co <- tryCatch(stats::coef(rsf_fit), error = function(e) NULL)
    if (is.null(co)) {
      return(tibble(bird_id = res$id, term = NA_character_, estimate = NA_real_))
    }

    tibble(
      bird_id = res$id,
      year_season = unname(id_year_lookup[[res$id]]),
      term = names(co),
      estimate = as.numeric(co)
    )
  }

  rsf_coef_table <- purrr::map_dfr(akde_results, ~purrr::possibly(fit_rsf_one, otherwise = NULL)(.x)) %>%
    dplyr::filter(!is.na(term))

  readr::write_csv(rsf_coef_table, file.path(table_dir, "irsf_coefficients_by_individual.csv"))

  rsf_coef_summary <- rsf_coef_table %>%
    group_by(year_season, term) %>%
    summarise(
      n_individuals = n(),
      median_estimate = median(estimate, na.rm = TRUE),
      q1_estimate = quantile(estimate, 0.25, na.rm = TRUE),
      q3_estimate = quantile(estimate, 0.75, na.rm = TRUE),
      .groups = "drop"
    )

  readr::write_csv(rsf_coef_summary, file.path(table_dir, "irsf_coefficients_summary_by_year.csv"))

  # One-sample tests: does each coefficient differ from 0 within each year?
  rsf_within_year_tests <- rsf_coef_table %>%
    group_by(year_season, term) %>%
    summarise(
      n_individuals = n(),
      median_estimate = median(estimate, na.rm = TRUE),
      p_value_wilcox_vs0 = tryCatch(wilcox.test(estimate, mu = 0, exact = FALSE)$p.value, error = function(e) NA_real_),
      .groups = "drop"
    ) %>%
    mutate(p_value_fdr = p.adjust(p_value_wilcox_vs0, method = "fdr"))

  readr::write_csv(rsf_within_year_tests, file.path(table_dir, "irsf_within_year_wilcox_vs0.csv"))

  # Between-year tests per coefficient (if both years available).
  if (length(unique(rsf_coef_table$year_season)) >= 2) {
    rsf_year_pair <- if (all(c("2025", "2026") %in% unique(rsf_coef_table$year_season))) {
      c("2025", "2026")
    } else {
      sort(unique(rsf_coef_table$year_season))[1:2]
    }
    y1 <- rsf_year_pair[1]
    y2 <- rsf_year_pair[2]

    rsf_between_year_tests <- rsf_coef_table %>%
      dplyr::filter(year_season %in% c(y1, y2)) %>%
      group_by(term) %>%
      group_modify(~{
        d <- .x
        tibble(
          n_1 = sum(d$year_season == y1),
          n_2 = sum(d$year_season == y2),
          median_1 = median(d$estimate[d$year_season == y1], na.rm = TRUE),
          median_2 = median(d$estimate[d$year_season == y2], na.rm = TRUE),
          p_value_wilcox_between_years = tryCatch(
            wilcox.test(estimate ~ year_season, data = d, exact = FALSE)$p.value,
            error = function(e) NA_real_
          )
        )
      }) %>%
      ungroup() %>%
      mutate(
        year_1 = y1,
        year_2 = y2,
        p_value_fdr = p.adjust(p_value_wilcox_between_years, method = "fdr"),
        .before = 1
      )

    readr::write_csv(rsf_between_year_tests, file.path(table_dir, "irsf_between_year_wilcox.csv"))
  }
}
```

## 4.1 Plot iRSF coefficient summaries
```{r rsf-coef-plot}
if (exists("rsf_coef_summary") && nrow(rsf_coef_summary) > 0) {
  fig_rsf_coef <- ggplot(
    rsf_coef_summary,
    aes(x = median_estimate, y = term, xmin = q1_estimate, xmax = q3_estimate, colour = year_season)
  ) +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_pointrange(position = position_dodge(width = 0.5)) +
    labs(
      title = "Autocorrelation-aware iRSF coefficients",
      subtitle = "Points are medians across individuals; bars are interquartile ranges",
      x = "Coefficient estimate",
      y = "Environmental covariate",
      colour = "Year"
    ) +
    theme_minimal()

  fig_rsf_coef

  ggsave(file.path(fig_dir, "fig_irsf_coefficients_by_year.png"), fig_rsf_coef, width = 9, height = 6, dpi = 300)
}
```

# 5. Notes for interpretation
- If sampling interval differs between years, report it explicitly.
- AKDE and iRSF here account for autocorrelation more rigorously than IID KDE + naive GLM.
- If you need strict comparability with mIBA/track2KBA outputs for policy workflows, run a parallel track2KBA KDE analysis as a sensitivity check.
- Runtime in section 3.3 is driven mostly by model selection (`ctmm.select`), weighted AKDE, and raster resolution; for exploratory runs use the default fast settings in section 1.3, then switch to slower robust settings for final analyses.

# 6. Reporting scaffold (Methods + Results text helper)
```{r reporting-scaffold}
fmt_num <- function(x, digits = 3) {
  ifelse(is.na(x), "NA", formatC(x, digits = digits, format = "f"))
}

methods_lines <- c(
  "Habitat use was quantified from at-sea GPS locations using autocorrelation-aware kernel density estimation (AKDE) in the ctmm framework.",
  "Individual movement models were fit from telemetry variograms and used to derive individual AKDE utilization distributions (UDs).",
  "Year-level mean UDs were obtained from individual UDs, and overlap between years was estimated using Bhattacharyya and encounter overlap metrics with uncertainty propagated by ctmm.",
  "To account for unbalanced sample sizes between years, overlap uncertainty and segregation strength were assessed with equal-bird bootstrap and year-label permutation procedures.",
  "Year differences in individual 95% AKDE area were tested with Wilcoxon rank-sum tests and effect sizes reported as Cliff's delta."
)

results_lines <- c("Run the analysis first to populate result tables.")

ov_path <- file.path(table_dir, "akde_overlap_by_year.csv")
ov_perm_path <- file.path(table_dir, "akde_overlap_permutation_summary.csv")
area_path <- file.path(table_dir, "akde_area95_year_wilcox.csv")

if (file.exists(ov_path) && file.exists(ov_perm_path) && file.exists(area_path)) {
  ov <- readr::read_csv(ov_path, show_col_types = FALSE)
  ov_perm <- readr::read_csv(ov_perm_path, show_col_types = FALSE)
  area <- readr::read_csv(area_path, show_col_types = FALSE)

  ov_bh <- ov %>% dplyr::filter(method == "Bhattacharyya") %>% dplyr::slice(1)
  ov_en <- ov %>% dplyr::filter(method == "Encounter") %>% dplyr::slice(1)
  area_1 <- area %>% dplyr::slice(1)
  perm_1 <- ov_perm %>% dplyr::slice(1)

  results_lines <- c(
    paste0(
      "Year-level Bhattacharyya overlap was ",
      fmt_num(ov_bh$overlap_estimate),
      " (",
      fmt_num(ov_bh$overlap_low),
      "-",
      fmt_num(ov_bh$overlap_high),
      "), and encounter overlap was ",
      fmt_num(ov_en$overlap_estimate),
      " (",
      fmt_num(ov_en$overlap_low),
      "-",
      fmt_num(ov_en$overlap_high),
      ")."
    ),
    paste0(
      "Permutation testing gave p(lower-tail) = ",
      fmt_num(perm_1$p_value_lower_tail, digits = 4),
      " for overlap lower than null expectation."
    ),
    paste0(
      "Individual 95% AKDE area comparison (",
      area_1$year_1, " vs ", area_1$year_2,
      ") gave W = ",
      fmt_num(area_1$statistic, digits = 0),
      ", p = ",
      fmt_num(area_1$p_value, digits = 4),
      ", Cliff's delta = ",
      fmt_num(area_1$cliffs_delta, digits = 3),
      "."
    )
  )
}

report_lines <- c(
  "METHODS (draft):",
  methods_lines,
  "",
  "RESULTS (draft):",
  results_lines
)

writeLines(report_lines, file.path(table_dir, "manuscript_methods_results_draft.txt"))
cat(paste(report_lines, collapse = "\n"), "\n")
```

# End
```{r end}
cat("05_habitat_use_kde_environment script completed.\n")
cat("Outputs directory:", output_dir, "\n")
```
