---
title: "06_hmm_behavioural_states"
output:
  html_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
# Show code in the knitted report, but suppress warnings/messages for readability.
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

# Purpose

This script fits a **2-state Hidden Markov Model (HMM)** to the retained rockhopper foraging trips.

This version has been rebuilt to follow the standard logic taught in movement-analysis workflows:

1. inspect the sampling structure  
2. evaluate candidate regular time resolutions  
3. split tracks at long gaps  
4. thin / regularise the data to a chosen interval  
5. derive step length and turning angle from the regularised data  
6. fit the HMM and decode behavioural states  

This is important because discrete-time HMMs are much easier to interpret when location data are approximately regular in time.

This script is designed to give you **control** over key preprocessing choices:
- candidate thinning resolutions
- final chosen resolution
- maximum allowed gap before splitting
- minimum retained subtrack duration

# Packages

```{r packages}
library(dplyr)        # data manipulation
library(readr)        # csv writing
library(lubridate)    # date-time handling and rounding
library(ggplot2)      # plotting
library(openxlsx)     # Excel outputs
library(momentuHMM)   # HMM fitting and movement preprocessing
library(tidyr)        # reshaping data
```

# Project folders

```{r folders}
# Use the current working directory as the project root.
project_dir <- getwd()

# Folder containing the analysis-ready objects from script 00.
input_dir <- file.path(project_dir, "outputs", "03_setup_and_load")

# Output folder for this script.
output_dir <- file.path(project_dir, "outputs", "06_hmm_behavioural_states")

# Create the output folder if needed.
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
```

# User controls

```{r user-controls}
# Candidate thinning resolutions (in minutes) to evaluate before selecting one.
candidate_resolutions_min <- c(5, 10, 15, 20, 30, 60)

# Final thinning resolution used for the HMM.
# Change this value after inspecting the resolution-evaluation outputs.
hmm_res_min <- 15

# Maximum allowed gap (in minutes) before a track is split into subtracks.
max_gap_min <- 60

# Minimum duration (in minutes) required for a subtrack to be retained after splitting.
min_subtrack_duration_min <- 60

# Number of HMM states.
nbStates <- 2
```

# Helper functions

```{r helpers}
# Simple plotting theme used throughout the script.
theme_paper <- function() {
  theme_minimal() +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(colour = "grey85", linewidth = 0.4),
      axis.text = element_text(colour = "grey20"),
      plot.title = element_text(face = "bold", hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5),
      legend.position = "right"
    )
}

# -------------------------------------------------------------------
# Split tracks at large time gaps
# -------------------------------------------------------------------
split_at_gap <- function(data, max_gap = 60, shortest_track = 0) {
  n_tracks <- length(unique(data$ID))                       # number of original tracks

  data$ID_old <- data$ID                                    # keep original track ID
  data$ID <- character(nrow(data))                          # initialise new split-track ID

  for(i_track in 1:n_tracks) {
    ind_this_track <- which(data$ID_old == unique(data$ID_old)[i_track])  # indices of this track
    track_length <- length(ind_this_track)                                 # number of rows in this track

    dtimes <- difftime(                                                    # consecutive time gaps in minutes
      data$time[ind_this_track[-1]],
      data$time[ind_this_track[-track_length]],
      units = "mins"
    )

    ind_gap <- c(0, which(dtimes > max_gap), track_length)                 # locations of large gaps

    subtrack_ID <- rep(1:(length(ind_gap) - 1), diff(ind_gap))             # assign subtrack number
    data$ID[ind_this_track] <- paste0(data$ID_old[ind_this_track], "-", subtrack_ID)
  }

  track_lengths <- sapply(unique(data$ID), function(id) {                  # calculate duration of each subtrack
    ind <- which(data$ID == id)
    difftime(data$time[ind[length(ind)]], data$time[ind[1]], units = "min")
  })

  ID_keep <- names(track_lengths)[which(track_lengths >= shortest_track)]  # keep only long-enough subtracks
  data <- subset(data, ID %in% ID_keep)

  return(data)
}

# -------------------------------------------------------------------
# Estimate the proportion of missing locations that would occur for a chosen resolution
# -------------------------------------------------------------------
p_na <- function(time, res) {
  time <- round_date(time, paste(res, "min"))                              # round timestamps to the chosen resolution
  time <- na.omit(time[time != dplyr::lead(time)])                         # remove duplicate rounded times
  max_n_loc <- length(seq(time[1], tail(time, 1) + res * 60,               # theoretical maximum number of locations
                          by = as.difftime(res, units = "mins")))
  n_NA <- max_n_loc - (length(time) + 1)                                   # missing observations implied by the chosen resolution
  n_NA / max_n_loc                                                         # proportion missing
}
```

# Load the analysis-ready inputs

```{r load-inputs}
tracks <- readRDS(file.path(input_dir, "tracks_clean.rds"))                # point-level tracking object
trips <- readRDS(file.path(input_dir, "trips_clean.rds"))                  # trip-level object
qa <- readRDS(file.path(input_dir, "qa_clean.rds"))                        # QA object
```

# Keep only points belonging to retained final trips

```{r subset-final-trip-points}
tracks_trip_points <- tracks %>%
  filter(!is.na(global_trip_id)) %>%                                       # keep only retained trip points
  arrange(bird_id, global_trip_id, datetime) %>%
  rename(ID = global_trip_id)

cat("Points in final retained trips:", nrow(tracks_trip_points), "\n")
cat("Trips represented in point data:", dplyr::n_distinct(tracks_trip_points$ID), "\n")
```

# Build the base HMM table

```{r base-hmm-table}
# Build a clean table with the variables needed for resolution testing and HMM input.
hmm_base <- tracks_trip_points %>%
  transmute(
    ID = ID,                                                   # each retained trip is an initial HMM track
    bird_id = bird_id,
    deploymentid = deploymentid,
    tagnumber = tagnumber,
    year_season = year_season,
    time = time,                                                       # helper expects a column called "time"
    x = long,
    y = lat
  ) %>%
  arrange(ID, time)
```

# Evaluate candidate thinning resolutions

```{r evaluate-resolutions}
# Estimate missingness for each candidate resolution within each trip.
resolution_eval <- tidyr::expand_grid(
  ID = unique(hmm_base$ID),
  res_min = candidate_resolutions_min
) %>%
  left_join(
    hmm_base %>%
      group_by(ID) %>%
      summarise(
        n_points = n(),
        start_time = min(time),
        end_time = max(time),
        .groups = "drop"
      ),
    by = "ID"
  ) %>%
  rowwise() %>%
  mutate(
    prop_missing = p_na(
      time = hmm_base$time[hmm_base$ID == ID],
      res = res_min
    )
  ) %>%
  ungroup()

# Summarise candidate resolutions across trips.
resolution_summary <- resolution_eval %>%
  group_by(res_min) %>%
  summarise(
    n_tracks = n(),
    mean_prop_missing = mean(prop_missing, na.rm = TRUE),
    median_prop_missing = median(prop_missing, na.rm = TRUE),
    max_prop_missing = max(prop_missing, na.rm = TRUE),
    .groups = "drop"
  )

resolution_summary
```

# Plot resolution diagnostics

```{r resolution-figures}
fig_resolution <- ggplot(resolution_summary, aes(x = factor(res_min), y = mean_prop_missing)) +
  geom_col() +
  labs(
    title = "Mean proportion of missing locations by candidate resolution",
    x = "Candidate resolution (minutes)",
    y = "Mean proportion missing"
  ) +
  theme_paper()

fig_resolution_box <- ggplot(resolution_eval, aes(x = factor(res_min), y = prop_missing)) +
  geom_boxplot() +
  labs(
    title = "Distribution of implied missingness across tracks",
    x = "Candidate resolution (minutes)",
    y = "Proportion missing"
  ) +
  theme_paper()

fig_resolution
fig_resolution_box
```

# Split tracks at large gaps

```{r split-at-gaps}
# Split tracks at large time gaps before thinning.
# This avoids forcing one HMM state process across biologically disconnected segments.
hmm_split <- split_at_gap(
  data = hmm_base,
  max_gap = max_gap_min,
  shortest_track = min_subtrack_duration_min
) %>%
  arrange(ID, time)

cat("Original number of retained trips:", dplyr::n_distinct(hmm_base$ID), "\n")
cat("Number of split subtracks retained:", dplyr::n_distinct(hmm_split$ID), "\n")
```

# Thin / regularise the tracks to the selected interval

```{r thin-tracks}
# Round timestamps to the chosen regular interval.
hmm_thinned <- hmm_split %>%
  mutate(
    time_round = round_date(time, paste(hmm_res_min, "min"))
  ) %>%
  arrange(ID, time, .by_group = FALSE) %>%
  group_by(ID, time_round) %>%
  slice(1) %>%                                                               # keep the first observation in each rounded time bin
  ungroup() %>%
  mutate(
    time = time_round
  ) %>%
  select(-time_round) %>%
  arrange(ID, time)

# Print a quick check.
cat("Number of rows after thinning:", nrow(hmm_thinned), "\n")
cat("Number of subtracks after thinning:", dplyr::n_distinct(hmm_thinned$ID), "\n")
```

# Re-check missingness after splitting and thinning

```{r post-thinning-check}
post_thinning_check <- hmm_thinned %>%
  group_by(ID) %>%
  summarise(
    n_points = n(),
    prop_missing_at_selected_res = p_na(time, hmm_res_min),
    .groups = "drop"
  )

post_thinning_summary <- post_thinning_check %>%
  summarise(
    n_tracks = n(),
    mean_prop_missing = mean(prop_missing_at_selected_res, na.rm = TRUE),
    median_prop_missing = median(prop_missing_at_selected_res, na.rm = TRUE),
    max_prop_missing = max(prop_missing_at_selected_res, na.rm = TRUE)
  )

post_thinning_summary
```

# Prepare movement streams for the HMM

```{r prepdata}
# Convert to a plain data.frame for reliable compatibility with momentuHMM::prepData().
hmm_thinned_df <- as.data.frame(hmm_thinned)

# Calculate step length and turning angle from the regularised data.
# With type = "LL", step lengths are calculated in kilometres
# and turning angles are based on initial bearings.
hmm_data <- momentuHMM::prepData(
  data = hmm_thinned_df,
  type = "LL",
  coordNames = c("x", "y")
) %>%
  tibble::as_tibble() %>%
  group_by(ID) %>%
  arrange(time, .by_group = TRUE) %>%
  mutate(
    dt_h = as.numeric(difftime(time, lag(time), units = "hours")),          # realised time lag in hours
    speed_kmh = step / dt_h                                                 # speed in km/h
  ) %>%
  ungroup()
```

# Clean the HMM input data

```{r clean-hmm-input}
# Keep only rows where step and angle are available.
# The first observation of each track cannot have both metrics.
hmm_data_fit <- hmm_data %>%
  filter(!is.na(step), !is.na(angle)) %>%
  filter(is.finite(step), is.finite(angle)) %>%
  arrange(ID, time)

cat("Rows used for HMM fitting:", nrow(hmm_data_fit), "\n")
cat("Tracks used for HMM fitting:", dplyr::n_distinct(hmm_data_fit$ID), "\n")
```

# Inspect the fitted-data distributions

```{r input-distributions}
fig_step_hist <- ggplot(hmm_data_fit, aes(x = step)) +
  geom_histogram(bins = 60) +
  labs(
    title = "Distribution of step length used in the HMM",
    x = "Step length (km)",
    y = "Count"
  ) +
  theme_paper()

fig_angle_hist <- ggplot(hmm_data_fit, aes(x = angle * 180 / pi)) +
  geom_histogram(bins = 60) +
  labs(
    title = "Distribution of turning angle used in the HMM",
    x = "Turning angle (degrees)",
    y = "Count"
  ) +
  theme_paper()

fig_step_hist
fig_angle_hist
```

# Define the HMM structure

```{r hmm-setup}
# Fit a 2-state HMM using:
# - gamma distribution for step length
# - von Mises distribution for turning angle
dist <- list(
  step = "gamma",
  angle = "vm"
)
```

# Choose starting values

```{r start-values}
# Use simple data-informed starting values.
# One state starts with shorter steps; the other with longer steps.
step_quants <- quantile(hmm_data_fit$step, probs = c(0.25, 0.75), na.rm = TRUE)

mu0 <- c(
  as.numeric(step_quants[1]),
  as.numeric(step_quants[2])
)

sigma0 <- c(
  max(mu0[1] / 2, 0.01),
  max(mu0[2] / 2, 0.01)
)

# Lower kappa = broader turning-angle distribution.
# Higher kappa = more concentrated turning around zero.
kappa0 <- c(0.5, 2)

Par0 <- list(
  step = c(mu0, sigma0),
  angle = kappa0
)

Par0
```

# Fit the 2-state HMM

```{r fit-hmm}
# Fit the hidden Markov model on the regularised movement data.
hmm_fit <- momentuHMM::fitHMM(
  data = hmm_data_fit,
  nbStates = nbStates,
  dist = dist,
  Par0 = Par0
)
```

# Inspect the fitted model

```{r inspect-model}
print(hmm_fit)

# Extract a few simple model diagnostics.
hmm_loglik <- as.numeric(logLik(hmm_fit))
hmm_aic <- AIC(hmm_fit)

model_info_tbl <- tibble::tibble(
  nb_states = nbStates,
  chosen_resolution_min = hmm_res_min,
  max_gap_min = max_gap_min,
  min_subtrack_duration_min = min_subtrack_duration_min,
  logLik = hmm_loglik,
  AIC = hmm_aic
)

model_info_tbl
```

# Decode the most likely state sequence

```{r decode-states}
# Use Viterbi decoding to assign the most likely state to each step.
decoded_states <- momentuHMM::viterbi(hmm_fit)

hmm_data_decoded <- hmm_data_fit %>%
  mutate(decoded_state = decoded_states)
```

# Summarise the decoded states

```{r state-summaries}
# Summarise movement properties by decoded state.
state_summary_tbl <- hmm_data_decoded %>%
  group_by(decoded_state) %>%
  summarise(
    n_steps = n(),
    mean_step_km = mean(step, na.rm = TRUE),
    sd_step_km = sd(step, na.rm = TRUE),
    mean_speed_kmh = mean(speed_kmh, na.rm = TRUE),
    sd_speed_kmh = sd(speed_kmh, na.rm = TRUE),
    mean_turn_deg = mean(angle * 180 / pi, na.rm = TRUE),
    mean_abs_turn_deg = mean(abs(angle * 180 / pi), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(mean_step_km)

# Assign a simple ecological label by ordering states from shorter to longer steps.
state_summary_tbl <- state_summary_tbl %>%
  mutate(
    behavioural_label = c("Searching-like", "Transit-like")
  )

state_summary_tbl
```

# Summarise state budgets by year

```{r state-budgets}
state_budget_by_year <- hmm_data_decoded %>%
  group_by(year_season, decoded_state) %>%
  summarise(
    n_steps = n(),
    .groups = "drop_last"
  ) %>%
  mutate(
    prop_steps = n_steps / sum(n_steps)
  ) %>%
  ungroup() %>%
  left_join(
    state_summary_tbl %>% select(decoded_state, behavioural_label),
    by = "decoded_state"
  )

state_budget_by_year
```

# Figures: state-dependent distributions

```{r state-distribution-figures}
fig_step_by_state <- ggplot(hmm_data_decoded, aes(x = step, fill = factor(decoded_state))) +
  geom_histogram(bins = 60, alpha = 0.7, position = "identity") +
  labs(
    title = "Step length distribution by decoded state",
    x = "Step length (km)",
    y = "Count",
    fill = "State"
  ) +
  theme_paper()

fig_angle_by_state <- ggplot(hmm_data_decoded, aes(x = angle * 180 / pi, fill = factor(decoded_state))) +
  geom_histogram(bins = 60, alpha = 0.7, position = "identity") +
  labs(
    title = "Turning-angle distribution by decoded state",
    x = "Turning angle (degrees)",
    y = "Count",
    fill = "State"
  ) +
  theme_paper()

fig_step_by_state
fig_angle_by_state
```

# Figure: state budget by year

```{r state-budget-figure}
fig_state_budget <- ggplot(
  state_budget_by_year,
  aes(x = year_season, y = prop_steps, fill = behavioural_label)
) +
  geom_col(position = "stack") +
  labs(
    title = "Decoded behavioural state budget by year",
    subtitle = paste0(
      "Regularised to ", hmm_res_min, "-minute intervals; split at gaps > ",
      max_gap_min, " minutes"
    ),
    x = "Year",
    y = "Proportion of steps",
    fill = "Behavioural label"
  ) +
  theme_paper()

fig_state_budget
```

# Figure: example regularised trip with decoded states

```{r example-trip-figure}
# Select one example track with many steps for visualisation.
example_trip_id <- hmm_data_decoded %>%
  count(ID, sort = TRUE) %>%
  slice(1) %>%
  pull(ID)

# Build plotting data by joining decoded states back to the thinned coordinates.
example_trip_plot_tbl <- hmm_data_decoded %>%
  filter(ID == example_trip_id) %>%
  left_join(
    hmm_thinned %>%
      select(ID, time, x, y),
    by = c("ID", "time")
  )

fig_example_trip <- ggplot(example_trip_plot_tbl, aes(x = x, y = y, colour = factor(decoded_state))) +
  geom_path(alpha = 0.6) +
  geom_point(size = 1.2) +
  labs(
    title = paste("Example regularised subtrack coloured by decoded state:", example_trip_id),
    x = "Longitude",
    y = "Latitude",
    colour = "State"
  ) +
  theme_paper()

fig_example_trip
```

# Save figures

```{r save-figures}
ggsave(file.path(output_dir, "fig_resolution_summary.png"), fig_resolution, width = 7, height = 5, dpi = 300)
ggsave(file.path(output_dir, "fig_resolution_boxplot.png"), fig_resolution_box, width = 7, height = 5, dpi = 300)
ggsave(file.path(output_dir, "fig_step_histogram_hmm_input.png"), fig_step_hist, width = 7, height = 5, dpi = 300)
ggsave(file.path(output_dir, "fig_turn_angle_histogram_hmm_input.png"), fig_angle_hist, width = 7, height = 5, dpi = 300)
ggsave(file.path(output_dir, "fig_step_by_state.png"), fig_step_by_state, width = 7, height = 5, dpi = 300)
ggsave(file.path(output_dir, "fig_turn_angle_by_state.png"), fig_angle_by_state, width = 7, height = 5, dpi = 300)
ggsave(file.path(output_dir, "fig_state_budget_by_year.png"), fig_state_budget, width = 7, height = 5, dpi = 300)
ggsave(file.path(output_dir, "fig_example_trip_by_state.png"), fig_example_trip, width = 7, height = 5, dpi = 300)
```

# Save tables

```{r save-tables}
readr::write_csv(resolution_eval, file.path(output_dir, "resolution_eval_by_track.csv"))
readr::write_csv(resolution_summary, file.path(output_dir, "resolution_summary.csv"))
readr::write_csv(post_thinning_check, file.path(output_dir, "post_thinning_check.csv"))
readr::write_csv(model_info_tbl, file.path(output_dir, "hmm_model_info.csv"))
readr::write_csv(hmm_data_decoded, file.path(output_dir, "hmm_data_decoded.csv"))
readr::write_csv(state_summary_tbl, file.path(output_dir, "hmm_state_summary.csv"))
readr::write_csv(state_budget_by_year, file.path(output_dir, "hmm_state_budget_by_year.csv"))
```

# Save model object

```{r save-model}
saveRDS(hmm_fit, file.path(output_dir, "hmm_fit_2state.rds"))
```

# Save workbook

```{r workbook}
wb <- openxlsx::createWorkbook()

openxlsx::addWorksheet(wb, "resolution_summary")
openxlsx::writeData(wb, "resolution_summary", resolution_summary)

openxlsx::addWorksheet(wb, "post_thinning_check")
openxlsx::writeData(wb, "post_thinning_check", post_thinning_check)

openxlsx::addWorksheet(wb, "model_info")
openxlsx::writeData(wb, "model_info", model_info_tbl)

openxlsx::addWorksheet(wb, "state_summary")
openxlsx::writeData(wb, "state_summary", state_summary_tbl)

openxlsx::addWorksheet(wb, "state_budget_by_year")
openxlsx::writeData(wb, "state_budget_by_year", state_budget_by_year)

openxlsx::saveWorkbook(
  wb,
  file.path(output_dir, "03_hmm_behavioural_states_outputs.xlsx"),
  overwrite = TRUE
)
```

# Interpretation summary

```{r interpretation-summary, results='asis'}
search_state <- state_summary_tbl %>%
  filter(behavioural_label == "Searching-like")

transit_state <- state_summary_tbl %>%
  filter(behavioural_label == "Transit-like")

cat("## Interpretation of HMM outputs\n\n")

cat(
  "This HMM was fitted after splitting tracks at gaps greater than **",
  max_gap_min,
  " minutes** and regularising the data to **",
  hmm_res_min,
  "-minute intervals**.\n\n",
  sep = ""
)

cat("A 2-state Hidden Markov Model was fitted to step length and turning angle derived from the regularised tracking data.\n\n")

cat(
  "The state with the shorter mean step length was interpreted as a **searching-like state**, ",
  "while the state with the longer mean step length was interpreted as a **transit-like state**.\n\n",
  sep = ""
)

cat(
  "The searching-like state had a mean step length of **",
  round(search_state$mean_step_km, 2),
  " km** and a mean absolute turning angle of **",
  round(search_state$mean_abs_turn_deg, 2),
  " degrees**. ",
  "The transit-like state had a mean step length of **",
  round(transit_state$mean_step_km, 2),
  " km** and a mean absolute turning angle of **",
  round(transit_state$mean_abs_turn_deg, 2),
  " degrees**.\n\n",
  sep = ""
)

cat(
  "These values are consistent with the ecological interpretation that searching behaviour is associated with shorter, more tortuous movement, ",
  "whereas transit behaviour is associated with longer, more directed movement.\n\n",
  sep = ""
)

cat(
  "The decoded state budget by year can be used to compare how much time birds spent in searching-like versus transit-like behaviour in 2025 and 2026. ",
  "These results provide a model-based complement to the descriptive threshold analysis carried out in script 02.\n"
)
```

# End

```{r end}
cat("03_hmm_behavioural_states completed successfully.\n")
cat("Outputs saved in:", output_dir, "\n")
```
