---
title: "Untitled"
author: "Diane Pavat"
date: "2026-03-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

# Purpose

This script estimates **space use** from retained penguin foraging-trip locations using **continuous-time autocorrelated kernel density estimation (AKDE)** with the `ctmm` package.

This is intentionally a step up from a simple classical KDE workflow because your data have two important features:

1. sampling may be **irregular in time**
2. locations are **autocorrelated**, as all movement data are

AKDE is designed to account for those issues directly, so it is generally more defensible than a basic KDE when working with animal tracking data.

This script does the following:

1. loads the clean tracking and trip objects
2. keeps only retained trip points
3. optionally excludes very extended trips for sensitivity analysis
4. converts the data to `ctmm` telemetry objects
5. fits individual continuous-time movement models
6. estimates individual AKDE home ranges
7. summarises **50%** and **95%** space-use areas
8. optionally builds **year-level mean UDs**
9. optionally constrains UDs to water using a hard boundary built from a land shapefile
10. saves outputs, figures, and interpretation text

# Why this workflow

This workflow is based on current best practice for irregular telemetry data.

The `ctmm::akde()` documentation states that AKDE calculates autocorrelated kernel density estimates from telemetry data and fitted continuous-time movement models, supports **hard boundaries** through the `SP` argument, and can **optimally weight** observations to account for sampling bias. It also explicitly cites work correcting for **missing and irregular data** in home-range estimation.

The ctmm AKDE vignette also notes that **irregular sampling** is a good use case for **weighted AKDE** (`weights=TRUE`), because uniformly weighted estimates can over-emphasise oversampled periods.

For marine animals crossing coastal landscapes, `{pathroutr}` is explicitly designed to reroute paths that cross land barriers, but its documentation also warns that rerouted paths are **not statistical models** and can introduce biases or imply unrealistic speeds. For a space-use analysis, that makes **boundary-constrained AKDE** preferable to manually rerouting the tracks themselves.

# Packages

```{r packages}
library(dplyr)
library(readr)
library(lubridate)
library(ggplot2)
library(openxlsx)
library(sf)
library(sp)
library(ctmm)
library(tidyr)
library(purrr)
```

# Project folders

```{r folders}
project_dir <- getwd()
input_dir <- file.path(project_dir, "outputs", "03_setup_and_load")
output_dir <- file.path(project_dir, "outputs", "07_akde_space_use_ctmm")

dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
```

# User controls

```{r user-controls}
exclude_extended_trips <- FALSE
extended_trip_threshold_km <- 250
land_shp <- NULL
boundary_buffer_m <- 20000
estimate_year_mean_ud <- TRUE
use_weighted_akde <- TRUE
use_debiased_akde <- TRUE
ud_levels <- c(0.50, 0.95)
```

# Helper functions

```{r helpers}
theme_paper <- function() {
  theme_minimal() +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(colour = "grey85", linewidth = 0.4),
      axis.text = element_text(colour = "grey20"),
      plot.title = element_text(face = "bold", hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5),
      legend.position = "right"
    )
}

extract_ud_area <- function(ud_obj, level_ud) {
  s <- summary(ud_obj, level.UD = level_ud, units = TRUE)
  ci_tbl <- as.data.frame(s$CI)
  ci_tbl$parameter <- rownames(ci_tbl)
  area_row <- ci_tbl %>% dplyr::filter(parameter == "area")
  tibble::tibble(
    ud_level = paste0(level_ud * 100, "%"),
    area_low = area_row[1, 1],
    area_est = area_row[1, 2],
    area_high = area_row[1, 3]
  )
}
```

# Load the analysis-ready inputs

```{r load-inputs}
tracks <- readRDS(file.path(input_dir, "tracks_clean.rds"))
trips <- readRDS(file.path(input_dir, "trips_clean.rds"))
qa <- readRDS(file.path(input_dir, "qa_clean.rds"))
```

# Keep only points belonging to retained final trips

```{r subset-final-trip-points}
tracks_trip_points <- tracks %>%
  filter(!is.na(global_trip_id)) %>%
  arrange(bird_id, global_trip_id, datetime)

cat("Points in retained final trips:", nrow(tracks_trip_points), "\n")
cat("Birds represented:", dplyr::n_distinct(tracks_trip_points$bird_id), "\n")
cat("Trips represented:", dplyr::n_distinct(tracks_trip_points$global_trip_id), "\n")
```

# Optionally exclude extended trips

```{r optional-extended-trip-filter}
trips <- trips %>%
  mutate(
    is_extended_trip = max_distance_colony_km > extended_trip_threshold_km
  )

extended_trip_ids <- trips %>%
  filter(is_extended_trip) %>%
  pull(global_trip_id)

tracks_use <- tracks_trip_points

if (exclude_extended_trips) {
  tracks_use <- tracks_use %>%
    filter(!global_trip_id %in% extended_trip_ids)
}

cat("Points used in AKDE workflow:", nrow(tracks_use), "\n")
cat("Birds used in AKDE workflow:", dplyr::n_distinct(tracks_use$bird_id), "\n")
```

# Build a hard water boundary if a land shapefile is provided

```{r build-boundary}
water_boundary_sp <- NULL

if (!is.null(land_shp)) {
  land_sf <- sf::st_read(land_shp, quiet = TRUE) %>%
    sf::st_transform(32721)

  pts_sf <- tracks_use %>%
    sf::st_as_sf(coords = c("long", "lat"), crs = 4326) %>%
    sf::st_transform(32721)

  bbox_poly <- sf::st_as_sfc(sf::st_bbox(pts_sf)) %>%
    sf::st_buffer(boundary_buffer_m)

  land_union <- sf::st_union(land_sf)
  water_poly <- sf::st_difference(bbox_poly, land_union)

  water_boundary_sp <- as(sf::st_as_sf(water_poly), "Spatial")
}
```

# Optional diagnostic: count segments that intersect land

```{r land-crossing-diagnostic}
land_crossing_count <- NA_integer_

if (!is.null(land_shp)) {
  land_sf <- sf::st_read(land_shp, quiet = TRUE) %>%
    sf::st_transform(32721)

  seg_tbl <- tracks_use %>%
    arrange(bird_id, datetime) %>%
    group_by(bird_id) %>%
    mutate(
      long_next = lead(long),
      lat_next = lead(lat)
    ) %>%
    ungroup() %>%
    filter(!is.na(long_next), !is.na(lat_next))

  if (nrow(seg_tbl) > 0) {
    seg_lines <- lapply(seq_len(nrow(seg_tbl)), function(i) {
      m <- matrix(
        c(seg_tbl$long[i], seg_tbl$lat[i],
          seg_tbl$long_next[i], seg_tbl$lat_next[i]),
        ncol = 2,
        byrow = TRUE
      )
      sf::st_linestring(m)
    })

    seg_sf <- sf::st_sf(
      geometry = sf::st_sfc(seg_lines, crs = 4326)
    ) %>%
      sf::st_transform(32721)

    land_crossing_count <- sum(lengths(sf::st_intersects(seg_sf, land_sf)) > 0)
  }
}

land_crossing_tbl <- tibble::tibble(
  land_boundary_used = !is.null(land_shp),
  n_segments_crossing_land = land_crossing_count
)

land_crossing_tbl
```

# Prepare Movebank-style input for `ctmm::as.telemetry()`

```{r prepare-telemetry-input}
ctmm_input <- tracks_use %>%
  transmute(
    individual.local.identifier = bird_id,
    timestamp = format(datetime, "%Y-%m-%d %H:%M:%S", tz = "UTC"),
    location.long = long,
    location.lat = lat,
    bird_id = bird_id,
    year_season = year_season
  )
```

# Convert to telemetry objects

```{r as-telemetry}
tele_list <- ctmm::as.telemetry(
  ctmm_input,
  timezone = "UTC"
)

length(tele_list)
names(tele_list)
```

# Fit individual continuous-time movement models

```{r fit-ctmm-models}
guess_list <- lapply(tele_list, function(dat) {
  ctmm::ctmm.guess(dat, interactive = FALSE)
})

fit_list <- Map(
  f = function(dat, guess) ctmm::ctmm.select(dat, guess, trace = 0),
  dat = tele_list,
  guess = guess_list
)

names(fit_list) <- names(tele_list)
```

# Estimate individual AKDEs

```{r estimate-individual-akde}
if (is.null(water_boundary_sp)) {
  ud_list <- ctmm::akde(
    tele_list,
    fit_list,
    weights = use_weighted_akde,
    debias = use_debiased_akde
  )
} else {
  ud_list <- ctmm::akde(
    tele_list,
    fit_list,
    weights = use_weighted_akde,
    debias = use_debiased_akde,
    SP = water_boundary_sp,
    SP.in = TRUE
  )
}

names(ud_list) <- names(tele_list)
```

# Summarise individual 50% and 95% areas

```{r summarise-individual-areas}
individual_area_tbl <- purrr::map_dfr(names(ud_list), function(id) {

  this_year <- tracks_use %>%
    filter(bird_id == id) %>%
    summarise(year_season = first(year_season)) %>%
    pull(year_season)

  out <- purrr::map_dfr(ud_levels, function(lvl) {
    extract_ud_area(ud_list[[id]], level_ud = lvl)
  })

  out %>%
    mutate(
      bird_id = id,
      year_season = this_year
    )
}) %>%
  select(bird_id, year_season, everything())

individual_area_tbl
```

# Summarise individual areas by year

```{r summarise-by-year}
individual_area_summary_by_year <- individual_area_tbl %>%
  group_by(year_season, ud_level) %>%
  summarise(
    n_birds = n(),
    mean_area = mean(area_est, na.rm = TRUE),
    sd_area = sd(area_est, na.rm = TRUE),
    min_area = min(area_est, na.rm = TRUE),
    max_area = max(area_est, na.rm = TRUE),
    .groups = "drop"
  )

individual_area_summary_by_year
```

# Estimate year-level mean UDs

```{r year-mean-uds}
year_ud_list <- NULL

if (estimate_year_mean_ud) {
  bird_year_tbl <- tracks_use %>%
    distinct(bird_id, year_season)

  year_ud_list <- split(
    ud_list,
    bird_year_tbl$year_season[match(names(ud_list), bird_year_tbl$bird_id)]
  )

  year_ud_list <- lapply(year_ud_list, function(x) {
    mean(x)
  })
}
```

# Extract year-level 50% and 95% polygons

```{r year-level-polygons}
year_ud_50_sf <- NULL
year_ud_95_sf <- NULL

if (!is.null(year_ud_list)) {
  year_ud_50_sf <- purrr::imap_dfr(year_ud_list, function(ud_obj, yr) {
    out <- ctmm::as.sf(ud_obj, level.UD = 0.50)
    out$year_season <- yr
    out
  })

  year_ud_95_sf <- purrr::imap_dfr(year_ud_list, function(ud_obj, yr) {
    out <- ctmm::as.sf(ud_obj, level.UD = 0.95)
    out$year_season <- yr
    out
  })
}
```

# Year-level mean UD areas

```{r year-level-areas}
year_mean_area_tbl <- NULL

if (!is.null(year_ud_list)) {
  year_mean_area_tbl <- purrr::imap_dfr(year_ud_list, function(ud_obj, yr) {
    purrr::map_dfr(ud_levels, function(lvl) {
      extract_ud_area(ud_obj, level_ud = lvl) %>%
        mutate(year_season = yr)
    })
  }) %>%
    select(year_season, everything())
}

year_mean_area_tbl
```

# Figures: individual AKDE area by year

```{r figures-area}
fig_individual_area <- ggplot(individual_area_tbl, aes(x = year_season, y = area_est, fill = ud_level)) +
  geom_boxplot() +
  labs(
    title = "Individual AKDE area by year",
    subtitle = "Estimated from individual bird AKDEs",
    x = "Year",
    y = "Area estimate",
    fill = "UD level"
  ) +
  theme_paper()

fig_individual_area
```

# Figures: year-level mean UD polygons

```{r figures-year-ud}
fig_year_ud <- NULL

if (!is.null(year_ud_50_sf) && !is.null(year_ud_95_sf)) {
  fig_year_ud <- ggplot() +
    geom_sf(data = year_ud_95_sf, fill = "grey70", alpha = 0.4, colour = "grey30") +
    geom_sf(data = year_ud_50_sf, fill = "red", alpha = 0.5, colour = "red4") +
    facet_wrap(~ year_season) +
    labs(
      title = "Year-level mean AKDE utilisation distributions",
      subtitle = "95% UD in grey, 50% UD in red",
      x = "Easting / Longitude",
      y = "Northing / Latitude"
    ) +
    theme_paper()

  fig_year_ud
}
```

# Save figures

```{r save-figures}
ggsave(
  file.path(output_dir, "fig_individual_akde_area_by_year.png"),
  fig_individual_area,
  width = 8,
  height = 5,
  dpi = 300
)

if (!is.null(fig_year_ud)) {
  ggsave(
    file.path(output_dir, "fig_year_mean_akde_ud.png"),
    fig_year_ud,
    width = 9,
    height = 5,
    dpi = 300
  )
}
```

# Save tables

```{r save-tables}
readr::write_csv(land_crossing_tbl, file.path(output_dir, "land_crossing_diagnostic.csv"))
readr::write_csv(individual_area_tbl, file.path(output_dir, "individual_akde_area_table.csv"))
readr::write_csv(individual_area_summary_by_year, file.path(output_dir, "individual_akde_area_summary_by_year.csv"))

if (!is.null(year_mean_area_tbl)) {
  readr::write_csv(year_mean_area_tbl, file.path(output_dir, "year_mean_akde_area_table.csv"))
}
```

# Save spatial outputs

```{r save-spatial-outputs}
if (!is.null(year_ud_50_sf)) {
  sf::write_sf(
    year_ud_50_sf,
    file.path(output_dir, "year_mean_ud_50.gpkg"),
    delete_dsn = TRUE
  )
}

if (!is.null(year_ud_95_sf)) {
  sf::write_sf(
    year_ud_95_sf,
    file.path(output_dir, "year_mean_ud_95.gpkg"),
    delete_dsn = TRUE
  )
}
```

# Save model objects

```{r save-model-objects}
saveRDS(fit_list, file.path(output_dir, "ctmm_fit_list.rds"))
saveRDS(ud_list, file.path(output_dir, "ctmm_ud_list.rds"))

if (!is.null(year_ud_list)) {
  saveRDS(year_ud_list, file.path(output_dir, "ctmm_year_mean_ud_list.rds"))
}
```

# Save workbook

```{r workbook}
wb <- openxlsx::createWorkbook()

openxlsx::addWorksheet(wb, "land_crossing_diagnostic")
openxlsx::writeData(wb, "land_crossing_diagnostic", land_crossing_tbl)

openxlsx::addWorksheet(wb, "individual_area_table")
openxlsx::writeData(wb, "individual_area_table", individual_area_tbl)

openxlsx::addWorksheet(wb, "individual_area_summary")
openxlsx::writeData(wb, "individual_area_summary", individual_area_summary_by_year)

if (!is.null(year_mean_area_tbl)) {
  openxlsx::addWorksheet(wb, "year_mean_area_table")
  openxlsx::writeData(wb, "year_mean_area_table", year_mean_area_tbl)
}

openxlsx::saveWorkbook(
  wb,
  file.path(output_dir, "04_akde_space_use_ctmm_outputs.xlsx"),
  overwrite = TRUE
)
```

# Interpretation summary

```{r interpretation-summary, results='asis'}
cat("## Interpretation of AKDE outputs\n\n")

cat(
  "This analysis estimated space use using **autocorrelated kernel density estimation (AKDE)**, ",
  "which is designed for tracking data that are irregular in time and autocorrelated. ",
  "Unlike a simple KDE, AKDE uses a fitted continuous-time movement model and therefore better respects the structure of animal movement data.\n\n",
  sep = ""
)

cat(
  "At the individual level, the **95% UD** can be interpreted as the overall space used by each bird, ",
  "whereas the **50% UD** represents the more intensively used core area. ",
  "The individual-area tables therefore allow you to compare overall and core-use area between years while preserving the individual-bird structure.\n\n",
  sep = ""
)

cat(
  "If year-level mean UDs were estimated, these represent an average utilisation surface across birds within a year. ",
  "This is often more appropriate than pooling all raw points directly, because it reduces bias from birds with disproportionate sample sizes.\n\n",
  sep = ""
)

cat(
  "If a land shapefile was provided, the AKDEs were constrained to the inferred water boundary using a hard spatial boundary. ",
  "This is preferable to manually rerouting tracks, because rerouted paths are not themselves statistical models and can introduce unrealistic behaviour.\n\n",
  sep = ""
)

cat(
  "These outputs should still be interpreted with care. ",
  "AKDE handles autocorrelation and irregular sampling more appropriately than a basic KDE, but results can still depend on track length, sample size, and whether very extended trips are included. ",
  "For that reason, it is good practice to compare full and sensitivity versions of the analysis.\n"
)
```

# End

```{r end}
cat("04_akde_space_use_ctmm completed successfully.\n")
cat("Outputs saved in:", output_dir, "\n")
```

