---
title: "04_akde_space_use_ctmm"
output:
  html_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

# Purpose

This script estimates **space use** from retained penguin foraging-trip locations using **continuous-time autocorrelated kernel density estimation (AKDE)** with the `ctmm` package.

This version runs the workflow twice:

1. **Full dataset**: includes all retained trips  
2. **Trimmed dataset**: excludes trips whose `max_distance_colony_km` exceeds a chosen threshold  

This is the safest structure for your study because it lets you compare whether one rare extended trip materially changes the space-use results.

# Packages

```{r packages}
library(dplyr)
library(readr)
library(lubridate)
library(ggplot2)
library(openxlsx)
library(sf)
library(sp)
library(ctmm)
library(tidyr)
library(purrr)
```

# Project folders

```{r folders}
project_dir <- getwd()
input_dir <- file.path(project_dir, "outputs", "03_setup_and_load")
output_dir <- file.path(project_dir, "outputs", "07_akde_space_use_ctmm")

dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
```

# User controls

```{r user-controls}
extended_trip_threshold_km <- 250
land_shp <- NULL
boundary_buffer_m <- 20000
estimate_year_mean_ud <- TRUE
use_weighted_akde <- TRUE
use_debiased_akde <- TRUE
ud_levels <- c(0.50, 0.95)
```

# Helper functions

```{r helpers}
theme_paper <- function() {
  theme_minimal() +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(colour = "grey85", linewidth = 0.4),
      axis.text = element_text(colour = "grey20"),
      plot.title = element_text(face = "bold", hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5),
      legend.position = "right"
    )
}

extract_ud_area <- function(ud_obj, level_ud) {
  s <- summary(ud_obj, level.UD = level_ud, units = TRUE)
  ci_tbl <- as.data.frame(s$CI)
  ci_tbl$parameter <- rownames(ci_tbl)
  area_row <- ci_tbl %>% dplyr::filter(parameter == "area")
  tibble::tibble(
    ud_level = paste0(level_ud * 100, "%"),
    area_low = area_row[1, 1],
    area_est = area_row[1, 2],
    area_high = area_row[1, 3]
  )
}
```

# Load the analysis-ready inputs

```{r load-inputs}
tracks <- readRDS(file.path(input_dir, "tracks_clean.rds"))
trips <- readRDS(file.path(input_dir, "trips_clean.rds"))
qa <- readRDS(file.path(input_dir, "qa_clean.rds"))
```

# Keep only points belonging to retained final trips

```{r subset-final-trip-points}
tracks_trip_points <- tracks %>%
  filter(!is.na(global_trip_id)) %>%
  arrange(bird_id, global_trip_id, datetime)

cat("Points in retained final trips:", nrow(tracks_trip_points), "\n")
cat("Birds represented:", dplyr::n_distinct(tracks_trip_points$bird_id), "\n")
cat("Trips represented:", dplyr::n_distinct(tracks_trip_points$global_trip_id), "\n")
```

# Flag extended trips and build both datasets

```{r build-full-and-trimmed-datasets}
trips <- trips %>%
  mutate(
    is_extended_trip = max_distance_colony_km > extended_trip_threshold_km
  )

extended_trip_ids <- trips %>%
  filter(is_extended_trip) %>%
  pull(global_trip_id)

tracks_use_full <- tracks_trip_points

tracks_use_trimmed <- tracks_trip_points %>%
  filter(!global_trip_id %in% extended_trip_ids)

readr::write_csv(tracks_use_full, file.path(output_dir, "tracks_use_full.csv"))
readr::write_csv(tracks_use_trimmed, file.path(output_dir, "tracks_use_trimmed.csv"))
readr::write_csv(trips, file.path(output_dir, "trips_with_extended_flag.csv"))

cat("FULL dataset - points:", nrow(tracks_use_full), "\n")
cat("FULL dataset - birds:", dplyr::n_distinct(tracks_use_full$bird_id), "\n")
cat("FULL dataset - trips:", dplyr::n_distinct(tracks_use_full$global_trip_id), "\n\n")

cat("TRIMMED dataset - points:", nrow(tracks_use_trimmed), "\n")
cat("TRIMMED dataset - birds:", dplyr::n_distinct(tracks_use_trimmed$bird_id), "\n")
cat("TRIMMED dataset - trips:", dplyr::n_distinct(tracks_use_trimmed$global_trip_id), "\n")
```

# Build a hard water boundary if a land shapefile is provided

```{r build-boundary}
# Path to Falklands land polygon shapefile
land_shp <- "C:/Users/r01dp24/OneDrive - University of Aberdeen/Documents/PhD/05-Data/01 - Shapefile/boundaries/falklands_boundaries.shp"

water_boundary_sp <- NULL
water_boundary_sf <- NULL

if (!is.null(land_shp)) {
  
  # Read land polygons and project them to UTM 21S
  land_sf <- sf::st_read(land_shp, quiet = TRUE) %>%
    sf::st_transform(32721)
  
  # Convert penguin points to sf and project them to the same CRS
  pts_sf <- tracks_use_full %>%
    sf::st_as_sf(coords = c("long", "lat"), crs = 4326) %>%
    sf::st_transform(32721)
  
  # Build a buffered bounding polygon around the tracking extent
  bbox_poly <- sf::st_as_sfc(sf::st_bbox(pts_sf)) %>%
    sf::st_buffer(boundary_buffer_m)
  
  # Merge all land polygons into one geometry
  land_union <- sf::st_union(land_sf)
  
  # Define water as study extent minus land
  water_boundary_sf <- sf::st_difference(bbox_poly, land_union)
  
  # Convert to Spatial for ctmm
  water_boundary_sp <- as(sf::st_as_sf(water_boundary_sf), "Spatial")
}

ggplot() +
  geom_sf(data = water_boundary_sf, fill = "lightblue", colour = "blue", alpha = 0.4) +
  geom_sf(data = land_sf, fill = "grey60", colour = "black") +
  geom_sf(data = pts_sf, size = 0.3, alpha = 0.3) +
  labs(title = "Water boundary used for AKDE") +
  theme_minimal()
```

# Store the two analysis datasets in a list

```{r analysis-dataset-list}
analysis_datasets <- list(
  full = tracks_use_full,
  trimmed = tracks_use_trimmed
)

names(analysis_datasets)
```

# Run AKDE workflow for both dataset versions

```{r run-akde-workflow}
analysis_results <- purrr::imap(
  analysis_datasets,
  function(tracks_use, dataset_name) {

    dataset_dir <- file.path(output_dir, dataset_name)
    dir.create(dataset_dir, recursive = TRUE, showWarnings = FALSE)

    ctmm_input <- tracks_use %>%
      transmute(
        individual.local.identifier = bird_id,
        timestamp = format(datetime, "%Y-%m-%d %H:%M:%S", tz = "UTC"),
        location.long = long,
        location.lat = lat,
        bird_id = bird_id,
        year_season = year_season
      )

    tele_list <- ctmm::as.telemetry(
      ctmm_input,
      timezone = "UTC"
    )

    guess_list <- lapply(tele_list, function(dat) {
      ctmm::ctmm.guess(dat, interactive = FALSE)
    })

    fit_list <- Map(
      f = function(dat, guess) ctmm::ctmm.select(dat, guess, trace = 0),
      dat = tele_list,
      guess = guess_list
    )
    names(fit_list) <- names(tele_list)

    if (is.null(water_boundary_sp)) {
      ud_list <- ctmm::akde(
        tele_list,
        fit_list,
        weights = use_weighted_akde,
        debias = use_debiased_akde
      )
    } else {
      ud_list <- ctmm::akde(
        tele_list,
        fit_list,
        weights = use_weighted_akde,
        debias = use_debiased_akde,
        SP = water_boundary_sp,
        SP.in = TRUE
      )
    }
    names(ud_list) <- names(tele_list)

    individual_area_tbl <- purrr::map_dfr(names(ud_list), function(id) {
      this_year <- tracks_use %>%
        filter(bird_id == id) %>%
        summarise(year_season = first(year_season)) %>%
        pull(year_season)

      out <- purrr::map_dfr(ud_levels, function(lvl) {
        extract_ud_area(ud_list[[id]], level_ud = lvl)
      })

      out %>%
        mutate(
          dataset_version = dataset_name,
          bird_id = id,
          year_season = this_year
        )
    }) %>%
      select(dataset_version, bird_id, year_season, everything())

    individual_area_summary_by_year <- individual_area_tbl %>%
      group_by(dataset_version, year_season, ud_level) %>%
      summarise(
        n_birds = n(),
        mean_area = mean(area_est, na.rm = TRUE),
        sd_area = sd(area_est, na.rm = TRUE),
        min_area = min(area_est, na.rm = TRUE),
        max_area = max(area_est, na.rm = TRUE),
        .groups = "drop"
      )

    year_ud_list <- NULL
    year_ud_50_sf <- NULL
    year_ud_95_sf <- NULL
    year_mean_area_tbl <- NULL

    if (estimate_year_mean_ud) {
      bird_year_tbl <- tracks_use %>%
        distinct(bird_id, year_season)

      year_ud_list <- split(
        ud_list,
        bird_year_tbl$year_season[match(names(ud_list), bird_year_tbl$bird_id)]
      )

      year_ud_list <- lapply(year_ud_list, function(x) {
        mean(x)
      })

      year_ud_50_sf <- purrr::imap_dfr(year_ud_list, function(ud_obj, yr) {
        out <- ctmm::as.sf(ud_obj, level.UD = 0.50)
        out$year_season <- yr
        out$dataset_version <- dataset_name
        out
      })

      year_ud_95_sf <- purrr::imap_dfr(year_ud_list, function(ud_obj, yr) {
        out <- ctmm::as.sf(ud_obj, level.UD = 0.95)
        out$year_season <- yr
        out$dataset_version <- dataset_name
        out
      })

      year_mean_area_tbl <- purrr::imap_dfr(year_ud_list, function(ud_obj, yr) {
        purrr::map_dfr(ud_levels, function(lvl) {
          extract_ud_area(ud_obj, level_ud = lvl) %>%
            mutate(
              dataset_version = dataset_name,
              year_season = yr
            )
        })
      }) %>%
        select(dataset_version, year_season, everything())
    }

    readr::write_csv(individual_area_tbl, file.path(dataset_dir, "individual_akde_area_table.csv"))
    readr::write_csv(individual_area_summary_by_year, file.path(dataset_dir, "individual_akde_area_summary_by_year.csv"))

    if (!is.null(year_mean_area_tbl)) {
      readr::write_csv(year_mean_area_tbl, file.path(dataset_dir, "year_mean_akde_area_table.csv"))
    }

    if (!is.null(year_ud_50_sf)) {
      sf::write_sf(
        year_ud_50_sf,
        file.path(dataset_dir, "year_mean_ud_50.gpkg"),
        delete_dsn = TRUE
      )
    }

    if (!is.null(year_ud_95_sf)) {
      sf::write_sf(
        year_ud_95_sf,
        file.path(dataset_dir, "year_mean_ud_95.gpkg"),
        delete_dsn = TRUE
      )
    }

    saveRDS(fit_list, file.path(dataset_dir, "ctmm_fit_list.rds"))
    saveRDS(ud_list, file.path(dataset_dir, "ctmm_ud_list.rds"))

    if (!is.null(year_ud_list)) {
      saveRDS(year_ud_list, file.path(dataset_dir, "ctmm_year_mean_ud_list.rds"))
    }

    wb <- openxlsx::createWorkbook()
    openxlsx::addWorksheet(wb, "individual_area_table")
    openxlsx::writeData(wb, "individual_area_table", individual_area_tbl)
    openxlsx::addWorksheet(wb, "individual_area_summary")
    openxlsx::writeData(wb, "individual_area_summary", individual_area_summary_by_year)

    if (!is.null(year_mean_area_tbl)) {
      openxlsx::addWorksheet(wb, "year_mean_area_table")
      openxlsx::writeData(wb, "year_mean_area_table", year_mean_area_tbl)
    }

    openxlsx::saveWorkbook(
      wb,
      file.path(dataset_dir, paste0("04_akde_outputs_", dataset_name, ".xlsx")),
      overwrite = TRUE
    )

    list(
      individual_area_tbl = individual_area_tbl,
      individual_area_summary_by_year = individual_area_summary_by_year,
      year_ud_50_sf = year_ud_50_sf,
      year_ud_95_sf = year_ud_95_sf,
      year_mean_area_tbl = year_mean_area_tbl
    )
  }
)
```

# Combine area tables across the two dataset versions

```{r combine-results}
combined_individual_area_tbl <- bind_rows(
  analysis_results$full$individual_area_tbl,
  analysis_results$trimmed$individual_area_tbl
)

combined_individual_area_summary_by_year <- bind_rows(
  analysis_results$full$individual_area_summary_by_year,
  analysis_results$trimmed$individual_area_summary_by_year
)

combined_year_mean_area_tbl <- bind_rows(
  analysis_results$full$year_mean_area_tbl,
  analysis_results$trimmed$year_mean_area_tbl
)

combined_individual_area_tbl
combined_individual_area_summary_by_year
combined_year_mean_area_tbl
```

# Figures: compare individual AKDE area by year and dataset version

```{r figures-area}
fig_individual_area <- ggplot(
  combined_individual_area_tbl,
  aes(x = year_season, y = area_est, fill = ud_level)
) +
  geom_boxplot() +
  facet_wrap(~ dataset_version) +
  labs(
    title = "Individual AKDE area by year",
    subtitle = "Comparison of full and trimmed datasets",
    x = "Year",
    y = "Area estimate",
    fill = "UD level"
  ) +
  theme_paper()

fig_individual_area
```

# Figures: compare year-level mean UD polygons

```{r figures-year-ud}
fig_year_ud <- NULL

year_ud_50_all <- bind_rows(
  analysis_results$full$year_ud_50_sf,
  analysis_results$trimmed$year_ud_50_sf
)

year_ud_95_all <- bind_rows(
  analysis_results$full$year_ud_95_sf,
  analysis_results$trimmed$year_ud_95_sf
)

if (!is.null(year_ud_50_all) && !is.null(year_ud_95_all) &&
    nrow(year_ud_50_all) > 0 && nrow(year_ud_95_all) > 0) {

  fig_year_ud <- ggplot() +
    geom_sf(data = year_ud_95_all, fill = "grey70", alpha = 0.4, colour = "grey30") +
    geom_sf(data = year_ud_50_all, fill = "red", alpha = 0.5, colour = "red4") +
    facet_grid(dataset_version ~ year_season) +
    labs(
      title = "Year-level mean AKDE utilisation distributions",
      subtitle = "95% UD in grey, 50% UD in red",
      x = "Easting / Longitude",
      y = "Northing / Latitude"
    ) +
    theme_paper()

  fig_year_ud
}
```

# Save combined figures

```{r save-figures}
ggsave(
  file.path(output_dir, "fig_individual_akde_area_by_year_full_vs_trimmed.png"),
  fig_individual_area,
  width = 9,
  height = 5,
  dpi = 300
)

if (!is.null(fig_year_ud)) {
  ggsave(
    file.path(output_dir, "fig_year_mean_akde_ud_full_vs_trimmed.png"),
    fig_year_ud,
    width = 10,
    height = 7,
    dpi = 300
  )
}
```

# Save combined tables

```{r save-combined-tables}
readr::write_csv(land_crossing_tbl, file.path(output_dir, "land_crossing_diagnostic.csv"))
readr::write_csv(combined_individual_area_tbl, file.path(output_dir, "combined_individual_akde_area_table.csv"))
readr::write_csv(combined_individual_area_summary_by_year, file.path(output_dir, "combined_individual_akde_area_summary_by_year.csv"))

if (!is.null(combined_year_mean_area_tbl) && nrow(combined_year_mean_area_tbl) > 0) {
  readr::write_csv(combined_year_mean_area_tbl, file.path(output_dir, "combined_year_mean_akde_area_table.csv"))
}
```

# Save combined workbook

```{r workbook}
wb <- openxlsx::createWorkbook()

openxlsx::addWorksheet(wb, "land_crossing_diagnostic")
openxlsx::writeData(wb, "land_crossing_diagnostic", land_crossing_tbl)

openxlsx::addWorksheet(wb, "combined_individual_area")
openxlsx::writeData(wb, "combined_individual_area", combined_individual_area_tbl)

openxlsx::addWorksheet(wb, "combined_area_summary")
openxlsx::writeData(wb, "combined_area_summary", combined_individual_area_summary_by_year)

if (!is.null(combined_year_mean_area_tbl) && nrow(combined_year_mean_area_tbl) > 0) {
  openxlsx::addWorksheet(wb, "combined_year_mean_area")
  openxlsx::writeData(wb, "combined_year_mean_area", combined_year_mean_area_tbl)
}

openxlsx::saveWorkbook(
  wb,
  file.path(output_dir, "04_akde_space_use_ctmm_full_vs_trimmed_outputs.xlsx"),
  overwrite = TRUE
)
```

# Interpretation summary

```{r interpretation-summary, results='asis'}
cat("## Interpretation of AKDE outputs

")

cat(
  "This analysis estimated space use using **autocorrelated kernel density estimation (AKDE)**, ",
  "which is designed for tracking data that are irregular in time and autocorrelated. ",
  "The workflow was run twice: once on the **full dataset** and once on a **trimmed dataset** excluding trips with maximum range greater than **",
  extended_trip_threshold_km,
  " km**.

",
  sep = ""
)

cat(
  "At the individual level, the **95% UD** represents the broader overall space used by each bird, ",
  "whereas the **50% UD** represents the more intensively used core area. ",
  "Comparing the full and trimmed versions allows you to assess whether a rare extended trip strongly changes the area estimates.

",
  sep = ""
)

cat(
  "If year-level mean UDs were estimated, these surfaces represent an average utilisation distribution across birds within each year. ",
  "This is generally more defensible than pooling all raw points directly, because it reduces bias from individuals with disproportionate sample sizes.

",
  sep = ""
)

cat(
  "If a land shapefile was provided, the UDs were constrained to water using a hard spatial boundary. ",
  "That is preferable to manually rerouting tracks, because rerouted paths are not themselves statistical models and can imply unrealistic movement. ",
  "The hard-boundary AKDE lets the space-use model respect the penguins' inability to cross land.

",
  sep = ""
)

cat(
  "The main comparison of interest is whether the **full** and **trimmed** results tell the same ecological story. ",
  "If they do, your conclusions are robust. ",
  "If they differ strongly, then the extended trip is an important but rare behaviour that should be discussed explicitly rather than silently removed.
"
)
```

# End

```{r end}
cat("04_akde_space_use_ctmm_full_and_trimmed completed successfully.
")
cat("Outputs saved in:", output_dir, "
")
```
